### 2021-11-04 mk2/manuscript.rmd
updated manuscript

---
title: "Venn diagrams indicate erroneous analyses in transcriptomics"
author: ""
date: "`r Sys.Date()`"
output:
  html_document
outputdupa:
  word_document:
    reference_docx: templates/template.docx
  pdf_document:
    latex_engine: xelatex
abstract: |
  Venn diagrams are frequently used to illustrate the results of
  differential expression analysis, and are an illustration of further
  downstream process of functional analysis. Here we show that both the use
  of Venn diagrams to find genes which are thought to be specific for a
  certain comparison as well as gene set enrichment analysis applied to
  such subsets is a fallacy and results in particularly misleading artifacts.
bibliography: manuscript.bib
link-citations: true
XXcsl: templates/nature-medicine.csl
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=FALSE,autodep=TRUE, results="hide")
```

```{r libraries,cache=FALSE}
library(tidyverse)
library(limma)
library(GEOquery)
library(DESeq2)
## install bioshmods from github.com/bihealth/bioshmods
library(bioshmods)
library(tmod)
library(cowplot)
library(stringr)
library(ggvenn)
library(ggplot2)
library(ggpval)
library(eulerr)
library(Biobase)
library(org.Hs.eg.db)
theme_set(theme_minimal() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()))
source("functions.R")
```

```{r setup}
padj_thr <- 0.05
lfc_thr  <- 1
```

## Introduction: Venn diagrams are a showcase of erroneous interpretation of interactions

Venn diagrams (VDs) are commonly used to visualize high throughput data
such as transcriptomic profiles. Frequently, a Venn diagram serves the
purpose of comparing the results from two experiments or two comparisons.
For example, a VD may illustrate up- or down-regulated genes in two strains
of mice upon infection, showing which genes are regulated in both strains,
and which are regulated only in one of them.  The VD serves
as a basis for the statement that the regulation of certain genes is
"specific" for one strain or another. We argue that not only is this
inference incorrect, but it may also lead to misleading – although
appealing – artifacts when combined with downstream analyses.

Consider a gene for which the expression has been analysed in four groups:
two different mouse strains (WT and KO) and two different experimental conditions
(naive versus infected). We find that the gene expression significantly
differs between infected and naive KO animals, but that there is no
significant difference in the WT strain. Such a gene may be incorrectly
considered as "specific" to KO, and will be accordingly entered in a Venn
diagram.

However, this is a common fallacy [@nieuwenhuis2011erroneous], since lack
of statistical significance is not the same as a significant lack of
difference. In other words, the fact that we failed to detect a significant
difference in the WT does not mean that the difference is absent and
significantly different from the difference in the KO. This "difference of
differences" is known is statistics as interaction (here, interaction
between strain and treatment).  In fact, the obtained p-values might be
just over the assumed threshold in one, but just under it in the other
strain (e.g. 0.009 vs 0.011), with the difference in both strains being
essentially the same. The correct analysis is to test the significance of
interaction using an ANOVA model. Notably, in 2011 Nieuwenhuis et al. found
that in over 150 papers in top neuroscience journals where the authors
could make this particular statistical error, half of them did.

Using VDs to show genes "specific" for a condition amounts, therefore, to
counting the times a comparison for a gene was statistically significant in
one condition, but not significant in another.

More disturbingly, Venn diagrams are a visual illustration of a procedure
that exacerbates this problem by applying a downstream analysis to
supposedly specific sets of genes. Several cells in a Venn diagram contain
variables that were selected based on the fact that they are significant in
one, but not significant in another comparison. Variables in a cell may
then be subsequently analysed to test whether a share a particular
characteristics.  For example, gene set enrichment analysis may be used to
interpret the biological function of genes that are "specific" to one
condition, but - supposedly - not to another. It turns out that due to the
fallacious nature of this procedure, this is likely to produce results that
appear reasonably in the context of the analysed biology. This will be
illustrated in the following with an example.

In this paper, we illustrate this rather simple statistical statement with
a real world data set, demonstrating how choosing this sort of approach
results in apparently sound gene set enrichment results, which are in fact
artifacts. Next, we dissect the underlying mechanism of how these artifacts
are generated.  Finally, we explore and demonstrate alternative approaches.

# Results

## Example: Transcriptomic changes due to Sars-Cov-2 infection

Consider two group of patients, G1 and G2. Each group contains 20
individuals. In both groups, there are patients who are either healthy ("no
infection") or infected with Sars-Cov-2 ("SC2"). Our aim is to understand
which genes or pathways are specifically upregulated by SC2 infection in G1
compared to G2, and vice versa. In the following, we used the data set
GSE156063 [@mick2020upper] to perform two such analyses, which arrive to opposite
conclusions.

```{r pheno_data_download,eval=FALSE}
## this sometimes fails. The RDS of the object is provided
Sys.setenv(VROOM_CONNECTION_SIZE=8*131072)

## with getGEO, we can only get the phenoData
geo <- getGEO("GSE156063")[[1]]
saveRDS(geo, file="GSE156063.geo")
```

```{r pheno_data_cleanup}
geo <- readRDS("GSE156063.geo")
covar <- as(phenoData(geo), "data.frame")

## remove columns with only one value
boring <- map_lgl(covar, ~ length(unique(.x)) == 1)
covar <- covar[ , !boring ]

## clean up
covar <- covar %>% 
  dplyr::rename(gender = "gender:ch1") %>%
  mutate(disease = gsub(" .*", "", .data[["disease state:ch1"]])) %>%
  mutate(label = description) %>%
  mutate(group = disease) %>%
  arrange(description) %>%
  dplyr::select(all_of(c("title", "label", "gender", "disease", "group")))
```

```{r featuredata_download}
if(!file.exists("GSE156063_swab_gene_counts.csv.gz")) {
  download.file("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE156nnn/GSE156063/suppl/GSE156063_swab_gene_counts.csv.gz",
                "GSE156063_swab_gene_counts.csv.gz")
}

counts <- read_csv("GSE156063_swab_gene_counts.csv.gz")
.tmp <- counts[[1]]
counts <- as.matrix(counts[,-1])
rownames(counts) <- .tmp
#counts <- counts[ , covar$description ]
counts <- counts[ , covar$label ]
lcpm   <- edgeR::cpm(counts, log=TRUE)
#stopifnot(all(colnames(counts) == covar$description))
stopifnot(all(colnames(counts) == covar$label))

annot <- data.frame(ENSEMBL = rownames(counts))

.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("ENTREZID"), keytype="ENSEMBL")
annot$ENTREZID <- .tmp[ match(annot$ENSEMBL, names(.tmp)) ]

.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("SYMBOL"), keytype="ENSEMBL")
annot$SYMBOL <- .tmp[ match(annot$ENSEMBL, names(.tmp)) ]

.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("GENENAME"), keytype="ENSEMBL")
annot$GENENAME <- .tmp[ match(annot$ENSEMBL, names(.tmp)) ]
```

```{r Preparation_of_the_covariates,cache=TRUE}
sel <- covar$group %in% c("no", "SC2")
counts <- counts[ , sel ]
covar  <- covar[ sel, ]
covar$group <- as.character(covar$group)

set.seed(0123)

#covar$disease <- covar$group

g1 <- covar %>% mutate(n=1:n()) %>%
  group_by(gender, disease) %>% slice_sample(n=10) %>% pull(n)
g2 <- covar %>% mutate(n=1:n()) %>% filter(!n %in% g1) %>%
  group_by(gender, disease) %>% slice_sample(n=10) %>% pull(n)

#g1 <- sample(1:nrow(covar), 30)
#g2 <- sample(setdiff(1:nrow(covar), g1), 30)

covar$group <- NA
covar$group[g1] <- "G1"
covar$group[g2] <- "G2"

covar$group.disease <- paste0(covar$group, '_', covar$disease)

sel <- c(g1, g2)

ds2 <- DESeqDataSetFromMatrix(counts[,sel], colData=covar[sel, ],
  design=~ 0 + group.disease )
```

```{r DESeq2,cache=TRUE}
## DESeq2 calculations

ds2 <- DESeq(ds2)
saveRDS(ds2, file="ds2_cached.rds")
```

```{r DE_analysis}
library(tidyverse)
library(tmod)
res <- list()
res$g1 <- results(ds2, contrast=c(-1, 1, 0, 0))
res$g2 <- results(ds2, contrast=c(0, 0, -1, 1))

res <- map(res, ~ .x %>% as.data.frame() %>%
  rownames_to_column("ENSEMBL") %>% 
  left_join(annot, by="ENSEMBL") %>%
  mutate(DEG=!is.na(padj) & abs(log2FoldChange) > lfc_thr & padj < padj_thr))

res.merged <- merge(res$g1, res$g2, by=c("ENSEMBL", "SYMBOL", "ENTREZID", "GENENAME"), suffixes=c(".g1", ".g2"))
```

First, we have performed differential gene expression analysis for each of
the groups G1 and G2 separately. For each comparison, we defined
genes with significantly different gene expression 
(differentially expressed genes, DEGs) 
by using an FDR threshold of 
`r padj_thr`
and absolute log~2~ fold change (LFC) threshold of
`r lfc_thr`. There were
`r sum(res$g1$DEG)` 
DE genes in the G1 group, and
`r sum(res$g2$DEG)`
in the G2 group. In total,
`r sum(res.merged$DEG.g1 & res.merged$DEG.g2)` 
DEGs were common for G1 and G2, 
`r sum(res.merged$DEG.g1 & !res.merged$DEG.g2)` 
DEGs were "specific" for G1, and 
`r sum(!res.merged$DEG.g1 & res.merged$DEG.g2)` 
were "specific" for G2 (see Fig. IncorrectAnalysis, panel *A*).
A naive interpretation of these results would indicate that while G1 shows
an almost non-existent transcriptomic response to COVID-19, individuals in
group G2 show massive regulation of gene expression. This (incorrectly)
suggests that there is a substantial difference between these two groups of
individuals.

```{r GO_db_prepare,cache=TRUE}
library(msigdbr)
msig_go_bp <- msigdbr(subcategory="GO:BP")
mset <- makeTmodFromDataFrame(msig_go_bp, 
                              feature_col="human_gene_symbol", 
                              module_col="gs_exact_source", 
                              title_col="gs_name")
mset$MODULES$N <- map_int(mset$MODULES$ID, ~ length(mset$MODULES2GENES[[.x]]))
mset <- mset[ mset$MODULES$N <= 50 & mset$MODULES$N >= 10 ]
```

```{r GSEA}
common <- res.merged %>% filter(DEG.g1 & DEG.g2) %>% pull(SYMBOL)
gsea_res <- map(res, ~ {
  fg <- .x %>% filter(padj < padj_thr & abs(log2FoldChange) > lfc_thr)
  fg <- setdiff(fg$SYMBOL, common) # only "specific" genes
  tmodHGtest(fg=fg, bg=.x$SYMBOL, mset=mset)
})
gsea_res_full <- map(res, ~ {
  fg <- .x %>% filter(padj < padj_thr & abs(log2FoldChange) > lfc_thr)
  fg <- setdiff(fg$SYMBOL, common) # only "specific" genes
  tmodHGtest(fg=fg, bg=.x$SYMBOL, mset=mset, qval = Inf)
})
gsea_res.merged <- merge(gsea_res$g1, gsea_res$g2, by=c("ID", "Title"), suffixes=c(".g1", ".g2"), all=T)
```

To understand which
pathways are upregulated in each of the two groups, we used a standard generation I
gene set enrichment analysis (GSEA) – a hypergeometric test – on the DEGs in each
group. Gene sets for GSEA were taken from the Gene Ontology (GO) database
(very large or very small gene sets were removed). For each group, we have
selected only genes which are DEGs in that group, but not the other,
mimicking a naive approach for finding "specifically" regulated pathways.
Here, a similar picture
emerged. Just
`r nrow(gsea_res$g1)` gene sets were significantly enriched in G1, and
`r nrow(gsea_res$g2)` gene sets were significantly enriched in G2. Again,
both the Venn Diagram (Fig. IncorrectAnalysis, B) and the results of
enrichments (Fig. IncorrectAnalysis, C and D) suggest that there is a
fundamental difference between the groups.

Importantly, the different GO terms enriched in the two groups were related
to infectious diseases, and may tempt to speculate about the underlying
biological differences between these two groups (e.g.  significance of
Toll like receptor 4 pathway in G1, but not G2; and, vice versa, significance of
response to interleukin 7 in G2, but not in G1).

Fig. IncorrectAnalysis. **Results of differential gene expression analysis
and gene set enrichment analysis using an incorrect Venn Diagram driven approach.**
**A**, Venn Diagram showing numbers of differentially expressed genes (DEG)
in each of the two groups, G1 and G2; **B**, Venn Diagram showing numbers
of significantly enriched GO terms in each of the two groups; **C** results
of gene set enrichment analysis for genes "specific" to group G1; **D**, results of gene set
enrichment analysis for genes "specific" to group G2 (only top 10 terms are shown).

```{r figIncorrectAnalysis, fig.width=15,fig.height=5.5}
p1 <- ggvenn(list(G1=res$g1 %>% filter(DEG) %>% pull(ENSEMBL),
                         G2=res$g2 %>% filter(DEG) %>% pull(ENSEMBL)),
             show_percentage=FALSE) +
      ggtitle("Differentially expressed genes")
p2 <- ggvenn(list(G1=gsea_res$g1 %>% pull(ID),
                         G2=gsea_res$g2 %>% pull(ID)),
              show_percentage=FALSE) +
      ggtitle("Enriched GO terms")

col1 <- plot_grid(p1, p2, labels="AUTO", nrow=2)

p3 <- plot_gsea(gsea_res$g1 %>% arrange(P.Value) %>% dplyr::slice(1:10)) + ggtitle("G1") +
  scale_x_discrete(labels=function(x) str_wrap(x, width=25))
#p3 <- plot_grid(p3, NULL, ncol=1, rel_heights=c(2, 1))
p4 <- plot_gsea(gsea_res$g2 %>% arrange(P.Value) %>% dplyr::slice(1:10)) + ggtitle("G2") +
  scale_x_discrete(labels=function(x) str_wrap(x, width=25))

plot_grid(plotlist = list(col1, p3, p4), labels=c('', 'C', 'D'), nrow=1)
```

The groups G1 and G2 were randomly sampled from the same data set.
In fact, repeated re-sampling always results in some genes being found to
be significantly different in one group, but not the other, despite the
fact that one does not expect any major differences between sets of
individuals randomly drawn from one population. Thus, the conclusions drawn
from a Venn Diagram – driven gene set enrichment analysis are artefactual. Closer
inspection of genes which belong to DEG in one group, but not the other
reveals the underlying statistical fallacy (Fig. BigReveal, *A* – *D*), that is, that difference
between significant and non-significant is, in itself, not statistically
significant [@gelman2006difference]. This does not necessarily mean that
there are no differences at all between these two groups, but that lack of
significance in one group and significance in the other group does not
correctly identify differences between groups.

```{r interaction}
res$int <- results(ds2, contrast=c(-1, 1, 1, -1)) %>%
  as.data.frame() %>%
  rownames_to_column("ENSEMBL") %>% 
  left_join(annot, by="ENSEMBL") %>%
  mutate(DEG=!is.na(padj) & abs(log2FoldChange) > lfc_thr & padj < 0.05)

fg <- res$int %>% filter(padj < padj_thr & abs(log2FoldChange) > lfc_thr)
if(nrow(fg) > 0) {
  gsea_res$int <-  tmodHGtest(fg=fg$SYMBOL, bg=res$int$SYMBOL, mset=mset)
}

tmod_res <- list()
tmod_res$int <- tmodCERNOtest(res$int %>% arrange(pvalue) %>% pull(SYMBOL),
                              mset=mset)
```

To find genes which are differentially regulated in the two groups, the
correct statistical approach is to calculate interaction between groups
(G1, G2) and disease status (no disease vs. COVID). For G1 and G2, no 
genes show a significant interaction at FDR < 0.1. In fact, this is not
surprising. The log~2~ fold changes for comparisons withing G1 and G2 are
correlated (Fig. BigReveal, *E*), and genes which have similar log~2~ fold
changes in both comparisons may have a significant p-value in one group,
but no significant p-value in the other group (e.g. Fig. BigReveal, *A*,
*C* and *D*).

Fig. BigReveal. **Genes which are significant in one comparison, but not
the other do not show a statistically significant interaction.** **A** -
**D**, examples of genes which are DEG in one group, but are not
significantly different in the other group. "Ctrl", healthy individuals; 
"SC2", Sars-Cov-2 infected patients. Values above the plot indicate FDR (p-values
corrected for multiple testing). **E**, correlation between log~2~ fold
changes in G1 and G2. Color indicates genes which are are significant in
one, but not significant in the other comparison. The overall Pearson correlation
coefficient between log~2~ fold changes is 
`r with(res.merged, format.pval(cor(log2FoldChange.g1, log2FoldChange.g2), digits=2))`.

```{r figBigReveal, fig.width=10, fig.height=5.5}
#id    <- gsub(" ", "_", toupper("response to interferon beta"))
#gstitle    <- "GOBP_TOLL_LIKE_RECEPTOR_4_SIGNALING_PATHWAY"
gstitle    <- "GOBP_INTERLEUKIN_7_MEDIATED_SIGNALING_PATHWAY"
gsid  <- gsea_res$g2 %>% filter(grepl(gstitle, Title)) %>% pull(ID)
gsgenes <- mset$MODULES2GENES[[gsid]]
sel   <- res.merged %>% filter(SYMBOL %in% gsgenes) %>% arrange(pvalue.g2) %>% 
  filter(DEG.g2 & !DEG.g1) %>%
  dplyr::slice(1:4) %>% pull(ENSEMBL)

sample_sel <- c(g1, g2)

group_lab <- gsub("_no", "/Ctrl", gsub("_SC2", "/SC2", covar$group.disease))

pees <- map(sel, ~ {
  s1 <- res$g1 %>% dplyr::slice(match(.x, ENSEMBL)) %>% pull(padj)
  s2 <- res$g2 %>% dplyr::slice(match(.x, ENSEMBL)) %>% pull(padj)
  id <- res$g1 %>% dplyr::slice(match(.x, ENSEMBL)) %>% pull(SYMBOL)

  ggplot_gene(lcpm[.x, sample_sel ], 
              group_lab[ sample_sel ],
              annotation=format.pval(c(s1, s2), digits=2), textsize=8,
              pval_text_adj=.5) +
            ggtitle(id)
})

part1 <- plot_grid(plotlist=pees, labels="AUTO", ncol=2)

tmp <- res.merged %>% mutate(specific= DEG.g1 != DEG.g2) %>%
                      mutate(tlr4=SYMBOL %in% gsgenes)

part2 <- ggplot(tmp, aes(x=log2FoldChange.g1, 
                                y=log2FoldChange.g2, 
                                color=specific,
                                symbol=tlr4)) + 
         scale_color_manual(values=c("TRUE"="red", "FALSE"="#666666")) +
         geom_point(alpha=.5) + xlab("G1") + ylab("G2") +
         geom_hline(yintercept=0, color="grey") +
         geom_vline(xintercept=0, color="grey") +
         geom_abline(slope=1, intercept=0, color="grey") +
         guides(color=FALSE)

plot_grid(part1, part2, rel_widths=c(2, 2), labels=c("", "E"))
```
Consequently, it is not possible to calculate gene set enrichment using a
hypergeometric test, as there are no DEGs for the interaction contrast.
Gene set enrichment using a second generation algorithm, relying on gene
ordering rather than selecting a set of DEGs [@zyla2019gene], does not show
any significant enrichment.

However, in the gene set enrichment analysis of the genes "specific" for a
given comparison – i.e., genes which are significant in that comparison,
but not significant in others – we have observed a number of terms
associated with immune response. It is a crucial point of this manuscript
to note that the spurious enrichments not only show significant p-values,
but also that the terms or pathways which appear in them are relevant to
the research hypothesis being tested. Below, we will show why these terms
(rather than random terms which have no obvious relevance for an infectious
disease) appear in the results.

## Artifacts arise because of false negatives

To understand how gene set enrichment actually gives significant
results in such randomly generated gene sets, despite absence of genes with
significant interaction, it is first necessary to consider the definition
of a differentially expressed gene in this context. Most often than not,
DEGs are defined by a threshold in p-value adjusted for multiple testing,
possibly combined with a threshold in log~2~ fold change. The commonly used
Benjamini-Hochberg procedure ensures that among genes for which the adjusted
p~adj~ < 0.05 there are at most 5% false positives.

This way, we can exert control over the false positive rate (FPR, type I
errors), keeping it at a relatively low level. However, we do not control
the false negative rate (FNR, type II errors).  In a powerful statistical
test (such as a t-test), the test power may achieve 80%; that is, the FNR
is 20%. Clearly, false negatives (FNs) occur at much higher rates than
false positives (FPs). In case of high throughput data sets, where the FPR is
controlled by Bejnamini-Hochberg procedure or a similar technique, the FNR
may be even as high as 80% [@white2019beyond].

```{r full_dataset}
ds2_full <- DESeqDataSetFromMatrix(counts, colData=covar, design=~ disease )
ds2_full<- DESeq(ds2_full)
res$full <- results(ds2_full, name="disease_SC2_vs_no") %>% as.data.frame() %>%
  rownames_to_column("ENSEMBL") %>% 
  left_join(annot, by="ENSEMBL") %>%
  mutate(DEG=!is.na(padj) & abs(log2FoldChange) > lfc_thr & padj < padj_thr)
```

```{r g_specific}
g1_spec <- res.merged %>% filter(DEG.g1 & !DEG.g2) %>% pull(ENSEMBL)
g2_spec <- res.merged %>% filter(!DEG.g1 & DEG.g2) %>% pull(ENSEMBL)
```

These FNs will occur
at a much higher rate within the sets of "specific" DEGs defined by the
non-overlaping areas of the VDs.
To illustrate this phenomenon, we have analysed the full data set from which G1 and G2
were drawn, comparing the 
`r sum(covar$disease == "no") `
healthy controls to 
`r sum(covar$disease == "SC2") `
COVID-19 patients. Of the
`r length(g1_spec)`
genes significant in G1, but not in G2, 
`r .n <- res$full %>% filter(DEG & ENSEMBL %in% g1_spec) %>% nrow; .n`
(`r sprintf("%.0f%%", .n/length(g1_spec) * 100)`) 
are significant in the full data set; 
of the
`r length(g2_spec)`
genes significant in G2, but not in G1, 
`r .n <- res$full %>% filter(DEG & ENSEMBL %in% g2_spec) %>% nrow; .n`
(`r sprintf("%.0f%%", .n/length(g2_spec) * 100)`) 
are significant in the full data set.
Given that G1 and G2 were sampled from the total population, and since the
FDR was set to `r padj_thr`, we can assume that between a third and a half of
the genes that appeared to be "specific" in the initial analysis were, in
fact, false negatives in one of the comparisons.

In other words, a substantial fraction of the "specific" genes are genes
that are in reality differentially expressed in both groups alike.
Therefore, if one is to perform a gene set enrichment analysis on one of
these "specific" groups of genes, then the enriched functions will be
related to the pathways and processes up- or down-regulated in both groups
due to the common factor (in this example, the COVID-19 disease), but
which are not related to differences between the two groups.

**Fig. RelationshipDEGs**. Area-proportional Venn diagram showing overlaps in DEGs between
G1, G2 and the full data set. The majority of genes which have been labeled
as DEGs in only one of the groups G1, G2 are DEGs when all data were
analysed.

```{r}
tmp <- merge(res.merged, res$full, by="ENSEMBL")
tmp <- tmp[ , c("DEG.g1", "DEG.g2", "DEG") ]
colnames(tmp) <- c("G1", "G2", "Full dataset")
plot(euler(tmp), quantities=list(type=c("counts", "percent")))
```

# Discussion

```{r lit_surves}
links <- system2("grep", args=c("https", "literature_survey.md"), stdout=TRUE)
n_tot <- length(links)
n_vd_only <- length(grep("vd only", links))
```

In an informal survey of articles from the journal "Scientific reports"
(chosen for its broad scope, large number of published articles and open
access), of the 238 articles from 2021 which used the term "venn diagram" and
"differential expression", at least 
`r n_tot`
were using Venn diagrams to compare
statistical significance with lack thereof by referring to "unique",
"specific", "solely regulated" or "exclusive" DEGs. Out of these, at least 
`r n_vd_only`
coupled the VDs with some form of gene set enrichment analysis on the
set of supposedly "specific" DEGs.

Drawing conclusions from comparing significance with lack thereof is a
common statistical fallacy. Just as absence of evidence is not evidence of
absence, the failure to reject the null hypothesis does not consitute the
same level of evidence as rejecting it. However, when such an incorrect
analysis is combined with downstream functional analysis – i.e., gene set
enrichments of genes "specific" to one or another comparison – the
resulting pathways or gene ontologies are misleadingly relevant, yielding
gene sets associated with immune answer for research hypotheses involving
an infectious disease, or cancer pathways if the underlying research
hypothesis involved cancer treatment. Such results may be hard to resist
for an involved researcher, especially if the correct analysis of
interactions does not show any significant differences.

That is not to say that VDs are not a useful tool, also in the context of
transcriptomics and gene set enrichments. For example, analysis of an
intersection of DEGs (i.e., by considering genes from the overlap in a VD)
is not an incorrect procedure.  Genes in the overlapping part of a VD are
significant in both (or all) comparisons, hence no comparison between
significance and non-significance is made.

While the use of Venn diagrams in combination with the incorrect gene set
enrichment analysis is common, it is by far not the only way of achieving
artifactual and incorrect conclusions by means of comparing significant
with non-significant results in transcriptomics and other high-throughput
applications. Our literature survey showed that at least two other ways are
widely spread. Firstly, the direct comparison of gene set enrichment
results: that is, drawing conclusions from the fact that a gene set
enrichment result was significant in one comparison only. Second, while
Venn diagrams are often used to illustrate the numbers of "specific" DEGs
and so present a mean to find examples of this fallacy in scientific
literature, researches often test for enrichment these "specific" genes
without using the phrase "Venn diagram" or even clearly stating how the
lists of "specific" genes were derived. In all these cases, the analysis
boils down to comparing results significant in one, but not in another
comparison.

# Methods

## Methods availability

This document has been written as an Rmarkdown file, containing all
statistical calculations required to replicate the findings and figures.
The markdown file, along with additional files required to recreate this
manuscript have been uploaded to [https://github.com/bihealth/manuscript_venn_diagrams](https://github.com/bihealth/manuscript_venn_diagrams).

## Data

The expression data as a count matrix has been downloaded from GEO,
accession GSE156063.

## Statistical analyses

For differential gene expression, the R package DESeq2, version
`r packageVersion("DESeq2")`
has been used. Gene set enrichments were done using either hypergeometric
test (where stated) or CERNO using the package tmod [@zyla2019gene],
version
`r packageVersion("tmod")`.
GO terms have been sourced from the R package msigdbr, version
`r packageVersion("msigdbr")`.

## Literature survey

An informal literature survey was performed using Google Scholar to
estimate the frequency of the incorrect use of Venn Diagrams. We searched
for articles mentioning the phrases "differential expression" and "venn
diagram" in the Scientific Reports journal from 2020. The journal was
selected because it represents a wide spectrum of research areas, the
articles are available through an open access license and there is a large
number of publications per year. For each of the papers identified, we
tested whether (i) the authors used the VD to show differentially expressed
transcripts significant in one comparison, but not another, (ii) the
authors discussed "unique", "non-overlapping" or "specific" regions of the
Venn diagram and (iii) whether this was coupled to gene set enrichment
analysis is any form. Articles which (i) focused only on the intersections
of the Venn diagrams (genes common to all conditions), (ii) which used the
Venn diagrams for a purpose other than to compare genes significant in one
condition, but not significant in another condition or (iii) for which a
clear-cut error could not be spotted were not considered incorrect.

### 2021-11-19 mk2/manuscript.rmd
updated manuscript

@@
 ---
-title: "Venn diagrams indicate erroneous analyses in transcriptomics"
-author: ""
+title: "Erroneous use of Venn diagram leads to artifacts in transcriptomic analyses"
+author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
 date: "`r Sys.Date()`"
+outputoff:
+  html_document
 output:
-  html_document
-outputdupa:
   word_document:
@@
     latex_engine: xelatex
-abstract: |
-  Venn diagrams are frequently used to illustrate the results of
-  differential expression analysis, and are an illustration of further
-  downstream process of functional analysis. Here we show that both the use
-  of Venn diagrams to find genes which are thought to be specific for a
-  certain comparison as well as gene set enrichment analysis applied to
-  such subsets is a fallacy and results in particularly misleading artifacts.
 bibliography: manuscript.bib
@@
 ---
+
+^1^ Berlin Institute of Health at Charité – Universitätsmedizin Berlin, Core Unit Bioinformatics, Charitéplatz 1, 10117 Berlin, Germany
+
+^†^ To whom the correspondence should be addressed
+
+## Abstract
+
+Venn diagrams are frequently used to illustrate the results of
+differential expression analysis and further
+downstream functional analysis. Here we show that both, the use
+of Venn diagrams to find genes which are thought to be specific for a
+certain comparison, as well as gene set enrichment analysis applied to
+such subsets is a fallacy. Moreover, the combination of incorrect statistical reasoning
+and gene set enrichment analysis may result in particularly misleading artifacts.
 
@@
 Consider a gene for which the expression has been analysed in four groups:
-two different mouse strains (WT and KO) and two different experimental conditions
+two different mouse strains (wildtype, WT and knockout, KO) and two different experimental conditions
 (naive versus infected). We find that the gene expression significantly
@@
 However, this is a common fallacy [@nieuwenhuis2011erroneous], since lack
-of statistical significance is not the same as a significant lack of
+of statistical significance is not the same as testing for lack of
 difference. In other words, the fact that we failed to detect a significant
 difference in the WT does not mean that the difference is absent and
-significantly different from the difference in the KO. This "difference of
-differences" is known is statistics as interaction (here, interaction
-between strain and treatment).  In fact, the obtained p-values might be
-just over the assumed threshold in one, but just under it in the other
-strain (e.g. 0.009 vs 0.011), with the difference in both strains being
-essentially the same. The correct analysis is to test the significance of
-interaction using an ANOVA model. Notably, in 2011 Nieuwenhuis et al. found
-that in over 150 papers in top neuroscience journals where the authors
-could make this particular statistical error, half of them did.
+itself significantly different from the difference in the KO. This "difference of
+differences" is known is statistics as interaction [@blalock1965theory].
+Here, it is the interaction between strain and treatment.  In fact, the
+obtained p-values might be just over the assumed p-value threshold in one, but just
+under it in the other strain (e.g. 0.009 vs 0.011), with the difference in
+both strains being essentially the same. The correct analysis is to test
+the significance of interaction using an ANOVA model. Notably, in 2011
+Nieuwenhuis et al. found that in over 150 papers in top neuroscience
+journals where the authors could make this particular statistical error,
+half of them did.
 
-Using VDs to show genes "specific" for a condition amounts, therefore, to
+Using VDs to show genes "specific" for a condition amounts to
 counting the times a comparison for a gene was statistically significant in
 one condition, but not significant in another.
-
-More disturbingly, Venn diagrams are a visual illustration of a procedure
+More disturbingly, VDs are a visual illustration of a procedure
 that exacerbates this problem by applying a downstream analysis to
 supposedly specific sets of genes. Several cells in a Venn diagram contain
-variables that were selected based on the fact that they are significant in
-one, but not significant in another comparison. Variables in a cell may
-then be subsequently analysed to test whether a share a particular
+numbers corresponding to features (such as genes), for which there was a significant test result in
+one, but not in another comparison. The sets of genes in a Venn diagram cell may
+then be subsequently analysed to test whether they share a particular
 characteristics.  For example, gene set enrichment analysis may be used to
-interpret the biological function of genes that are "specific" to one
-condition, but - supposedly - not to another. It turns out that due to the
-fallacious nature of this procedure, this is likely to produce results that
-appear reasonably in the context of the analysed biology. This will be
-illustrated in the following with an example.
+interpret the biological function of genes that are considered "specific" to one
+condition. It turns out that due to the
+fallacious nature of this procedure, it is likely to produce results that
+appear to make sense in the given biological context.
 
@@
 a real world data set, demonstrating how choosing this sort of approach
-results in apparently sound gene set enrichment results, which are in fact
+results in apparently sound gene set enrichment results which are, in fact,
 artifacts. Next, we dissect the underlying mechanism of how these artifacts
-are generated.  Finally, we explore and demonstrate alternative approaches.
+are generated.  Finally, we discuss alternative approaches.
 
-# Results
+## Results
 
-## Example: Transcriptomic changes due to Sars-Cov-2 infection
+### Transcriptomic changes due to Sars-Cov-2 infection
 
-Consider two group of patients, G1 and G2. Each group contains 20
-individuals. In both groups, there are patients who are either healthy ("no
-infection") or infected with Sars-Cov-2 ("SC2"). Our aim is to understand
-which genes or pathways are specifically upregulated by SC2 infection in G1
-compared to G2, and vice versa. In the following, we used the data set
-GSE156063 [@mick2020upper] to perform two such analyses, which arrive to opposite
-conclusions.
+```{r}
+n_stratum <- 10 # 2 strata per condition 
+```
+
+Consider two group of patients, G1 and G2. Each group contains 
+`r n_stratum * 4`
+individuals. In both groups, there is an equal number of healthy
+individuals (labeled "Ctrl" on figures below) 
+or patients infected with Sars-Cov-2 (labeled "SC2"). Our aim is to understand
+the differences between G1 and G2 in the response to infection. For
+example, we ask which genes or pathways are specifically upregulated by SC2
+infection in G1 as compared to G2, and vice versa. In the following, we
+used the data set GSE156063 [@mick2020upper] in two approaches (an
+incorrect and the correct one) which arrive at opposite conclusions.
 
@@
 ```{r featuredata_download}
+## the counts must be downloaded from GEO separately.
 if(!file.exists("GSE156063_swab_gene_counts.csv.gz")) {
@@
 g1 <- covar %>% mutate(n=1:n()) %>%
-  group_by(gender, disease) %>% slice_sample(n=10) %>% pull(n)
+  group_by(gender, disease) %>% slice_sample(n=n_stratum) %>% pull(n)
 g2 <- covar %>% mutate(n=1:n()) %>% filter(!n %in% g1) %>%
-  group_by(gender, disease) %>% slice_sample(n=10) %>% pull(n)
+  group_by(gender, disease) %>% slice_sample(n=n_stratum) %>% pull(n)
 
@@
 First, we have performed differential gene expression analysis for each of
-the groups G1 and G2 separately. For each comparison, we defined
+the groups G1 and G2 separately using standard bioinformatic tools. For each comparison, we defined
 genes with significantly different gene expression 
@@
 `r sum(res.merged$DEG.g1 & !res.merged$DEG.g2)` 
-DEGs were "specific" for G1, and 
+DEGs were significant in G1 only ("specific" for G1), and 
 `r sum(!res.merged$DEG.g1 & res.merged$DEG.g2)` 
-were "specific" for G2 (see Fig. IncorrectAnalysis, panel *A*).
-A naive interpretation of these results would indicate that while G1 shows
-an almost non-existent transcriptomic response to COVID-19, individuals in
-group G2 show massive regulation of gene expression. This (incorrectly)
-suggests that there is a substantial difference between these two groups of
-individuals.
+were significant in G2 only ("specific" for G2; see Fig. 1, panel *A*).
+A naive interpretation of these results implies that there is a substantial
+difference between these two groups of individuals, as evidenced by a small
+overlap in commonly regulated genes, while the majority of DEGs is
+significant in one comparison only.
 
@@
 library(msigdbr)
+min_mod_size <- 10
+max_mod_size <- 50
 msig_go_bp <- msigdbr(subcategory="GO:BP")
@@
 mset$MODULES$N <- map_int(mset$MODULES$ID, ~ length(mset$MODULES2GENES[[.x]]))
-mset <- mset[ mset$MODULES$N <= 50 & mset$MODULES$N >= 10 ]
+mset <- mset[ mset$MODULES$N <= max_mod_size & 
+              mset$MODULES$N >= min_mod_size ]
 ```
@@
 gene set enrichment analysis (GSEA) – a hypergeometric test – on the DEGs in each
-group. Gene sets for GSEA were taken from the Gene Ontology (GO) database
-(very large or very small gene sets were removed). For each group, we have
+group. Gene sets for GSEA were taken from the Gene Ontology (GO) database.
+Gene sets with more than `r max_mod_size` or 
+fewer than `r min_mod_size` genes were removed.
+For each group, we have
 selected only genes which are DEGs in that group, but not the other,
@@
 `r nrow(gsea_res$g2)` gene sets were significantly enriched in G2. Again,
-both the Venn Diagram (Fig. IncorrectAnalysis, B) and the results of
-enrichments (Fig. IncorrectAnalysis, C and D) suggest that there is a
-fundamental difference between the groups.
+both the Venn diagram (Fig. 1, B) and the results of
+enrichments (Fig. 1, C and D) suggest that there is a
+fundamental difference between the groups, and that the groups have little
+in common in their response to the virus.
 
@@
 
-Fig. IncorrectAnalysis. **Results of differential gene expression analysis
-and gene set enrichment analysis using an incorrect Venn Diagram driven approach.**
-**A**, Venn Diagram showing numbers of differentially expressed genes (DEG)
-in each of the two groups, G1 and G2; **B**, Venn Diagram showing numbers
+**Fig. 1. Results of differential gene expression analysis
+and gene set enrichment analysis using an incorrect Venn diagram driven approach.**
+**A**, Venn diagram showing numbers of differentially expressed genes (DEG)
+in each of the two groups, G1 and G2; **B**, Venn diagram showing numbers
 of significantly enriched GO terms in each of the two groups; **C** results
@@
 
-```{r figIncorrectAnalysis, fig.width=15,fig.height=5.5}
+```{r fig1, fig.width=15,fig.height=5.5,dpi=300,dev="png"}
 p1 <- ggvenn(list(G1=res$g1 %>% filter(DEG) %>% pull(ENSEMBL),
@@
 individuals randomly drawn from one population. Thus, the conclusions drawn
-from a Venn Diagram – driven gene set enrichment analysis are artefactual. Closer
+from a Venn diagram-driven gene set enrichment analysis are artefactual. Closer
 inspection of genes which belong to DEG in one group, but not the other
-reveals the underlying statistical fallacy (Fig. BigReveal, *A* – *D*), that is, that difference
+reveals the underlying statistical fallacy (Fig. 2, *A* – *D*), that is, that difference
 between significant and non-significant is, in itself, not statistically
@@
 
+```{r correlation_coefs}
+## calculate correlation coefficients for various groups of genes
+## all correlations are significant
+
+cors <- list()
+cors$all <- with(res.merged, 
+                cor(log2FoldChange.g1, log2FoldChange.g2))
+cors$sign <- with(res.merged %>% filter(DEG.g1 | DEG.g2), 
+                cor(log2FoldChange.g1, log2FoldChange.g2))
+cors$g1 <- with(res.merged %>% filter(DEG.g1 & !DEG.g2), 
+                cor(log2FoldChange.g1, log2FoldChange.g2))
+cors$g2 <- with(res.merged %>% filter(!DEG.g1 & DEG.g2), 
+                cor(log2FoldChange.g1, log2FoldChange.g2))
+
+```
+
 To find genes which are differentially regulated in the two groups, the
 correct statistical approach is to calculate interaction between groups
-(G1, G2) and disease status (no disease vs. COVID). For G1 and G2, no 
-genes show a significant interaction at FDR < 0.1. In fact, this is not
+(G1, G2) and disease status (no disease vs. COVID). While it may be argued
+that a test for interaction has lower power than a test for a simple
+contrast, no 
+genes show a significant interaction even at FDR < 0.1. In fact, this is not
 surprising. The log~2~ fold changes for comparisons withing G1 and G2 are
-correlated (Fig. BigReveal, *E*), and genes which have similar log~2~ fold
-changes in both comparisons may have a significant p-value in one group,
-but no significant p-value in the other group (e.g. Fig. BigReveal, *A*,
-*C* and *D*).
+strongly correlated (Fig. 2, *E*). For all significant genes, the Pearson correlation
+coefficient is
+`r format.pval(cors$sign, digits=2)`, 
+while for genes
+exclusively significant in G1 or G2 (genes "specific" to G1 or G2), it is
+`r format.pval(cors$g1, digits=2)` 
+and
+`r format.pval(cors$g2, digits=2)`, 
+respectively. Thus, genes which are significant in one, but not in the
+other comparison often have have similar log~2~ fold changes (e.g. Fig.
+2, *A*, *C* and *D*).
 
-Fig. BigReveal. **Genes which are significant in one comparison, but not
+**Fig. 2. Genes which are significant in one comparison, but not
 the other do not show a statistically significant interaction.** **A** -
@@
 changes in G1 and G2. Color indicates genes which are are significant in
-one, but not significant in the other comparison. The overall Pearson correlation
+one, but not significant in the other comparison; red indicates genes significant in G1, 
+blue indicates genes significant in G2. The overall Pearson correlation
 coefficient between log~2~ fold changes is 
@@
 
-```{r figBigReveal, fig.width=10, fig.height=5.5}
+```{r fig2, fig.width=10, fig.height=5.5, dpi=300,dev="png"}
 #id    <- gsub(" ", "_", toupper("response to interferon beta"))
@@
 
-tmp <- res.merged %>% mutate(specific= DEG.g1 != DEG.g2) %>%
+tmp <- res.merged %>% mutate(specific= 
+                             ifelse(DEG.g1 & !DEG.g2,
+                                    "G1", 
+                                    ifelse(!DEG.g1 & DEG.g2, "G2", "NS"))) %>%
                       mutate(tlr4=SYMBOL %in% gsgenes)
@@
                                 y=log2FoldChange.g2, 
-                                color=specific,
-                                symbol=tlr4)) + 
-         scale_color_manual(values=c("TRUE"="red", "FALSE"="#666666")) +
+                                color=specific)) + 
+         scale_color_manual(values=c(G1="red", G2="blue", NS="#666666")) +
          geom_point(alpha=.5) + xlab("G1") + ylab("G2") +
@@
 ```
-Consequently, it is not possible to calculate gene set enrichment using a
+Consequently, it is not possible to calculate gene set enrichment for the
+interaction using a
 hypergeometric test, as there are no DEGs for the interaction contrast.
-Gene set enrichment using a second generation algorithm, relying on gene
-ordering rather than selecting a set of DEGs [@zyla2019gene], does not show
-any significant enrichment.
+Gene set enrichment using a second generation algorithm (CERNO), relying on the
+ordering of genes according to their raw p-values from the interaction
+contrast rather than selecting a
+set of DEGs [@zyla2019gene], does not show any significant enrichment.
 
-However, in the gene set enrichment analysis of the genes "specific" for a
-given comparison – i.e., genes which are significant in that comparison,
-but not significant in others – we have observed a number of terms
-associated with immune response. It is a crucial point of this manuscript
-to note that the spurious enrichments not only show significant p-values,
-but also that the terms or pathways which appear in them are relevant to
-the research hypothesis being tested. Below, we will show why these terms
-(rather than random terms which have no obvious relevance for an infectious
-disease) appear in the results.
+### Artifacts arise because of false negatives
 
-## Artifacts arise because of false negatives
+It is worth noting that in the gene set enrichment analysis of the genes
+"specific" for a given comparison – i.e., genes which are significant in
+that comparison, but not significant in others – we have observed a number
+of terms associated with immune response. It is a crucial point of this
+manuscript to note that the spurious enrichments not only show significant
+p-values, but also that the terms or pathways which appear in them are
+relevant to the research hypothesis being tested. Below, we will show why
+these terms (rather than random terms which have no obvious relevance for
+an infectious disease) appear in the results.
 
@@
 significant interaction, it is first necessary to consider the definition
-of a differentially expressed gene in this context. Most often than not,
+of a differentially expressed gene in this context. More often than not,
 DEGs are defined by a threshold in p-value adjusted for multiple testing,
 possibly combined with a threshold in log~2~ fold change. The commonly used
-Benjamini-Hochberg procedure ensures that among genes for which the adjusted
-p~adj~ < 0.05 there are at most 5% false positives.
+Benjamini-Hochberg procedure [@benjamini1995controlling] ensures that among
+genes for which the adjusted p~adj~ < 0.05 there are at most 5% false
+positives.
 
@@
 the false negative rate (FNR, type II errors).  In a powerful statistical
-test (such as a t-test), the test power may achieve 80%; that is, the FNR
-is 20%. Clearly, false negatives (FNs) occur at much higher rates than
-false positives (FPs). In case of high throughput data sets, where the FPR is
-controlled by Bejnamini-Hochberg procedure or a similar technique, the FNR
-may be even as high as 80% [@white2019beyond].
+test (such as a t-test), the test power in a typical application will
+rarely achieve more than 80%. For example, even for large
+effects (Cohen's d > 0.8) and type I error rate of 0.05, a t-test only achieves
+80% power with at least 
+`r floor(pwr.t.test(power=.8, d=0.8)$n)`
+samples per group. For small effects (Cohen's d > 0.2), the required number
+of samples is at least
+`r floor(pwr.t.test(power=.8, d=0.2)$n)`
+per group.
+Even assuming a test power of 80%, the FNR is 20%. Clearly, false negatives (FNs) occur
+at much higher rates than false positives (FPs). In case of high throughput
+data sets, where the FPR is controlled by Bejnamini-Hochberg procedure or a
+similar technique, the FNR may be even as high as 80% [@white2019beyond].
 
@@
 These FNs will occur
-at a much higher rate within the sets of "specific" DEGs defined by the
-non-overlaping areas of the VDs.
+at a much higher rate within the sets of DEGs defined by the
+non-overlaping areas of the VDs, that is DEGs considered to be "specific"
+for one group or other in a naive approach.
 To illustrate this phenomenon, we have analysed the full data set from which G1 and G2
-were drawn, comparing the 
+were drawn (Fig. 3), comparing the 
 `r sum(covar$disease == "no") `
@@
 Given that G1 and G2 were sampled from the total population, and since the
-FDR was set to `r padj_thr`, we can assume that between a third and a half of
+FDR was set to `r padj_thr`, we can assume that at least between a third and a half of
 the genes that appeared to be "specific" in the initial analysis were, in
@@
 
-**Fig. RelationshipDEGs**. Area-proportional Venn diagram showing overlaps in DEGs between
-G1, G2 and the full data set. The majority of genes which have been labeled
+**Fig. 3. Area-proportional Venn diagram showing overlaps in DEGs between
+G1, G2 and the full data set.** The majority of genes which have been labeled
 as DEGs in only one of the groups G1, G2 are DEGs when all data were
@@
 
-```{r}
+```{r fig3,dpi=300,dev="png"}
 tmp <- merge(res.merged, res$full, by="ENSEMBL")
@@
 
-# Discussion
+## Discussion
 
@@
 Drawing conclusions from comparing significance with lack thereof is a
-common statistical fallacy. Just as absence of evidence is not evidence of
+common statistical fallacy [@gelman2006difference]. Just as absence of evidence is not evidence of
 absence, the failure to reject the null hypothesis does not consitute the
@@
 
-That is not to say that VDs are not a useful tool, also in the context of
-transcriptomics and gene set enrichments. For example, analysis of an
+That is not to say that VDs are not a useful tool, even in the context of
+transcriptomics and gene set enrichments, if used correctly. For example, analysis of an
 intersection of DEGs (i.e., by considering genes from the overlap in a VD)
@@
 
+As an alternative to Venn diagrams two approaches can be considered.
+Firstly, as shown above, a test for interaction can reveal genes for which
+the impact of treatment significantly differ between the groups. As a mean
+of visualization, we recommend plotting the log~2~ fold changes in one
+comparison versus the log~2~ fold changes in the other comparison. In
+addition, a concordance / discordance analysis may provide additional
+insight in genes which are regulated either in the same direction or in the
+opposite directions in both groups [@domaszewska2017concordant].
+
 While the use of Venn diagrams in combination with the incorrect gene set
-enrichment analysis is common, it is by far not the only way of achieving
+enrichment analysis is common, it is not the only way of achieving
 artifactual and incorrect conclusions by means of comparing significant
@@
 applications. Our literature survey showed that at least two other ways are
-widely spread. Firstly, the direct comparison of gene set enrichment
+common. Firstly, the direct comparison of gene set enrichment
 results: that is, drawing conclusions from the fact that a gene set
-enrichment result was significant in one comparison only. Second, while
-Venn diagrams are often used to illustrate the numbers of "specific" DEGs
-and so present a mean to find examples of this fallacy in scientific
-literature, researches often test for enrichment these "specific" genes
-without using the phrase "Venn diagram" or even clearly stating how the
-lists of "specific" genes were derived. In all these cases, the analysis
-boils down to comparing results significant in one, but not in another
-comparison.
+enrichment result was significant in one comparison only.
+Second, while Venn diagrams are often used to illustrate the numbers of
+"specific" DEGs and so present a mean to find examples of this fallacy in
+scientific literature, researches often test for enrichment these
+"specific" genes without using the phrase "Venn diagram" or even clearly
+stating how the lists of "specific" genes were derived. In all these cases,
+the analysis boils down to comparing results significant in one, but not in
+another comparison.
 
-# Methods
+The use of Venn diagrams to illustrate specific differences between
+comparisons should therefore be abandoned in favor of statistically correct
+approaches. Furthermore, gene set enrichment analysis must never be applied
+to sets of genes defined as significant in one comparison, but not the
+other.
 
-## Methods availability
+## Methods
+
+### Methods availability
 
@@
 
-## Data
+### Data
 
@@
 
-## Statistical analyses
+### Statistical analyses
 
+Power calculation was done using the R package pwr, version
+`r packageVersion("pwr")`.
 For differential gene expression, the R package DESeq2, version
@@
 has been used. Gene set enrichments were done using either hypergeometric
-test (where stated) or CERNO using the package tmod [@zyla2019gene],
+test (where stated) or the CERNO test using the package tmod [@zyla2019gene],
 version
@@
 
-## Literature survey
+### Literature survey
 
 An informal literature survey was performed using Google Scholar to
-estimate the frequency of the incorrect use of Venn Diagrams. We searched
-for articles mentioning the phrases "differential expression" and "venn
-diagram" in the Scientific Reports journal from 2020. The journal was
+estimate the frequency of the incorrect use of Venn diagrams. We searched
+for articles from 2020 mentioning the phrases "differential expression" and "venn
+diagram" in the Scientific Reports journal. The journal was
 selected because it represents a wide spectrum of research areas, the
@@
 number of publications per year. For each of the papers identified, we
-tested whether (i) the authors used the VD to show differentially expressed
+checked whether (i) the authors used the VD to show differentially expressed
 transcripts significant in one comparison, but not another, (ii) the
@@
 analysis is any form. Articles which (i) focused only on the intersections
-of the Venn diagrams (genes common to all conditions), (ii) which used the
+of the Venn diagrams (genes common to all conditions), or (ii) which used the
 Venn diagrams for a purpose other than to compare genes significant in one
 condition, but not significant in another condition or (iii) for which a
-clear-cut error could not be spotted were not considered incorrect.
+clear-cut error could not be indicated past any reasonable doubt were not
+considered incorrect.
+
+## Bibliography

### 2021-11-19 mk2/manuscript.rmd
small fixes

@@
 ---
-title: "Erroneous use of Venn diagram leads to artifacts in transcriptomic analyses"
+title: "Erroneous use of Venn diagrams leads to artifacts in transcriptomic analyses"
 author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
@@
     reference_docx: templates/template.docx
+outputoff2: 
   pdf_document:
     latex_engine: xelatex
+    keep_tex: true
 bibliography: manuscript.bib
@@
 of Venn diagrams to find genes which are thought to be specific for a
-certain comparison, as well as gene set enrichment analysis applied to
-such subsets is a fallacy. Moreover, the combination of incorrect statistical reasoning
-and gene set enrichment analysis may result in particularly misleading artifacts.
+certain comparison, as well as gene set enrichment analysis applied to such
+subsets is a fallacy. Since genes which show a significant change in one
+condition but not in another are likely to be false negatives in the latter
+case, these genes considered to be specific for one condition are likely
+characteristic for both of the compared groups, and enriched in functions
+related to the research hypothesis. Thus, the combination of incorrect
+statistical reasoning and gene set enrichment analysis may result in
+particularly misleading artifacts.
 

### 2021-12-30 mk3/manuscript.rmd
added revision 3

--- 2021-11-19 mk2/manuscript.rmd
@@
 ---
-title: "Erroneous use of Venn diagram leads to artifacts in transcriptomic analyses"
+title: "Venn diagrams may indicate erroneous statistical approaches leading to artifacts in transcriptomic analyses"
 author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
 date: "`r Sys.Date()`"
+output:
+  html_document
 outputoff:
-  html_document
-output:
   word_document:
     reference_docx: templates/template.docx
+outputoff2: 
   pdf_document:
     latex_engine: xelatex
+    keep_tex: true
 bibliography: manuscript.bib
@@
 of Venn diagrams to find genes which are thought to be specific for a
-certain comparison, as well as gene set enrichment analysis applied to
-such subsets is a fallacy. Moreover, the combination of incorrect statistical reasoning
-and gene set enrichment analysis may result in particularly misleading artifacts.
+certain comparison, as well as gene set enrichment analysis applied to such
+subsets is a fallacy.  A statistically correct approach involves testing for interaction.
+Moreover, since genes which show a significant change in one condition but
+not in another are likely to be false negatives in the latter case, these
+genes considered to be specific for one condition are likely characteristic
+for both of the compared groups, and enriched in functions related to the
+research hypothesis. Thus, the combination of incorrect statistical
+analysis and gene set enrichment analysis may result in particularly
+misleading artifacts.
 
@@
 library(Biobase)
+library(pander)
 library(org.Hs.eg.db)
+library(pwr)
 theme_set(theme_minimal() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()))
@@
 of mice upon infection, showing which genes are regulated in both strains,
-and which are regulated only in one of them.  The VD serves
-as a basis for the statement that the regulation of certain genes is
-"specific" for one strain or another. We argue that not only is this
+and which are regulated only in one of them.  The VD serves as a basis for
+the statement that the regulation of certain genes is "specific" for one
+strain or another. A gene is, in this context, considered "specific" or
+"unique" for a given comparison when it shows a statistically significant
+difference in that comparison only.  We argue that not only is this
 inference incorrect, but it may also lead to misleading – although
@@
 
-Consider two group of patients, G1 and G2. Each group contains 
+Consider two group of patients, G1 and G2 (Tab. 1). Each group contains 
 `r n_stratum * 4`
@@
 incorrect and the correct one) which arrive at opposite conclusions.
+
+**Table 1** Overall design in the case study: transcriptomic changes due to
+Sars-Cov-2 infection. The table shows number of patients in each
+combination of study group / disease status.
+
+```{r results="markdown"}
+tab <- as.data.frame(matrix(2 * n_stratum, ncol=2, nrow=2))
+tab <- data.frame(c("", "Disease status", ""), c("", "Sars-Cov-2 infection", "Another infection"), 
+  c("Group 1 (G1)", tab[[1]]), c("Group 2 (G2)", tab[[2]]))
+
+colnames(tab) <- c("", "", "Study group", "")
+pander(tab)
+```
 
@@
 pathways are upregulated in each of the two groups, we used a standard generation I
-gene set enrichment analysis (GSEA) – a hypergeometric test – on the DEGs in each
-group. Gene sets for GSEA were taken from the Gene Ontology (GO) database.
+gene set enrichment analysis – a hypergeometric test – on the DEGs in each
+group. Gene sets for the gene set enrichment analysis were taken from the Gene Ontology (GO) database.
 Gene sets with more than `r max_mod_size` or 
@@
 ```
+
 Consequently, it is not possible to calculate gene set enrichment for the
@@
 
-```{r lit_surves}
-links <- system2("grep", args=c("https", "literature_survey.md"), stdout=TRUE)
-n_tot <- length(links)
-n_vd_only <- length(grep("vd only", links))
+```{r lit_surveys}
+get_keywords <- function(file, pattern="^([a-zA-Z]+): *(.*)") {
+  ll <- readLines(file)
+  ll <- ll[ grepl(pattern, ll) & !grepl("^https://", ll) ]
+
+  ret <- gsub(pattern, "\\2", ll)
+  names(ret) <- gsub(pattern, "\\1", ll)
+  ret <- as.list(ret)
+  ret$file <- file
+  ret$incorrect_count <- get_incorrect_counts(file)
+  ret$incorrect_count_vdonly <- get_incorrect_counts(file, vdonly=TRUE)
+  ret
+}
+
+get_incorrect_counts <- function(file, vdonly=FALSE) {
+
+  ll <- readLines(file)
+
+  pat <- "^https://"
+  if(vdonly) {
+    sum(grepl(pat, ll) & grepl("vd only", ll))
+  } else {
+    sum(grepl(pat, ll))
+  }
+
+}
+
+files <- list.files(pattern="literature_survey.*.md")
+
+finfo <- map(files, get_keywords) 
+names(finfo) <- map_chr(finfo, ~ .x[["journal"]])
 ```
 
-In an informal survey of articles from the journal "Scientific reports"
-(chosen for its broad scope, large number of published articles and open
-access), of the 238 articles from 2021 which used the term "venn diagram" and
+**Tab. 2** Results of the informal literature survey. 
+We searched for papers using Google Scholar and the keywords "venn diagram"
+and "differential expression".
+*Journal*, journal title; *Years*, publication dates; *Total*, total number
+of papers found using the search phrase; *Analysed*, total number of papers
+which have been analysed for correctness; *Incorrect*, total number
+(percentage) of
+papers in which the Venn diagram was combined with comparing significance
+to non-significance; *Enrichment*, total numebr (percentage) of papers
+which combined the Venn diagram with a gene set enrichment analysis.
+
+```{r results="markdown"}
+tab <- map_dfr(finfo, ~ 
+               data.frame(Journal=.x$journal,
+                          Years=.x$years,
+                          Total=.x$total,
+                          Analysed=.x$analysed,
+                          Incorrect=.x$incorrect_count,
+                          Enrichment=.x$incorrect_count - .x$incorrect_count_vdonly))  %>%
+  mutate(Analysed=as.numeric(Analysed)) %>%
+  mutate(Total=as.numeric(Total))
+
+tab <- rbind(tab, 
+             data.frame(Journal="**Total:**", Years="", Total=sum(tab$Total),
+                        Analysed=sum(tab$Analysed),
+                        Incorrect=sum(tab$Incorrect),
+                        Enrichment=sum(tab$Enrichment)))
+tab2 <- tab %>%
+  mutate(Incorrect=sprintf("%d (%.0f%%)", Incorrect, 100 * Incorrect / Analysed)) %>%
+  mutate(Enrichment=sprintf("%d (%.0f%%)", Enrichment, 100 * Enrichment / Analysed))
+
+pander(tab2)
+```
+
+In an informal survey of articles from three journals (Tab. 2) we found
+that of the 
+`r tab2$Analysed[nrow(tab2)]` analysed articles
+ which used the term "venn diagram" and
 "differential expression", at least 
-`r n_tot`
+`r tab2$Incorrect[nrow(tab2)]`
 were using Venn diagrams to compare
@@
 "specific", "solely regulated" or "exclusive" DEGs. Out of these, at least 
-`r n_vd_only`
+`r tab$Enrichment[nrow(tab2)]`
 coupled the VDs with some form of gene set enrichment analysis on the
@@
 
-As an alternative to Venn diagrams two approaches can be considered.
-Firstly, as shown above, a test for interaction can reveal genes for which
-the impact of treatment significantly differ between the groups. As a mean
-of visualization, we recommend plotting the log~2~ fold changes in one
-comparison versus the log~2~ fold changes in the other comparison. In
-addition, a concordance / discordance analysis may provide additional
-insight in genes which are regulated either in the same direction or in the
-opposite directions in both groups [@domaszewska2017concordant].
+As an alternative to Venn diagrams and the downstream gene set enrichment
+analysis, two approaches can be considered.
 
-While the use of Venn diagrams in combination with the incorrect gene set
+Statistical analysis, as shown above, involves a test for interaction which
+can reveal genes for which the impact of treatment significantly differ
+between the groups. The results can then be plugged into a gene set
+enrichment analysis the usual way.  Unfortunately, this has two major
+drawbacks. Firstly, effect sizes (log~2~ fold changes) of the interaction
+term are harder to interpret than log~2~ fold changes in a direct, group vs
+group comparison.  The effect size in an interaction is negative if the
+log~2~ fold change in the first comparison is larger than the log~2~ fold
+change in the second comparison; this is, however, irrespective of whether
+the differences in the individual comparisons are negative or positive,
+which makes it harder to separate the differences in genes up-regulated in
+one or both groups.
+
+The second problem may arise if the changes are similar in both
+comparisons, but of larger magnitude in one of them. For example, in a time
+series context, the changes may be more pronounced at a later time point.
+In this case, the analysis will show that the processes enriched for the
+interaction term are the same as those enriched in each of the comparisons
+individually.  While the results of the gene set enrichment analysis in
+this context are correct, the result may not be what the researchers
+intended – processes which qualitatively (rather than quantitatively)
+differ between the comparisons.
+
+An alternative approach, discordance/concordance analysis, has been proposed by @domaszewska2017concordant,
+aiming at identifying processes which qualitatively differ between the two
+comparisons. Here, a heuristic score ("disco score") has been defined which depends on the
+effect sizes and p-values in both comparisons. The sign of the score
+depends on whether the effects have the same sign (concordant; genes
+up-regulated in both comparisons or down-regulated in both comparisons) or
+opposite signs (discordant; genes up-regulated in one, but down-regulated
+in the other comparison, and genes down-regulated in one, but up-regulated
+in the other comparison). While the score does not allow the calculation of
+a p-value and does not present an alternative to an analysis of
+interaction, it can facilitate both visualization and further analysis
+using a gene set enrichment algorithm.
+
+Visualization of interaction for individual genes is straightforward (see
+for example Fig. 2, *A-D*). However, the point of VDs is to show a grand
+overview of the whole analysis summarising thousands of results for the
+analysed genes. As an alternative of such an overview, we suggest plotting
+the log~2 fold changes in one comparison against log~2~ fold changes in the
+second comparison. This allows an intuitive assessment of the differences
+between the two comparisons, in especially with combination of color coding
+the genes either are significant in the interaction or using the disco
+score (see Fig. 2, *E* for an example).
+
+While the use of VDs in combination with the incorrect gene set
 enrichment analysis is common, it is not the only way of achieving
@@
 with non-significant results in transcriptomics and other high-throughput
-applications. Our literature survey showed that at least two other ways are
-common. Firstly, the direct comparison of gene set enrichment
-results: that is, drawing conclusions from the fact that a gene set
-enrichment result was significant in one comparison only.
-Second, while Venn diagrams are often used to illustrate the numbers of
-"specific" DEGs and so present a mean to find examples of this fallacy in
-scientific literature, researches often test for enrichment these
-"specific" genes without using the phrase "Venn diagram" or even clearly
-stating how the lists of "specific" genes were derived. In all these cases,
-the analysis boils down to comparing results significant in one, but not in
-another comparison.
+applications. VDs appear to be symptomatic of the assumption that if a gene
+is significantly regulated in one comparison, but does not show a
+significant difference in another one, then this gene is "specific" to this
+first comparison. However, VDs may also lead one to this reasoning; for
+example, the VD on Fig. 1A practically begs the question: what do these 
+`r sum(res.merged$DEG.g1 & !res.merged$DEG.g2)` 
+genes, which are significant in G1 only, have in common?
+
+Our literature survey showed that even in cases where a VD is absent from a
+paper, at least two other ways are not uncommon. Firstly, the direct
+comparison of gene set enrichment results: that is, drawing conclusions
+from the fact that a gene set enrichment result was significant in one
+comparison only.  Second, while Venn diagrams are often used to illustrate
+the numbers of "specific" DEGs and so present a mean to find examples of
+this fallacy in scientific literature, researches often test for enrichment
+these "specific" genes without using the phrase "Venn diagram" or even
+clearly stating how the lists of "specific" genes were derived. In all
+these cases, the analysis boils down to comparing results significant in
+one, but not in another comparison.
 

### 2022-01-03 mk4/montecarlo.R
added revision 4

@@ -1,0 +1,81 @@
+## monte carlo simulation: 100 replicates of the procedure shown in the
+## paper
+
+## output directory
+outdir <- "montecarlo"
+dir.create(path=outdir, showWarnings = FALSE)
+
+## number of replicates
+n_sim <- 100
+
+geo <- readRDS("GSE156063.geo")
+covar <- as(phenoData(geo), "data.frame")
+
+## remove columns with only one value
+boring <- map_lgl(covar, ~ length(unique(.x)) == 1)
+covar <- covar[ , !boring ] 
+
+## clean up
+covar <- covar %>% 
+  dplyr::rename(gender = "gender:ch1") %>%
+  mutate(disease = gsub(" .*", "", .data[["disease state:ch1"]])) %>%
+  mutate(label = description) %>%
+  mutate(group = disease) %>%
+  arrange(description) %>%
+  dplyr::select(all_of(c("title", "label", "gender", "disease", "group")))
+
+counts <- read_csv("GSE156063_swab_gene_counts.csv.gz")
+.tmp <- counts[[1]]
+counts <- as.matrix(counts[,-1])
+rownames(counts) <- .tmp
+#counts <- counts[ , covar$description ]
+counts <- counts[ , covar$label ]
+lcpm   <- edgeR::cpm(counts, log=TRUE)
+#stopifnot(all(colnames(counts) == covar$description))
+stopifnot(all(colnames(counts) == covar$label))
+
+annot <- data.frame(ENSEMBL = rownames(counts))
+
+.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("ENTREZID"), keytype="ENSEMBL")
+annot$ENTREZID <- .tmp[ match(annot$ENSEMBL, names(.tmp)) ]
+
+.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("SYMBOL"), keytype="ENSEMBL")
+annot$SYMBOL <- .tmp[ match(annot$ENSEMBL, names(.tmp)) ]
+
+.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("GENENAME"), keytype="ENSEMBL")
+annot$GENENAME <- .tmp[ match(annot$ENSEMBL, names(.tmp)) ]
+
+sel <- covar$group %in% c("no", "SC2")
+counts <- counts[ , sel ]
+covar  <- covar[ sel, ]
[…]

### 2022-01-05 mk4/response_to_editor.md
updated response to editor

Thanks for your responses to addressing the reviewer's concerns. I would
like to ask the authors to address the following concerns regarding the
paper. These are essential before accepting the paper.

The authors state that "Here we show that both, the use of Venn diagrams to
find genes which are thought to be specific for a certain comparison, as
well as gene set enrichment analysis applied to such subsets is a fallacy.
A statistically correct approach involves testing for interaction."

To start with, no one uses "Venn diagrams to find genes which are thought to be
specific for a certain comparison". It is just a visualization tool.

**We do agree in principle. However, the term "venn diagram analysis" (and
related formulations) is firmly (even though rather unfortunately)
established in the field of differential expression and many authors are
using it essentially as a shorthand for stating that the supposedly
treatment specific genes are found by comparing which genes are significant
in one, but not the other comparison. We have now explained this in more
detail in the Introduction as follows:**

XXX

The flaw
is not in the Venn diagram but in how one could interpret it. Also, applying
gene set enrichment analysis to such subsets could be done even without the use
of Venn diagrams.

**We fully agree; we have already addressed this issue in the
"Discussion".**

The title and abstract are misleading and more journalistic
than scientific.  An analogous example to how the authors presented the paper
would be to condemn the use of a knife for murders.

**We understand that point. However, we do observe a substantial fraction of
papers using Venn diagrams misinterpreting the results. The given analogy
would only work if every fourth use of a knife resulted in a homicide. This
point has already been addressed in the Discussion. Nonetheless we have now
removed references to Venn diagrams in the title and in the abstract.**

This style is misleading
and not acceptable for a scientific paper. The title and abstract should be
completely reformatted to focus on the underlying issue.

**We have completely rewritten the title and the abstract and hope that
they are now acceptable.**

The authors state that "We have noticed a widespread fallacy related to gene
set enrichment analysis" however, no single example (or reference) has been
provided.

**We do not want to single out a paper or point fingers at specific
authors.  However, given that the sources of all our calculations are
available from github, we placed there the list of links to papers which –
in our view – have incorrectly interpreted Venn diagrams. For example, the
file `literature_survey_sci_immunol.md` includes six (out of 14 analysed)
links to papers in Science Immunology which incorrectly interpreted Venn
diagrams. The file `literature_survey_scirep.md` contains 73 links to
papers from Scientific Reports (out of 238 reviewed).**

The literature review results with incorrect use cases should be provided as
supplementary materials to support the claims made in the paper. 
Justification
for the claims made in the paper is essential.

**They have been all along. We have now included the following statement in the
Methods section:**

*"The results of the literature survey (including links to papers classified
as incorrectly analysing the interaction) are included in the manuscript
sources."*

XXX

Authors state that "Of the 431 genes significant in G1, but not in G2, 199
(46%) are significant in the full data set; of the 278 genes significant in G2,
but not in G1, 99 (36%) are significant in the full data set. Given that G1 and
G2 were sampled from the total population, and since the FDR was set to 0.05,
we can assume that at least between a third and a half of the genes that
appeared to be specific" in the initial analysis were, in fact, false negatives
in one of the comparisons." The conclusion made is based on the assumption that
there are no false positives when analyzing the whole dataset. 
This is not
justifiable.

**This conclusion is based on the assumption that the fraction of FP among
significant genes is controlled by the Benjamini-Hochberg procedure, in
other words, that the false discovery rate (FDR) is not more than 5% (as
stated in the quoted paragraph).**

Also, splitting samples of a dataset into two groups results in
smaller sample sizes for each group.

**We agree, but the total number of samples in the analysis is still 80,
with 40 samples per group and 20 per group/treatment combination. This is
not an unusually small sample size for this type of transcriptomic
analysis (see below for more on the chosen sample size). Also, the design of our in
silico experiment *requires* splitting the data set into two randomly selected groups,
because that is the whole idea of the paper: take a situation in which 
there should not be any significant enrichments (because there are no
real differences between the groups) and show that with an incorrect (but
widely spread) procedure you will get convincing evidence for differences
between groups.**

This comes with two major shortcomings:
(1) the test for differential expression of genes are more sensitive to noise
and outliers, resulting in false positives

**We find the claim that the number of false positives is higher for small
sample sizes surprising. Statistical theory tells us that the FDR is
controlled by the BH false discovery rate controlling procedure.  If BH
(which is a widely accepted method for the analysis of DE data) is correct,
we should not expect more than 5% FP among significant results
independently of sample size.  See Benjamini, Yoav, and Yosef Hochberg.
"Controlling the false discovery rate: a practical and powerful approach to
multiple testing." Journal of the Royal statistical society: series B
(Methodological) 57.1 (1995): 289-300. We are not aware of any sources that
would claim otherwise.**

, and (2) the test for differential
expression of genes might not have the power to reject the null hypothesis (no
differential expression), resulting in introducing false negatives.

**We fully agree that FNs are the core of the problem! However, we know
from other studies (and basic statistics) that even in most favorable
circumstances the FNR will likely be above 20%; and that in case of transcriptomic
studies, the FNR may be much larger (please see the relevant fragment already
included in the paper). In other words, it does not matter whether one
looks at 80 samples or 193 samples: FNs are bound to occur. That is the
reason – we claim – for the fact that the gene set enrichments found in
this manner are related to the studied problem. However, we have addressed
this isssue by testing different sample sizes (see below).**

Therefore,
the design of the experiment is biased in favour of the "desired" conclusion.

**We have mimicked an RNA-Seq experiment using a sample size
well above the median used by others precisely to avoid this allegation. We
have now included a relevant reference (Baccarella et al.) which includes a
literature survey on RNA sample size in real world data. In there you will
see that half of the studies used sample sizes sizes of 6 or less per
group.

Nonetheless, we think that showing how the artifacts depend on sample size
is, in itself, interesting, not least because it shows that sample size has
a direct effect on the number of observed artifacts: the larger the sample
size, the more artifacts are observed. We have now included a section
comparing the results for 10, 20 (our example) and 40 samples per
group/treatment combination (thus, total sample sizes of 40, 80 and 160).
The reason for this is, of course, that the more DEGs you observe, the more
powerfull the hypergeometric test is.

For each sample size, we have run 100 replicates of the group sampling,
replicated the whole procedure and gathered all test results. Results are
shown on the new Fig. 3, C and D.

However, we also think that the core of the problem is the incorrect
statistical procedure (selecting genes by comparing significance with lack
thereof) which will not become more correct by increasing the sample size,
even should we fail to observe any artifacts (which is not the case,
artifacts appear even for abs(lfc) >= 3).**

XXX

The authors use a small log-fold change value of 1 to define the
differential expression. This leads to a high number of false positives.

**Statistical theory (e.g. the above reference to the original Benjamini &
Hochberg paper) tells us that the false discovery rate in differential
expression analysis is controlled by the BH procedure, not by the log fold
change threshold, except incidentally (beause log fold change is correlated
to the p-value, thus setting a more conservative log2 fold change results
in eliminating most of the larger p-values). Genes with large fold changes
may still not be significant and vice versa, true positives may have small
log2 fold changes. Therefore you cannot predict what will be the precise
effect of raising the LFC threshold on FDR except in very broad terms,
whereas setting the FDR threshold gives you clear information about the
expected proportion of false positives in your results.

This is analogous to using an effect size filter (say, Cohen's d) on
results of a t-test. The p-value from the t-test controls the type I error
rate, but a statistically significant difference does not necessarily
indicate a biologically meaningfull effect. For this, we turn to the effect
size (see for example Sullivan, Gail M., and Richard Feinn. "Using effect
size—or why the P value is not enough." Journal of graduate medical
education 4.3 (2012): 279-282.).

Thus, the threshold serves to eliminate genes that do show statistically
significant change (even if they are true positives), but which are not
biologically relevant (because the effect size, for which log2 fold change
is a proxy, is too small). The proportion of false positives is 5% even with no
log2 FC treshold.  Nonetheless, we have compared the effect of the log2
fold change threshold on the results.**

Higher values of log-fold change should also be used, and the results
should be compared and discussed.

**We have now included a section in the manuscript which shows the
dependence of the artifacts on the log2 fold change thresholds used.
For this, we have replicated the whole procedure of group randomization 100
times, and repeated the analysis for different threshold values. Summary
overview can be found in the new Fig. 3.**

The paper also needs to be edited to have a more scientific tone, improve long and complicated sentences, and correct the typos and grammatical errors.

**We thank the Editor for this suggestion. We did our best, and we hope
that the manuscript reads now much better.**

### 2022-01-05 mk4/manuscript.rmd
added revision 4
updated response to editor

--- 2021-12-30 mk3/manuscript.rmd
@@
 lfc_thr  <- 1
+n_stratum <- 20 # 2 strata per condition 
 ```
@@
 ### Transcriptomic changes due to Sars-Cov-2 infection
-
-```{r}
-n_stratum <- 10 # 2 strata per condition 
-```
 
@@
 
-```{r Preparation_of_the_covariates,cache=TRUE}
+```{r Preparation_of_the_covariates,cache=FALSE}
 sel <- covar$group %in% c("no", "SC2")
@@
 
-set.seed(0123)
+set.seed(20220105)
 
@@
 
-#g1 <- sample(1:nrow(covar), 30)
-#g2 <- sample(setdiff(1:nrow(covar), g1), 30)
-
 covar$group <- NA
@@
 
-sel <- c(g1, g2)
-
-ds2 <- DESeqDataSetFromMatrix(counts[,sel], colData=covar[sel, ],
-  design=~ 0 + group.disease )
 ```
 
-```{r DESeq2,cache=TRUE}
+```{r DESeq2,cache=FALSE}
 ## DESeq2 calculations
+## manual cache, since the operation takes a long time
+sel <- !is.na(covar$group)
 
+ds2 <- DESeqDataSetFromMatrix(counts[,sel], colData=covar[sel, ], design=~ 0 + group.disease )
+
+ds2_file <- "ds2_cached.rds"
+if(!file.exists(ds2_file)) {
+  message("Running DESeq2")
   ds2 <- DESeq(ds2)
-saveRDS(ds2, file="ds2_cached.rds")
+  saveRDS(ds2, file=ds2_file)
+} else {
+  message("Reading ds2 from manual cache")
+  ds2 <- readRDS(ds2_file)
+}
 ```
@@
 (differentially expressed genes, DEGs) 
-by using an FDR threshold of 
+by using an false discovery rate (FDR) threshold of 
 `r padj_thr`
@@
 
+### Influence of sample size and cut-off thresholds on number of artifacts
+
+In the example above, the groups have been randomly sampled from a larger
+data set only once. To investigate whether the sampling had an impact on
+the artifacts, we have repeated the above procedure for 100 replicates,
+each containing a different set of samples randomly assigned to the two
+groups.  Furthermore, we tested whether the selection of the log~2~ fold
+change threshold
+or different sample size might have an impact on the extent of the arising
+artifacts. To this end, we have tested `r length(lfc_thresholds)` different
+threshold values for the log~2~ fold changes and three sample sizes: 20, 40 or 80 samples
+per group (corresponding to 10, 20 or 40 samples per group/treatment
+combination).
+
+**Fig. 3.Influence of log~2~ fold change treshold (A, C) and the sample
+size (B, D) on the number of genes significant in one, but not in the other
+group (A, B) and the number of significant gene sets found in one group, but
+not the other (C, D).**  Top row shows the influence of log~2~ fold change
+threshold for sample size of 40 per group (20 per group/treatment
+combination), bottom row shows the influence of sample size for a log~2~
+fold change threshold set to `r lfc_thr`.
+
+```{r montecarlo,cache=TRUE,fig.width=5.5,fig.height=5.5}
+## note: given that the monte carlo simulation is quite time consuming, we
+## have included the code for it in another file, `montecarlo.R`. If trying
+## to replicate this manuscript, please execute that file first to generate
+## the necessary replicates.
+
+outdir <- "montecarlo"
+
+merged_res <- readRDS(file.path(outdir, "merged_res.rds"))
+
+lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
+
+by_lfc_thr <- map_dfc(lfc_thresholds, ~ 
+                      get_specific_num(merged_res[["10"]], lfc_thr=.x,
+                                       padj_thr=padj_thr))
+colnames(by_lfc_thr) <- lfc_thresholds
+by_lfc_thr <- by_lfc_thr %>% pivot_longer(everything(), names_to = "lfc")
+
+g1 <- ggplot(by_lfc_thr, aes(x=lfc, y=value)) + geom_boxplot() +
+  ylab("Number of DEGs") + xlab("log2 fold change threshold")
+
+gsea_files <- list.files(path=outdir, pattern="gsea_.*rds", full.names=TRUE)
+gsea_by_lfc_thr_files <- gsea_files[ grepl("gsea_res_10", gsea_files) ]
+names(gsea_by_lfc_thr_files) <- gsub("montecarlo/gsea_res_10_([0-9.]*)\\.rds", "\\1",
+                                     gsea_by_lfc_thr_files)
+gsea_by_lfc_thr_files <- gsea_by_lfc_thr_files[ order(as.numeric(names(gsea_by_lfc_thr_files))) ]
+gsea_by_lfc_thr <- map_dfc(gsea_by_lfc_thr_files, gsea_specific_num)
+gsea_by_lfc_thr <- gsea_by_lfc_thr %>% pivot_longer(everything(), names_to = "lfc")
+
+g2 <- ggplot(gsea_by_lfc_thr, aes(x=lfc, y=value)) + geom_boxplot() +
+  ylab("Number of enriched gene sets") + xlab("log2 fold change threshold")
+
+by_ssize <- map_dfc(c("5", "10", "20"), 
+                    ~ get_specific_num(merged_res[[.x]], lfc_thr=1,
+                                       padj_thr=padj_thr))
+colnames(by_ssize) <- c(10, 20, 40)
+by_ssize <- by_ssize %>% pivot_longer(everything(), names_to = "ss")
+
+g3 <- ggplot(by_ssize, aes(x=ss, y=value)) + geom_boxplot() +
+  ylab("Number of DEGs") + xlab("Number of samples per group/treatment")
+
+gsea_by_ss_files <- gsea_files[ grepl("gsea_res_[0-9]*_1\\.rds", gsea_files) ]
+names(gsea_by_ss_files) <- gsub("montecarlo/gsea_res_([0-9]*)_.*\\.rds", "\\1",
+                                     gsea_by_ss_files)
+gsea_by_ss_files <- gsea_by_ss_files[ order(as.numeric(names(gsea_by_ss_files))) ]
+gsea_by_ss <- imap(gsea_by_ss_files, ~ {
+                     data.frame(ss=.y, value= gsea_specific_num(.x))
+                                     })
+gsea_by_ss <- Reduce(rbind, gsea_by_ss)
+gsea_by_ss$ss <- factor(as.numeric(gsea_by_ss$ss) * 2, levels=c(10, 20, 40))
+
+g4 <- ggplot(gsea_by_ss, aes(x=ss, y=value)) + geom_boxplot() +
+  ylab("Number of enriched gene sets") + xlab("Number of samples per group/treatment")
+
+plot_grid(g1, g2, g3, g4, labels = LETTERS[1:4])
+```
+
 ### Artifacts arise because of false negatives
@@
 To illustrate this phenomenon, we have analysed the full data set from which G1 and G2
-were drawn (Fig. 3), comparing the 
+were drawn (Fig. 4), comparing the 
 `r sum(covar$disease == "no") `
@@
 
-**Fig. 3. Area-proportional Venn diagram showing overlaps in DEGs between
+**Fig. 4. Area-proportional Venn diagram showing overlaps in DEGs between
 G1, G2 and the full data set.** The majority of genes which have been labeled

### 2022-01-05 mk4/functions.R

@@ -79,5 +79,31 @@ ggplot_disco <- function(df,
 }
 
 
+## return the number of "specific" enrichments in the given file
+gsea_specific_num <- function(file_name, gsea_auc_thr=0.5, gsea_pval_thr=0.05) {
+
+  res <- readRDS(file_name)
+
+  map_int(res, ~ {
+    n1 <- .x %>% filter(adj.P.Val.g1 < gsea_pval_thr) %>% nrow()
+    n2 <- .x %>% filter(adj.P.Val.g2 < gsea_pval_thr) %>% nrow()
+    n1 + n2
+    })
+
+}
+
+## returns a numeric vector. Each number represents the number of
+## "specific" genes (in both groups) for one replicate.
+get_specific_num <- function(merged_res, lfc_thr=0, padj_thr=0.05) {
+
+  n_spec <- map_int(merged_res, ~ {
+    res <- .x %>%
+      mutate(DEG.g1 = !is.na(padj.g1) & abs(log2FoldChange.g1) > lfc_thr & padj.g1 < padj_thr) %>%
+      mutate(DEG.g2 = !is.na(padj.g2) & abs(log2FoldChange.g2) > lfc_thr & padj.g2 < padj_thr) 
+    sum((res$DEG.g1 & !res$DEG.g2) | (!res$DEG.g1 & res$DEG.g2))
+  })
+
+  n_spec
+}
 
 

### 2022-01-05 mk4/montecarlo.R

@@ -8,6 +8,17 @@ dir.create(path=outdir, showWarnings = FALSE)
 ## number of replicates
 n_sim <- 100
 
+## number of replicates per stratum (2 x n_stratum for one treatment/group)
+## total sample size is 8 x n_stratum
+n_stratum <- c(5, 10, 20)
+
+## thresholds for the analysis
+padj_thr <- 0.05
+lfc_thr  <- 1.5
+
+## lfc thresholds to test
+lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
+
 geo <- readRDS("GSE156063.geo")
 covar <- as(phenoData(geo), "data.frame")
 
@@ -50,7 +61,7 @@ counts <- counts[ , sel ]
 covar  <- covar[ sel, ]
 covar$group <- as.character(covar$group)
 
-one_replicate_ds2 <- function(covar, counts) {
+one_replicate_ds2 <- function(covar, counts, n_stratum) {
   g1 <- covar %>% mutate(n=1:n()) %>%
     group_by(gender, disease) %>% slice_sample(n=n_stratum) %>% pull(n)
   g2 <- covar %>% mutate(n=1:n()) %>% filter(!n %in% g1) %>%
@@ -72,10 +83,99 @@ one_replicate_ds2 <- function(covar, counts) {
 
 
 ## save the DS2 objects from each replication
-for(i in 1:n_sim) {
-  ds2 <- one_replicate_ds2(covar, counts)
+for(nstrat in n_stratum) {
+  # 2 x nstrat per group/treatment, 4 x nstrat per group
+  message(sprintf("Group size %d", nstrat * 2))
+  for(i in 1:n_sim) {
 
-  file_name <- file.path(outdir, sprintf("ds2_%d.rds", i))
-  saveRDS(ds2, file=file_name)
+    message(i)
+    ds2_file_name <- file.path(outdir, sprintf("ds2_%d_%d.rds", nstrat, i))
+
+    if(!file.exists(ds2_file_name)) {
+      ds2 <- one_replicate_ds2(covar, counts, nstrat)
+      saveRDS(ds2, file=ds2_file_name)
+    }
+
+  }
 }
+
+ds2_files <- list.files(path = outdir, pattern = "ds2.*\\.rds", full.names = TRUE)
+
+## get the results lists for the two contrasts (G1 and G2, respectively)
[…]

### 2022-01-06 mk4/response_to_editor.md
added presentation

@@
 The reason for this is, of course, that the more DEGs you observe, the more
-powerfull the hypergeometric test is.
+powerful the hypergeometric test is.
 
@@
 statistical procedure (selecting genes by comparing significance with lack
-thereof) which will not become more correct by increasing the sample size,
-even should we fail to observe any artifacts (which is not the case,
-artifacts appear even for abs(lfc) >= 3).**
+thereof) which will not become more correct by increasing the sample size.**
 
@@
 is a proxy, is too small). The proportion of false positives is 5% even with no
-log2 FC treshold.  Nonetheless, we have compared the effect of the log2
-fold change threshold on the results.**
+log2 FC treshold.
+
+Increasing the threshold can, of course, result in not observing any
+enrichments (see Fig. 3, A and B). Indeed, it does for some replicates.
+This is not surprising – without a sufficient number of DEGs, no
+enrichments can be calculated using a hypergeometric test. However, even
+though the effect of increasing log2 fold change may sometimes eliminate the
+symptoms, it does not cure the disease – the core problem of erroneous
+analysis of interactions remains.
+
+Nonetheless, we have compared the effect of the log2
+fold change threshold on the results (Fig. 3).**
 
@@
 For this, we have replicated the whole procedure of group randomization 100
-times, and repeated the analysis for different threshold values. Summary
-overview can be found in the new Fig. 3.**
+times. Summary overview can be found in the new Fig. 3. Bottom line: yes,
+for many replicates the artifacts disappear when the log2 fold change threshold is high –
+because the DEGs disappear. However, this is not generally so: even for LFC
+greater or equal 3 and total sample size of 160 the artifacts are still
+present in 78 out of 100 replicates.**
 

### 2022-01-06 presentation/manuscript.rmd
added presentation

--- 2022-01-05 mk4/manuscript.rmd
@@
 ---
-title: "Venn diagrams may indicate erroneous statistical approaches leading to artifacts in transcriptomic analyses"
+title: ""
 author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
@@
 lfc_thr  <- 1
-n_stratum <- 20 # 2 strata per condition 
+n_stratum <- 10 # 2 strata (male and female) per condition 
+lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
 ```
@@
 
-set.seed(20220105)
+set.seed(0123)
 
@@
 ds2_file <- "ds2_cached.rds"
-if(!file.exists(ds2_file)) {
+if(TRUE || !file.exists(ds2_file)) {
   message("Running DESeq2")
@@
 
-```{r GSEA}
+```{r GSEA,cache=FALSE}
 common <- res.merged %>% filter(DEG.g1 & DEG.g2) %>% pull(SYMBOL)
@@
 })
+message("G1 ", nrow(gsea_res$g1), " G2 ", nrow(gsea_res$g2))
 gsea_res.merged <- merge(gsea_res$g1, gsea_res$g2, by=c("ID", "Title"), suffixes=c(".g1", ".g2"), all=T)
+saveRDS(gsea_res.merged, file="gsea_res_merged.rds")
 ```
@@
 In the example above, the groups have been randomly sampled from a larger
-data set only once. To investigate whether the sampling had an impact on
+data set only once. Arguably, the observations might differ if the groups
+were to be resampled. To investigate whether the sampling had an impact on
 the artifacts, we have repeated the above procedure for 100 replicates,
@@
 groups.  Furthermore, we tested whether the selection of the log~2~ fold
-change threshold
-or different sample size might have an impact on the extent of the arising
-artifacts. To this end, we have tested `r length(lfc_thresholds)` different
-threshold values for the log~2~ fold changes and three sample sizes: 20, 40 or 80 samples
-per group (corresponding to 10, 20 or 40 samples per group/treatment
-combination).
+change threshold or different sample size might have an impact on the
+extent of the arising artifacts. To this end, we have tested `r
+length(lfc_thresholds)` different threshold values for the log~2~ fold
+changes and three sample sizes: 20, 40 or 80 samples per group
+(corresponding to 10, 20 or 40 samples per group/treatment combination).
 
-**Fig. 3.Influence of log~2~ fold change treshold (A, C) and the sample
+**Fig. 3. Influence of log~2~ fold change treshold (A, C) and the sample
 size (B, D) on the number of genes significant in one, but not in the other
@@
 
-```{r montecarlo,cache=TRUE,fig.width=5.5,fig.height=5.5}
+```{r montecarlo,cache=TRUE,fig.width=7.5,fig.height=6.5}
 ## note: given that the monte carlo simulation is quite time consuming, we
@@
 ```
+
+Setting a higher log~2~ fold change threshold reduces the number of
+DEGs and of observed artifacts (Fig. 3, A and B). However, even for log~2~
+threshold of 3 (DEGs defined by 8-fold change and FDR < `r pval_thr`) the
+number of replicates in which artifacts can be observed is 
+`r sum(gsea_by_lfc_thr %>% filter(lfc == 3) %>% pull(value) > 0)`
+(out of 100),
+and in at least 
+`r sum(gsea_by_lfc_thr %>% filter(lfc == 3) %>% pull(value) > 0)`
+replicates, 5 or more gene sets were significantly
+enriched. Thus, setting a more conservative threshold cannot protect from
+the arising artifacts.
+
+Contrary to our expectations, the number of artifacts rose with sample size
+(Fig. 3, C and D). For total sample size of 160 (40 samples per
+group/treatment combination) the number of artifacts was almost 
+`r round(
+  mean(gsea_by_ss %>% filter(ss == 40) %>% pull(value)) /
+  mean(gsea_by_ss %>% filter(ss == 20) %>% pull(value)))
+`
+times higher than for total sample size of 80.
 

### 2022-01-06 presentation/venn_diagrams_weiner_2022_01_06.rmd
added presentation

--- 2022-01-06 presentation/manuscript.rmd
@@
 ---
-title: ""
-author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
+title: "Venn diagrams indicate erroneous analyses in transcriptomics"
+author: "January Weiner"
 date: "`r Sys.Date()`"
+outputdupa: html_document
 output:
-  html_document
-outputoff:
-  word_document:
-    reference_docx: templates/template.docx
-outputoff2: 
-  pdf_document:
-    latex_engine: xelatex
-    keep_tex: true
-bibliography: manuscript.bib
-link-citations: true
-XXcsl: templates/nature-medicine.csl
+  xaringan::moon_reader:
+    self-contained: true
+    css: ["default", "files/cubi-fonts.css", "files/style.css" ]
+    lib_dir: libs
+    nature:
+      highlightStyle: github
+      countIncrementalSlides: false
+      ratio: "16:9"
+      beforeInit: "files/macros.js"
+toc: no
 ---
 
-^1^ Berlin Institute of Health at Charité – Universitätsmedizin Berlin, Core Unit Bioinformatics, Charitéplatz 1, 10117 Berlin, Germany
+```{r,echo=FALSE}
+## Set default options for the knitr RMD processing
+knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=FALSE,autodep=TRUE, results="hide")
+library(pander)
+library(knitr)
+library(kableExtra)
+library(tidyverse)
+library(tmod)
+library(tidyverse)
+library(limma)
+library(GEOquery)
+library(DESeq2)
+## install bioshmods from github.com/bihealth/bioshmods
+library(bioshmods)
+library(tmod)
+library(cowplot)
+library(stringr)
+library(ggvenn)
+library(ggplot2)
+library(ggpval)
+library(eulerr)
+library(Biobase)
+library(org.Hs.eg.db)
+theme_set(theme_minimal() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()))
+source("functions.R")
+```
 
-^†^ To whom the correspondence should be addressed
+```{r echo=FALSE,include=FALSE,eval=TRUE}
+options(crayon.enabled = TRUE)
+options(crayon.colors = 256)
+knitr::knit_hooks$set(output = function(x, options){
+  paste0(
+    '<pre class="r-output"><code>',
+    fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
+    '</code></pre>'
+  )
+})
 
-## Abstract
-
-Venn diagrams are frequently used to illustrate the results of
-differential expression analysis and further
-downstream functional analysis. Here we show that both, the use
-of Venn diagrams to find genes which are thought to be specific for a
-certain comparison, as well as gene set enrichment analysis applied to such
-subsets is a fallacy.  A statistically correct approach involves testing for interaction.
-Moreover, since genes which show a significant change in one condition but
-not in another are likely to be false negatives in the latter case, these
-genes considered to be specific for one condition are likely characteristic
-for both of the compared groups, and enriched in functions related to the
-research hypothesis. Thus, the combination of incorrect statistical
-analysis and gene set enrichment analysis may result in particularly
-misleading artifacts.
+## this is an ugly, ugly hack, but otherwise crayon does not LISTEN TO REASON!!!
+num_colors <- function(forget=TRUE) 256
+library(crayon)
+assignInNamespace("num_colors", num_colors, pos="package:crayon")
+```
 
@@
 lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
-```
-
-## Introduction: Venn diagrams are a showcase of erroneous interpretation of interactions
-
-Venn diagrams (VDs) are commonly used to visualize high throughput data
-such as transcriptomic profiles. Frequently, a Venn diagram serves the
-purpose of comparing the results from two experiments or two comparisons.
-For example, a VD may illustrate up- or down-regulated genes in two strains
-of mice upon infection, showing which genes are regulated in both strains,
-and which are regulated only in one of them.  The VD serves as a basis for
-the statement that the regulation of certain genes is "specific" for one
-strain or another. A gene is, in this context, considered "specific" or
-"unique" for a given comparison when it shows a statistically significant
-difference in that comparison only.  We argue that not only is this
-inference incorrect, but it may also lead to misleading – although
-appealing – artifacts when combined with downstream analyses.
-
-Consider a gene for which the expression has been analysed in four groups:
-two different mouse strains (wildtype, WT and knockout, KO) and two different experimental conditions
-(naive versus infected). We find that the gene expression significantly
-differs between infected and naive KO animals, but that there is no
-significant difference in the WT strain. Such a gene may be incorrectly
-considered as "specific" to KO, and will be accordingly entered in a Venn
-diagram.
-
-However, this is a common fallacy [@nieuwenhuis2011erroneous], since lack
-of statistical significance is not the same as testing for lack of
-difference. In other words, the fact that we failed to detect a significant
-difference in the WT does not mean that the difference is absent and
-itself significantly different from the difference in the KO. This "difference of
-differences" is known is statistics as interaction [@blalock1965theory].
-Here, it is the interaction between strain and treatment.  In fact, the
-obtained p-values might be just over the assumed p-value threshold in one, but just
-under it in the other strain (e.g. 0.009 vs 0.011), with the difference in
-both strains being essentially the same. The correct analysis is to test
-the significance of interaction using an ANOVA model. Notably, in 2011
-Nieuwenhuis et al. found that in over 150 papers in top neuroscience
-journals where the authors could make this particular statistical error,
-half of them did.
-
-Using VDs to show genes "specific" for a condition amounts to
-counting the times a comparison for a gene was statistically significant in
-one condition, but not significant in another.
-More disturbingly, VDs are a visual illustration of a procedure
-that exacerbates this problem by applying a downstream analysis to
-supposedly specific sets of genes. Several cells in a Venn diagram contain
-numbers corresponding to features (such as genes), for which there was a significant test result in
-one, but not in another comparison. The sets of genes in a Venn diagram cell may
-then be subsequently analysed to test whether they share a particular
-characteristics.  For example, gene set enrichment analysis may be used to
-interpret the biological function of genes that are considered "specific" to one
-condition. It turns out that due to the
-fallacious nature of this procedure, it is likely to produce results that
-appear to make sense in the given biological context.
-
-In this paper, we illustrate this rather simple statistical statement with
-a real world data set, demonstrating how choosing this sort of approach
-results in apparently sound gene set enrichment results which are, in fact,
-artifacts. Next, we dissect the underlying mechanism of how these artifacts
-are generated.  Finally, we discuss alternative approaches.
-
-## Results
-
-### Transcriptomic changes due to Sars-Cov-2 infection
-
-Consider two group of patients, G1 and G2 (Tab. 1). Each group contains 
-`r n_stratum * 4`
-individuals. In both groups, there is an equal number of healthy
-individuals (labeled "Ctrl" on figures below) 
-or patients infected with Sars-Cov-2 (labeled "SC2"). Our aim is to understand
-the differences between G1 and G2 in the response to infection. For
-example, we ask which genes or pathways are specifically upregulated by SC2
-infection in G1 as compared to G2, and vice versa. In the following, we
-used the data set GSE156063 [@mick2020upper] in two approaches (an
-incorrect and the correct one) which arrive at opposite conclusions.
-
-**Table 1** Overall design in the case study: transcriptomic changes due to
-Sars-Cov-2 infection. The table shows number of patients in each
-combination of study group / disease status.
-
-```{r results="markdown"}
-tab <- as.data.frame(matrix(2 * n_stratum, ncol=2, nrow=2))
-tab <- data.frame(c("", "Disease status", ""), c("", "Sars-Cov-2 infection", "Another infection"), 
-  c("Group 1 (G1)", tab[[1]]), c("Group 2 (G2)", tab[[2]]))
-
-colnames(tab) <- c("", "", "Study group", "")
-pander(tab)
 ```
@@
 
-```{r DESeq2,cache=FALSE}
+```{r DESeq2,cache=TRUE}
 ## DESeq2 calculations
@@
 
-First, we have performed differential gene expression analysis for each of
-the groups G1 and G2 separately using standard bioinformatic tools. For each comparison, we defined
-genes with significantly different gene expression 
-(differentially expressed genes, DEGs) 
-by using an false discovery rate (FDR) threshold of 
-`r padj_thr`
-and absolute log~2~ fold change (LFC) threshold of
-`r lfc_thr`. There were
-`r sum(res$g1$DEG)` 
-DE genes in the G1 group, and
-`r sum(res$g2$DEG)`
-in the G2 group. In total,
-`r sum(res.merged$DEG.g1 & res.merged$DEG.g2)` 
-DEGs were common for G1 and G2, 
-`r sum(res.merged$DEG.g1 & !res.merged$DEG.g2)` 
-DEGs were significant in G1 only ("specific" for G1), and 
-`r sum(!res.merged$DEG.g1 & res.merged$DEG.g2)` 
-were significant in G2 only ("specific" for G2; see Fig. 1, panel *A*).
-A naive interpretation of these results implies that there is a substantial
-difference between these two groups of individuals, as evidenced by a small
-overlap in commonly regulated genes, while the majority of DEGs is
-significant in one comparison only.
-
 ```{r GO_db_prepare,cache=TRUE}
@@
 
-```{r GSEA,cache=FALSE}
+```{r GSEA,cache=TRUE}
 common <- res.merged %>% filter(DEG.g1 & DEG.g2) %>% pull(SYMBOL)
@@
 ```
-
-To understand which
-pathways are upregulated in each of the two groups, we used a standard generation I
-gene set enrichment analysis – a hypergeometric test – on the DEGs in each
-group. Gene sets for the gene set enrichment analysis were taken from the Gene Ontology (GO) database.
-Gene sets with more than `r max_mod_size` or 
-fewer than `r min_mod_size` genes were removed.
-For each group, we have
-selected only genes which are DEGs in that group, but not the other,
-mimicking a naive approach for finding "specifically" regulated pathways.
-Here, a similar picture
-emerged. Just
-`r nrow(gsea_res$g1)` gene sets were significantly enriched in G1, and
-`r nrow(gsea_res$g2)` gene sets were significantly enriched in G2. Again,
-both the Venn diagram (Fig. 1, B) and the results of
-enrichments (Fig. 1, C and D) suggest that there is a
-fundamental difference between the groups, and that the groups have little
-in common in their response to the virus.
-
-Importantly, the different GO terms enriched in the two groups were related
-to infectious diseases, and may tempt to speculate about the underlying
-biological differences between these two groups (e.g.  significance of
-Toll like receptor 4 pathway in G1, but not G2; and, vice versa, significance of
-response to interleukin 7 in G2, but not in G1).
-
-**Fig. 1. Results of differential gene expression analysis
-and gene set enrichment analysis using an incorrect Venn diagram driven approach.**
-**A**, Venn diagram showing numbers of differentially expressed genes (DEG)
-in each of the two groups, G1 and G2; **B**, Venn diagram showing numbers
-of significantly enriched GO terms in each of the two groups; **C** results
-of gene set enrichment analysis for genes "specific" to group G1; **D**, results of gene set
-enrichment analysis for genes "specific" to group G2 (only top 10 terms are shown).
-
-```{r fig1, fig.width=15,fig.height=5.5,dpi=300,dev="png"}
-p1 <- ggvenn(list(G1=res$g1 %>% filter(DEG) %>% pull(ENSEMBL),
-                         G2=res$g2 %>% filter(DEG) %>% pull(ENSEMBL)),
-             show_percentage=FALSE) +
-      ggtitle("Differentially expressed genes")
-p2 <- ggvenn(list(G1=gsea_res$g1 %>% pull(ID),
-                         G2=gsea_res$g2 %>% pull(ID)),
-              show_percentage=FALSE) +
-      ggtitle("Enriched GO terms")
-
-col1 <- plot_grid(p1, p2, labels="AUTO", nrow=2)
-
-p3 <- plot_gsea(gsea_res$g1 %>% arrange(P.Value) %>% dplyr::slice(1:10)) + ggtitle("G1") +
-  scale_x_discrete(labels=function(x) str_wrap(x, width=25))
-#p3 <- plot_grid(p3, NULL, ncol=1, rel_heights=c(2, 1))
-p4 <- plot_gsea(gsea_res$g2 %>% arrange(P.Value) %>% dplyr::slice(1:10)) + ggtitle("G2") +
-  scale_x_discrete(labels=function(x) str_wrap(x, width=25))
-
-plot_grid(plotlist = list(col1, p3, p4), labels=c('', 'C', 'D'), nrow=1)
-```
-
-The groups G1 and G2 were randomly sampled from the same data set.
-In fact, repeated re-sampling always results in some genes being found to
-be significantly different in one group, but not the other, despite the
-fact that one does not expect any major differences between sets of
-individuals randomly drawn from one population. Thus, the conclusions drawn
-from a Venn diagram-driven gene set enrichment analysis are artefactual. Closer
-inspection of genes which belong to DEG in one group, but not the other
-reveals the underlying statistical fallacy (Fig. 2, *A* – *D*), that is, that difference
-between significant and non-significant is, in itself, not statistically
-significant [@gelman2006difference]. This does not necessarily mean that
-there are no differences at all between these two groups, but that lack of
-significance in one group and significance in the other group does not
-correctly identify differences between groups.
 
@@
 
-To find genes which are differentially regulated in the two groups, the
-correct statistical approach is to calculate interaction between groups
-(G1, G2) and disease status (no disease vs. COVID). While it may be argued
-that a test for interaction has lower power than a test for a simple
-contrast, no 
-genes show a significant interaction even at FDR < 0.1. In fact, this is not
-surprising. The log~2~ fold changes for comparisons withing G1 and G2 are
-strongly correlated (Fig. 2, *E*). For all significant genes, the Pearson correlation
-coefficient is
-`r format.pval(cors$sign, digits=2)`, 
-while for genes
-exclusively significant in G1 or G2 (genes "specific" to G1 or G2), it is
-`r format.pval(cors$g1, digits=2)` 
-and
-`r format.pval(cors$g2, digits=2)`, 
-respectively. Thus, genes which are significant in one, but not in the
-other comparison often have have similar log~2~ fold changes (e.g. Fig.
-2, *A*, *C* and *D*).
+```{r full_dataset,cache=TRUE}
+ds2_full <- DESeqDataSetFromMatrix(counts, colData=covar, design=~ disease )
+ds2_full<- DESeq(ds2_full)
+res$full <- results(ds2_full, name="disease_SC2_vs_no") %>% as.data.frame() %>%
+  rownames_to_column("ENSEMBL") %>% 
+  left_join(annot, by="ENSEMBL") %>%
+  mutate(DEG=!is.na(padj) & abs(log2FoldChange) > lfc_thr & padj < padj_thr)
+```
 
-**Fig. 2. Genes which are significant in one comparison, but not
-the other do not show a statistically significant interaction.** **A** -
-**D**, examples of genes which are DEG in one group, but are not
-significantly different in the other group. "Ctrl", healthy individuals; 
-"SC2", Sars-Cov-2 infected patients. Values above the plot indicate FDR (p-values
-corrected for multiple testing). **E**, correlation between log~2~ fold
-changes in G1 and G2. Color indicates genes which are are significant in
-one, but not significant in the other comparison; red indicates genes significant in G1, 
-blue indicates genes significant in G2. The overall Pearson correlation
-coefficient between log~2~ fold changes is 
-`r with(res.merged, format.pval(cor(log2FoldChange.g1, log2FoldChange.g2), digits=2))`.
+```{r g_specific}
+g1_spec <- res.merged %>% filter(DEG.g1 & !DEG.g2) %>% pull(ENSEMBL)
+g2_spec <- res.merged %>% filter(!DEG.g1 & DEG.g2) %>% pull(ENSEMBL)
+```
 
-```{r fig2, fig.width=10, fig.height=5.5, dpi=300,dev="png"}
+## Absence of evidence is not evidence of absence
+
+![ ](images/absence_evidence_2.webp)
+
+.myfootnote[
+*Nieuwenhuis, Sander, Birte U. Forstmann, and Eric-Jan Wagenmakers.
+"Erroneous analyses of interactions in neuroscience: a problem of
+significance." Nature neuroscience 14.9 (2011): 1105-1107."*
+]
+
+---
+
+## Half of the researchers get it wrong
+
+```{r results="markdown"}
+foo <- read.table(text="Nature\tScience\tNature Neuroscience\tNeuron\tJournal of Neuroscience\tSummed
+Total reviewed\t34\t45\t117\t106\t211\t513
+Correct count\t3\t9\t17\t13\t36\t78
+Error count\t7\t11\t16\t15\t30\t79", sep="\t")
+colnames(foo) <- gsub("\\.", " ", colnames(foo))
+kable(foo)
+```
+
+.myfootnote[
+*Nieuwenhuis, Sander, Birte U. Forstmann, and Eric-Jan Wagenmakers.
+"Erroneous analyses of interactions in neuroscience: a problem of
+significance." Nature neuroscience 14.9 (2011): 1105-1107."*
+]
+
+---
+
+.pull-left[
+
+Consider two group of patients, G1 and G2. Each group contains 
+`r n_stratum * 4`
+individuals.
+
+In each group, there is an equal number of healthy
+individuals (labeled "Ctrl")
+or patients infected with Sars-Cov-2 (labeled "SC2"). Our aim is to understand
+the differences between G1 and G2 in the response to infection.
+
+]
+
+.myfootnote[
+Weiner, Obermayer & Beule, "Erroneous use of Venn diagrams leads to
+artifacts in transcriptomic analyses", submitted.
+]
+
+---
+
+**Table 1** Overall design in the case study: transcriptomic changes due to
+Sars-Cov-2 infection. The table shows number of patients in each
+combination of study group / disease status.
+
+```{r results="markdown"}
+tab <- as.data.frame(matrix(2 * n_stratum, ncol=2, nrow=2))
+tab <- data.frame(c("", "Disease status", ""), c("", "Sars-Cov-2 infection", "Another infection"), 
+  c("Group 1 (G1)", tab[[1]]), c("Group 2 (G2)", tab[[2]]))
+
+colnames(tab) <- c("", "", "Study group", "")
+kable(tab)
+```
+
+---
+
+## "Specific" ("unique", "characteristic" genes)
+
+.pull-left[
+
+```{r fig1_1, fig.width=5,fig.height=5,dpi=300,dev="png"}
+p1 <- ggvenn(list(G1=res$g1 %>% filter(DEG) %>% pull(ENSEMBL),
+                         G2=res$g2 %>% filter(DEG) %>% pull(ENSEMBL)),
+             show_percentage=FALSE) +
+      ggtitle("Differentially expressed genes")
+p1
+
+```
+
+]
+
+.pull-right[
+
+  Venn diagrams illustrate a procedure to select genes which are thought to
+  be specific or characteristic for a given comparison.
+
+  **Genes which are significant in one condition, but not the other, are
+  considered "specific".**
+
+  This is incorrect: difference between significant and non-significant is
+  not significant!
+
+]
+
+---
+
+## The groups G1 and G2 do not differ
+
+We have sampled 80 samples from a larger data set and randomly assigned
+them to two group, G1 and G2
+
+---
+
+## The reality
+
+.pull-left[
+
+```{r fig2, fig.width=5, fig.height=5, dpi=300,dev="png"}
 #id    <- gsub(" ", "_", toupper("response to interferon beta"))
@@
 
-plot_grid(part1, part2, rel_widths=c(2, 2), labels=c("", "E"))
+part1
 ```
 
-Consequently, it is not possible to calculate gene set enrichment for the
-interaction using a
-hypergeometric test, as there are no DEGs for the interaction contrast.
-Gene set enrichment using a second generation algorithm (CERNO), relying on the
-ordering of genes according to their raw p-values from the interaction
-contrast rather than selecting a
-set of DEGs [@zyla2019gene], does not show any significant enrichment.
+]
 
-### Influence of sample size and cut-off thresholds on number of artifacts
+.pull-right[
 
-In the example above, the groups have been randomly sampled from a larger
-data set only once. Arguably, the observations might differ if the groups
-were to be resampled. To investigate whether the sampling had an impact on
-the artifacts, we have repeated the above procedure for 100 replicates,
-each containing a different set of samples randomly assigned to the two
-groups.  Furthermore, we tested whether the selection of the log~2~ fold
-change threshold or different sample size might have an impact on the
-extent of the arising artifacts. To this end, we have tested `r
-length(lfc_thresholds)` different threshold values for the log~2~ fold
-changes and three sample sizes: 20, 40 or 80 samples per group
-(corresponding to 10, 20 or 40 samples per group/treatment combination).
+Genes significant in one, but not the other comparison behave very
+similarly
 
-**Fig. 3. Influence of log~2~ fold change treshold (A, C) and the sample
-size (B, D) on the number of genes significant in one, but not in the other
-group (A, B) and the number of significant gene sets found in one group, but
-not the other (C, D).**  Top row shows the influence of log~2~ fold change
-threshold for sample size of 40 per group (20 per group/treatment
-combination), bottom row shows the influence of sample size for a log~2~
-fold change threshold set to `r lfc_thr`.
+]
+
+---
+
+## The reality
+
+.pull-left[
+
+```{r}
+part2
+```
+
+]
+
+.pull-right[
+
+Genes significant in one, but not the other comparison behave very
+similarly
+
+]
+
+---
+
+```{r lit_surveys}
+get_keywords <- function(file, pattern="^([a-zA-Z]+): *(.*)") {
+  ll <- readLines(file)
+  ll <- ll[ grepl(pattern, ll) & !grepl("^https://", ll) ]
+
+  ret <- gsub(pattern, "\\2", ll)
+  names(ret) <- gsub(pattern, "\\1", ll)
+  ret <- as.list(ret)
+  ret$file <- file
+  ret$incorrect_count <- get_incorrect_counts(file)
+  ret$incorrect_count_vdonly <- get_incorrect_counts(file, vdonly=TRUE)
+  ret
+}
+
+get_incorrect_counts <- function(file, vdonly=FALSE) {
+
+  ll <- readLines(file)
+
+  pat <- "^https://"
+  if(vdonly) {
+    sum(grepl(pat, ll) & grepl("vd only", ll))
+  } else {
+    sum(grepl(pat, ll))
+  }
+
+}
+
+files <- list.files(pattern="literature_survey.*.md")
+
+finfo <- map(files, get_keywords) 
+names(finfo) <- map_chr(finfo, ~ .x[["journal"]])
+```
+
+**Tab. 2** Results of the informal literature survey.
+
+.pull-left[
+We searched for papers using Google Scholar and the keywords "venn diagram"
+and "differential expression".
+]
+
+.pull-right[
+
+```{r results="markdown"}
+tab <- map_dfr(finfo, ~ 
+               data.frame(Journal=.x$journal,
+                          Years=.x$years,
+                          Total=.x$total,
+                          Analysed=.x$analysed,
+                          Incorrect=.x$incorrect_count,
+                          Enrichment=.x$incorrect_count - .x$incorrect_count_vdonly))  %>%
+  mutate(Analysed=as.numeric(Analysed)) %>%
+  mutate(Total=as.numeric(Total))
+
+tab <- rbind(tab, 
+             data.frame(Journal="**Total:**", Years="", Total=sum(tab$Total),
+                        Analysed=sum(tab$Analysed),
+                        Incorrect=sum(tab$Incorrect),
+                        Enrichment=sum(tab$Enrichment)))
+tab2 <- tab %>%
+  mutate(Incorrect=sprintf("%d (%.0f%%)", Incorrect, 100 * Incorrect / Analysed)) %>%
+  mutate(Enrichment=sprintf("%d (%.0f%%)", Enrichment, 100 * Enrichment / Analysed))
+
+kable(tab2)
+```
+
+]
+
+---
+
+## It gets weirder
+
+Often, researchers ask the question: what are these "specific" genes?
+
+.pull-left[
+
+```{r fig1_1b, fig.width=5,fig.height=5,dpi=300,dev="png"}
+p1 <- ggvenn(list(G1=res$g1 %>% filter(DEG) %>% pull(ENSEMBL),
+                         G2=res$g2 %>% filter(DEG) %>% pull(ENSEMBL)),
+             show_percentage=FALSE)
+p1
+
+```
+
+]
+
+--
+
+.pull-right[
+
+Classical approach involves testing for gene set enrichment for the group
+of "specific" genes.
+
+]
+
+---
+
+## Here are the results.
+
+```{r fig1, fig.width=15,fig.height=6.5,dpi=300,dev="png"}
+p1 <- ggvenn(list(G1=res$g1 %>% filter(DEG) %>% pull(ENSEMBL),
+                         G2=res$g2 %>% filter(DEG) %>% pull(ENSEMBL)),
+             show_percentage=FALSE) +
+      ggtitle("Differentially expressed genes")
+p2 <- ggvenn(list(G1=gsea_res$g1 %>% pull(ID),
+                         G2=gsea_res$g2 %>% pull(ID)),
+              show_percentage=FALSE) +
+      ggtitle("Enriched GO terms")
+
+col1 <- plot_grid(p1, p2, labels="AUTO", nrow=2)
+
+p3 <- plot_gsea(gsea_res$g1 %>% arrange(P.Value) %>% dplyr::slice(1:10)) + ggtitle("G1") +
+  scale_x_discrete(labels=function(x) str_wrap(x, width=25))
+#p3 <- plot_grid(p3, NULL, ncol=1, rel_heights=c(2, 1))
+p4 <- plot_gsea(gsea_res$g2 %>% arrange(P.Value) %>% dplyr::slice(1:10)) + ggtitle("G2") +
+  scale_x_discrete(labels=function(x) str_wrap(x, width=25))
+
+plot_grid(plotlist = list(col1, p3, p4), labels=c('', 'C', 'D'), nrow=1)
+```
+
+---
+
+## Wait, what?
+
+.pull-left[
+
+These enrichments don't look random!
+
+]
+
+.pull-right[
+
+```{r fig1_c, fig.width=10,fig.height=5.5,dpi=300,dev="png"}
+plot_grid(p3, p4, nrow=1)
+```
+
+]
+
+---
+
+## Here is what happens
+
+---
+
+## "Specific" genes are likely FNs
+
+```{r fig3,dpi=300,dev="png"}
+tmp <- merge(res.merged, res$full, by="ENSEMBL")
+tmp <- tmp[ , c("DEG.g1", "DEG.g2", "DEG") ]
+colnames(tmp) <- c("G1", "G2", "Full dataset")
+plot(euler(tmp), quantities=list(type=c("counts", "percent")))
+```
+
+---
+
+## Can we get rid of it in another way?
 
@@
 
-Setting a higher log~2~ fold change threshold reduces the number of
-DEGs and of observed artifacts (Fig. 3, A and B). However, even for log~2~
-threshold of 3 (DEGs defined by 8-fold change and FDR < `r pval_thr`) the
-number of replicates in which artifacts can be observed is 
-`r sum(gsea_by_lfc_thr %>% filter(lfc == 3) %>% pull(value) > 0)`
-(out of 100),
-and in at least 
-`r sum(gsea_by_lfc_thr %>% filter(lfc == 3) %>% pull(value) > 0)`
-replicates, 5 or more gene sets were significantly
-enriched. Thus, setting a more conservative threshold cannot protect from
-the arising artifacts.
-
-Contrary to our expectations, the number of artifacts rose with sample size
-(Fig. 3, C and D). For total sample size of 160 (40 samples per
-group/treatment combination) the number of artifacts was almost 
-`r round(
-  mean(gsea_by_ss %>% filter(ss == 40) %>% pull(value)) /
-  mean(gsea_by_ss %>% filter(ss == 20) %>% pull(value)))
-`
-times higher than for total sample size of 80.
-
-### Artifacts arise because of false negatives
-
-It is worth noting that in the gene set enrichment analysis of the genes
-"specific" for a given comparison – i.e., genes which are significant in
-that comparison, but not significant in others – we have observed a number
-of terms associated with immune response. It is a crucial point of this
-manuscript to note that the spurious enrichments not only show significant
-p-values, but also that the terms or pathways which appear in them are
-relevant to the research hypothesis being tested. Below, we will show why
-these terms (rather than random terms which have no obvious relevance for
-an infectious disease) appear in the results.
-
-To understand how gene set enrichment actually gives significant
-results in such randomly generated gene sets, despite absence of genes with
-significant interaction, it is first necessary to consider the definition
-of a differentially expressed gene in this context. More often than not,
-DEGs are defined by a threshold in p-value adjusted for multiple testing,
-possibly combined with a threshold in log~2~ fold change. The commonly used
-Benjamini-Hochberg procedure [@benjamini1995controlling] ensures that among
-genes for which the adjusted p~adj~ < 0.05 there are at most 5% false
-positives.
-
-This way, we can exert control over the false positive rate (FPR, type I
-errors), keeping it at a relatively low level. However, we do not control
-the false negative rate (FNR, type II errors).  In a powerful statistical
-test (such as a t-test), the test power in a typical application will
-rarely achieve more than 80%. For example, even for large
-effects (Cohen's d > 0.8) and type I error rate of 0.05, a t-test only achieves
-80% power with at least 
-`r floor(pwr.t.test(power=.8, d=0.8)$n)`
-samples per group. For small effects (Cohen's d > 0.2), the required number
-of samples is at least
-`r floor(pwr.t.test(power=.8, d=0.2)$n)`
-per group.
-Even assuming a test power of 80%, the FNR is 20%. Clearly, false negatives (FNs) occur
-at much higher rates than false positives (FPs). In case of high throughput
-data sets, where the FPR is controlled by Bejnamini-Hochberg procedure or a
-similar technique, the FNR may be even as high as 80% [@white2019beyond].
-
-```{r full_dataset}
-ds2_full <- DESeqDataSetFromMatrix(counts, colData=covar, design=~ disease )
-ds2_full<- DESeq(ds2_full)
-res$full <- results(ds2_full, name="disease_SC2_vs_no") %>% as.data.frame() %>%
-  rownames_to_column("ENSEMBL") %>% 
-  left_join(annot, by="ENSEMBL") %>%
-  mutate(DEG=!is.na(padj) & abs(log2FoldChange) > lfc_thr & padj < padj_thr)
-```
-
-```{r g_specific}
-g1_spec <- res.merged %>% filter(DEG.g1 & !DEG.g2) %>% pull(ENSEMBL)
-g2_spec <- res.merged %>% filter(!DEG.g1 & DEG.g2) %>% pull(ENSEMBL)
-```
-
-These FNs will occur
-at a much higher rate within the sets of DEGs defined by the
-non-overlaping areas of the VDs, that is DEGs considered to be "specific"
-for one group or other in a naive approach.
-To illustrate this phenomenon, we have analysed the full data set from which G1 and G2
-were drawn (Fig. 4), comparing the 
-`r sum(covar$disease == "no") `
-healthy controls to 
-`r sum(covar$disease == "SC2") `
-COVID-19 patients. Of the
-`r length(g1_spec)`
-genes significant in G1, but not in G2, 
-`r .n <- res$full %>% filter(DEG & ENSEMBL %in% g1_spec) %>% nrow; .n`
-(`r sprintf("%.0f%%", .n/length(g1_spec) * 100)`) 
-are significant in the full data set; 
-of the
-`r length(g2_spec)`
-genes significant in G2, but not in G1, 
-`r .n <- res$full %>% filter(DEG & ENSEMBL %in% g2_spec) %>% nrow; .n`
-(`r sprintf("%.0f%%", .n/length(g2_spec) * 100)`) 
-are significant in the full data set.
-Given that G1 and G2 were sampled from the total population, and since the
-FDR was set to `r padj_thr`, we can assume that at least between a third and a half of
-the genes that appeared to be "specific" in the initial analysis were, in
-fact, false negatives in one of the comparisons.
-
-In other words, a substantial fraction of the "specific" genes are genes
-that are in reality differentially expressed in both groups alike.
-Therefore, if one is to perform a gene set enrichment analysis on one of
-these "specific" groups of genes, then the enriched functions will be
-related to the pathways and processes up- or down-regulated in both groups
-due to the common factor (in this example, the COVID-19 disease), but
-which are not related to differences between the two groups.
-
-**Fig. 4. Area-proportional Venn diagram showing overlaps in DEGs between
-G1, G2 and the full data set.** The majority of genes which have been labeled
-as DEGs in only one of the groups G1, G2 are DEGs when all data were
-analysed.
-
-```{r fig3,dpi=300,dev="png"}
-tmp <- merge(res.merged, res$full, by="ENSEMBL")
-tmp <- tmp[ , c("DEG.g1", "DEG.g2", "DEG") ]
-colnames(tmp) <- c("G1", "G2", "Full dataset")
-plot(euler(tmp), quantities=list(type=c("counts", "percent")))
-```
-
-## Discussion
-
-```{r lit_surveys}
-get_keywords <- function(file, pattern="^([a-zA-Z]+): *(.*)") {
-  ll <- readLines(file)
-  ll <- ll[ grepl(pattern, ll) & !grepl("^https://", ll) ]
-
-  ret <- gsub(pattern, "\\2", ll)
-  names(ret) <- gsub(pattern, "\\1", ll)
-  ret <- as.list(ret)
-  ret$file <- file
-  ret$incorrect_count <- get_incorrect_counts(file)
-  ret$incorrect_count_vdonly <- get_incorrect_counts(file, vdonly=TRUE)
-  ret
-}
-
-get_incorrect_counts <- function(file, vdonly=FALSE) {
-
-  ll <- readLines(file)
-
-  pat <- "^https://"
-  if(vdonly) {
-    sum(grepl(pat, ll) & grepl("vd only", ll))
-  } else {
-    sum(grepl(pat, ll))
-  }
-
-}
-
-files <- list.files(pattern="literature_survey.*.md")
-
-finfo <- map(files, get_keywords) 
-names(finfo) <- map_chr(finfo, ~ .x[["journal"]])
-```
-
-**Tab. 2** Results of the informal literature survey. 
-We searched for papers using Google Scholar and the keywords "venn diagram"
-and "differential expression".
-*Journal*, journal title; *Years*, publication dates; *Total*, total number
-of papers found using the search phrase; *Analysed*, total number of papers
-which have been analysed for correctness; *Incorrect*, total number
-(percentage) of
-papers in which the Venn diagram was combined with comparing significance
-to non-significance; *Enrichment*, total numebr (percentage) of papers
-which combined the Venn diagram with a gene set enrichment analysis.
-
-```{r results="markdown"}
-tab <- map_dfr(finfo, ~ 
-               data.frame(Journal=.x$journal,
-                          Years=.x$years,
-                          Total=.x$total,
-                          Analysed=.x$analysed,
-                          Incorrect=.x$incorrect_count,
-                          Enrichment=.x$incorrect_count - .x$incorrect_count_vdonly))  %>%
-  mutate(Analysed=as.numeric(Analysed)) %>%
-  mutate(Total=as.numeric(Total))
-
-tab <- rbind(tab, 
-             data.frame(Journal="**Total:**", Years="", Total=sum(tab$Total),
-                        Analysed=sum(tab$Analysed),
-                        Incorrect=sum(tab$Incorrect),
-                        Enrichment=sum(tab$Enrichment)))
-tab2 <- tab %>%
-  mutate(Incorrect=sprintf("%d (%.0f%%)", Incorrect, 100 * Incorrect / Analysed)) %>%
-  mutate(Enrichment=sprintf("%d (%.0f%%)", Enrichment, 100 * Enrichment / Analysed))
-
-pander(tab2)
-```
-
-In an informal survey of articles from three journals (Tab. 2) we found
-that of the 
-`r tab2$Analysed[nrow(tab2)]` analysed articles
- which used the term "venn diagram" and
-"differential expression", at least 
-`r tab2$Incorrect[nrow(tab2)]`
-were using Venn diagrams to compare
-statistical significance with lack thereof by referring to "unique",
-"specific", "solely regulated" or "exclusive" DEGs. Out of these, at least 
-`r tab$Enrichment[nrow(tab2)]`
-coupled the VDs with some form of gene set enrichment analysis on the
-set of supposedly "specific" DEGs.
-
-Drawing conclusions from comparing significance with lack thereof is a
-common statistical fallacy [@gelman2006difference]. Just as absence of evidence is not evidence of
-absence, the failure to reject the null hypothesis does not consitute the
-same level of evidence as rejecting it. However, when such an incorrect
-analysis is combined with downstream functional analysis – i.e., gene set
-enrichments of genes "specific" to one or another comparison – the
-resulting pathways or gene ontologies are misleadingly relevant, yielding
-gene sets associated with immune answer for research hypotheses involving
-an infectious disease, or cancer pathways if the underlying research
-hypothesis involved cancer treatment. Such results may be hard to resist
-for an involved researcher, especially if the correct analysis of
-interactions does not show any significant differences.
-
-That is not to say that VDs are not a useful tool, even in the context of
-transcriptomics and gene set enrichments, if used correctly. For example, analysis of an
-intersection of DEGs (i.e., by considering genes from the overlap in a VD)
-is not an incorrect procedure.  Genes in the overlapping part of a VD are
-significant in both (or all) comparisons, hence no comparison between
-significance and non-significance is made.
-
-As an alternative to Venn diagrams and the downstream gene set enrichment
-analysis, two approaches can be considered.
-
-Statistical analysis, as shown above, involves a test for interaction which
-can reveal genes for which the impact of treatment significantly differ
-between the groups. The results can then be plugged into a gene set
-enrichment analysis the usual way.  Unfortunately, this has two major
-drawbacks. Firstly, effect sizes (log~2~ fold changes) of the interaction
-term are harder to interpret than log~2~ fold changes in a direct, group vs
-group comparison.  The effect size in an interaction is negative if the
-log~2~ fold change in the first comparison is larger than the log~2~ fold
-change in the second comparison; this is, however, irrespective of whether
-the differences in the individual comparisons are negative or positive,
-which makes it harder to separate the differences in genes up-regulated in
-one or both groups.
-
-The second problem may arise if the changes are similar in both
-comparisons, but of larger magnitude in one of them. For example, in a time
-series context, the changes may be more pronounced at a later time point.
-In this case, the analysis will show that the processes enriched for the
-interaction term are the same as those enriched in each of the comparisons
-individually.  While the results of the gene set enrichment analysis in
-this context are correct, the result may not be what the researchers
-intended – processes which qualitatively (rather than quantitatively)
-differ between the comparisons.
-
-An alternative approach, discordance/concordance analysis, has been proposed by @domaszewska2017concordant,
-aiming at identifying processes which qualitatively differ between the two
-comparisons. Here, a heuristic score ("disco score") has been defined which depends on the
-effect sizes and p-values in both comparisons. The sign of the score
-depends on whether the effects have the same sign (concordant; genes
-up-regulated in both comparisons or down-regulated in both comparisons) or
-opposite signs (discordant; genes up-regulated in one, but down-regulated
-in the other comparison, and genes down-regulated in one, but up-regulated
-in the other comparison). While the score does not allow the calculation of
-a p-value and does not present an alternative to an analysis of
-interaction, it can facilitate both visualization and further analysis
-using a gene set enrichment algorithm.
-
-Visualization of interaction for individual genes is straightforward (see
-for example Fig. 2, *A-D*). However, the point of VDs is to show a grand
-overview of the whole analysis summarising thousands of results for the
-analysed genes. As an alternative of such an overview, we suggest plotting
-the log~2 fold changes in one comparison against log~2~ fold changes in the
-second comparison. This allows an intuitive assessment of the differences
-between the two comparisons, in especially with combination of color coding
-the genes either are significant in the interaction or using the disco
-score (see Fig. 2, *E* for an example).
-
-While the use of VDs in combination with the incorrect gene set
-enrichment analysis is common, it is not the only way of achieving
-artifactual and incorrect conclusions by means of comparing significant
-with non-significant results in transcriptomics and other high-throughput
-applications. VDs appear to be symptomatic of the assumption that if a gene
-is significantly regulated in one comparison, but does not show a
-significant difference in another one, then this gene is "specific" to this
-first comparison. However, VDs may also lead one to this reasoning; for
-example, the VD on Fig. 1A practically begs the question: what do these 
-`r sum(res.merged$DEG.g1 & !res.merged$DEG.g2)` 
-genes, which are significant in G1 only, have in common?
-
-Our literature survey showed that even in cases where a VD is absent from a
-paper, at least two other ways are not uncommon. Firstly, the direct
-comparison of gene set enrichment results: that is, drawing conclusions
-from the fact that a gene set enrichment result was significant in one
-comparison only.  Second, while Venn diagrams are often used to illustrate
-the numbers of "specific" DEGs and so present a mean to find examples of
-this fallacy in scientific literature, researches often test for enrichment
-these "specific" genes without using the phrase "Venn diagram" or even
-clearly stating how the lists of "specific" genes were derived. In all
-these cases, the analysis boils down to comparing results significant in
-one, but not in another comparison.
-
-The use of Venn diagrams to illustrate specific differences between
-comparisons should therefore be abandoned in favor of statistically correct
-approaches. Furthermore, gene set enrichment analysis must never be applied
-to sets of genes defined as significant in one comparison, but not the
-other.
-
-## Methods
-
-### Methods availability
-
-This document has been written as an Rmarkdown file, containing all
-statistical calculations required to replicate the findings and figures.
-The markdown file, along with additional files required to recreate this
-manuscript have been uploaded to [https://github.com/bihealth/manuscript_venn_diagrams](https://github.com/bihealth/manuscript_venn_diagrams).
-
-### Data
-
-The expression data as a count matrix has been downloaded from GEO,
-accession GSE156063.
-
-### Statistical analyses
-
-Power calculation was done using the R package pwr, version
-`r packageVersion("pwr")`.
-For differential gene expression, the R package DESeq2, version
-`r packageVersion("DESeq2")`
-has been used. Gene set enrichments were done using either hypergeometric
-test (where stated) or the CERNO test using the package tmod [@zyla2019gene],
-version
-`r packageVersion("tmod")`.
-GO terms have been sourced from the R package msigdbr, version
-`r packageVersion("msigdbr")`.
-
-### Literature survey
-
-An informal literature survey was performed using Google Scholar to
-estimate the frequency of the incorrect use of Venn diagrams. We searched
-for articles from 2020 mentioning the phrases "differential expression" and "venn
-diagram" in the Scientific Reports journal. The journal was
-selected because it represents a wide spectrum of research areas, the
-articles are available through an open access license and there is a large
-number of publications per year. For each of the papers identified, we
-checked whether (i) the authors used the VD to show differentially expressed
-transcripts significant in one comparison, but not another, (ii) the
-authors discussed "unique", "non-overlapping" or "specific" regions of the
-Venn diagram and (iii) whether this was coupled to gene set enrichment
-analysis is any form. Articles which (i) focused only on the intersections
-of the Venn diagrams (genes common to all conditions), or (ii) which used the
-Venn diagrams for a purpose other than to compare genes significant in one
-condition, but not significant in another condition or (iii) for which a
-clear-cut error could not be indicated past any reasonable doubt were not
-considered incorrect.
-
-## Bibliography
+---

### 2022-01-06 presentation/functions.R

@@ -1,0 +1,109 @@
+plot_gsea <- function(df, es_col="E", pval_col="adj.P.Val") {
+
+  df <- df %>% arrange(-.data[[pval_col]])
+  df$log10_pval <- -log10(df[[pval_col]])
+
+	df$Title <- .cleanup_ids(df$Title)
+	df$Title <- factor(df$Title, levels=df$Title)
+  
+  ggplot(df, 
+         aes_string(x="Title", y="log10_pval", fill=es_col)) + geom_bar(stat="identity") +
+         xlab("") +
+         ylab("-log10(FDR)") +
+         coord_flip()
+
+
+}
+
+.cleanup_ids <- function(ids) {
+  ids <- gsub("_", " ", ids)
+
+  #return(ids)
+
+  ids <- strsplit(ids, " ")
+  min.l <- min(sapply(ids, length))
+
+  max_prefix <- 0
+  for(i in 1:min.l) {
+
+    if(length(unique(sapply(ids, function(x) x[i]))) == 1L) {
+      max_prefix <- i
+    }
+  }
+
+  if(max_prefix > 0) {
+    ids <- lapply(ids, function(x) x[ -1:-max_prefix ])
+  }
+
+  ids <- unlist(lapply(ids, paste, collapse=" "))
+
+  if(!any(grepl("[a-z]", ids))) {
+    ids <- tolower(ids)
+    substr(ids, 1, 1) <- toupper(substr(ids, 1, 1))
+  }
+
+  ids
+}
+
+
+ggplot_gene <- function(x, group, 
+                        groups=list(1:2, 3:4),
+                        annotation=NULL, ...) {
+
+  df <- data.frame(Expression=x, Group=group)
+
+  p1 <- ggplot(df, aes(x=Group, y=Expression)) + 
+      geom_jitter(width = .2, size=1, alpha=.5) +
+      geom_boxplot(fill=NA, outlier.shape = NA) 
+      
+  if(!is.null(annotation)) {
+    p1 <- add_pval(p1, groups, annotation = annotation, ...)
+  }
+
+  p1
+
+}
+
+ggplot_disco <- function(df,
+                         lfc1="log2FoldChange.g1",
+                         lfc2="log2FoldChange.g2",
+                         pval1="padj.g1",
+                         pval2="padj.g2") {
+
+  df$disco <- 1
+
+
+
+
+
+}
+
+
[…]

### 2022-01-16 mk4/response_to_editor.md
major rewrite

@@
 
-**We do agree in principle. However, the term "venn diagram analysis" (and
-related formulations) is firmly (even though rather unfortunately)
-established in the field of differential expression and many authors are
-using it essentially as a shorthand for stating that the supposedly
-treatment specific genes are found by comparing which genes are significant
-in one, but not the other comparison. We have now explained this in more
-detail in the Introduction as follows:**
-
-XXX
+**We do agree ; we have stated that ourselves in the
+Discussion. We have now reformulated the Abstract and Introduction.**
 
@@
 **We fully agree; we have already addressed this issue in the
-"Discussion".**
+"Discussion". However...**
 
@@
 **We understand that point. However, we do observe a substantial fraction of
-papers using Venn diagrams misinterpreting the results. The given analogy
-would only work if every fourth use of a knife resulted in a homicide. This
-point has already been addressed in the Discussion. Nonetheless we have now
-removed references to Venn diagrams in the title and in the abstract.**
+papers using Venn diagrams misinterpreting the results. To continue on the
+metaphor, the given analogy would only work if every fourth use of a knife
+resulted in a homicide. This point has already been addressed in the
+Discussion. Nonetheless we have now removed references to Venn diagrams in
+the title and in the abstract.**
 
@@
 diagrams. The file `literature_survey_scirep.md` contains 73 links to
-papers from Scientific Reports (out of 238 reviewed).**
+papers from Scientific Reports (out of 238 reviewed). We now mention that
+explicitely in "Methods".**
 
@@
 not an unusually small sample size for this type of transcriptomic
-analysis (see below for more on the chosen sample size). Also, the design of our in
+analysis (see below for more on the chosen sample size). In fact, in your
+2019 paper on the influence of sample size on gene set enrichments that was
+the largest group size you have used.
+Also, the design of our in
 silico experiment *requires* splitting the data set into two randomly selected groups,
@@
 
-**We fully agree that FNs are the core of the problem! However, we know
-from other studies (and basic statistics) that even in most favorable
-circumstances the FNR will likely be above 20%; and that in case of transcriptomic
-studies, the FNR may be much larger (please see the relevant fragment already
-included in the paper). In other words, it does not matter whether one
-looks at 80 samples or 193 samples: FNs are bound to occur. That is the
-reason – we claim – for the fact that the gene set enrichments found in
-this manner are related to the studied problem. However, we have addressed
-this isssue by testing different sample sizes (see below).**
+**We fully agree that FNs are the core of the problem! In fact, we have
+chosen the FDR of 0.05 to minimize the number of false negatives (since a
+more conservative threshold results in higher false negative rate).**
+
+**However, we know from other studies (and basic statistics) that even in
+most favorable circumstances the FNR will likely be above 20%; and that in
+case of transcriptomic studies, the FNR may be much larger (please see the
+relevant fragment already included in the paper). In other words, it does
+not matter whether one looks at 80 samples or 193 samples: FNs are bound to
+occur. That is the reason – we claim – for the fact that the gene set
+enrichments found in this manner are related to the studied problem.
+However, we have addressed this isssue by testing different sample sizes
+(see below).**
 
@@
 see that half of the studies used sample sizes sizes of 6 or less per
-group.
+group.**
 
-Nonetheless, we think that showing how the artifacts depend on sample size
+**Nonetheless, we think that showing how the artifacts depend on sample size
 is, in itself, interesting, not least because it shows that sample size has
@@
 The reason for this is, of course, that the more DEGs you observe, the more
-powerful the hypergeometric test is.
+powerful the hypergeometric test is.**
 
-For each sample size, we have run 100 replicates of the group sampling,
+**For each sample size, we have run 100 replicates of the group sampling,
 replicated the whole procedure and gathered all test results. Results are
-shown on the new Fig. 3, C and D.
+shown on the new Fig. 4, C and D.**
 
-However, we also think that the core of the problem is the incorrect
+**However, we also think that the core of the problem is the incorrect
 statistical procedure (selecting genes by comparing significance with lack
@@
 whereas setting the FDR threshold gives you clear information about the
-expected proportion of false positives in your results.
+expected proportion of false positives in your results.**
 
-This is analogous to using an effect size filter (say, Cohen's d) on
+**This is analogous to using an effect size filter (say, Cohen's d) on
 results of a t-test. The p-value from the t-test controls the type I error
@@
 size—or why the P value is not enough." Journal of graduate medical
-education 4.3 (2012): 279-282.).
+education 4.3 (2012): 279-282.).**
 
-Thus, the threshold serves to eliminate genes that do show statistically
-significant change (even if they are true positives), but which are not
+**Thus, the threshold serves to eliminate genes that do show statistically
+significant change (that is, they are true positives), but which are not
 biologically relevant (because the effect size, for which log2 fold change
 is a proxy, is too small). The proportion of false positives is 5% even with no
-log2 FC treshold.
+log2 FC treshold.**
 
-Increasing the threshold can, of course, result in not observing any
-enrichments (see Fig. 3, A and B). Indeed, it does for some replicates.
+**Increasing the threshold can, of course, result in not observing any
+enrichments (see Fig. 4, A and B). Indeed, it does for some replicates.
 This is not surprising – without a sufficient number of DEGs, no
@@
 symptoms, it does not cure the disease – the core problem of erroneous
-analysis of interactions remains.
+analysis of interactions remains, and many replicates have artifactual
+enrichments even at high log fold change thresholds.**
 
-Nonetheless, we have compared the effect of the log2
-fold change threshold on the results (Fig. 3).**
+**Nonetheless, we have compared the effect of the log2
+fold change threshold on the results (Fig. 4).**
 
@@
 For this, we have replicated the whole procedure of group randomization 100
-times. Summary overview can be found in the new Fig. 3. Bottom line: yes,
+times. Summary overview can be found in the new Fig. 4. Bottom line: yes,
 for many replicates the artifacts disappear when the log2 fold change threshold is high –

### 2022-01-16 mk4/manuscript.rmd
major rewrite

@@
 ---
-title: "Venn diagrams may indicate erroneous statistical approaches leading to artifacts in transcriptomic analyses"
+title: ""
 author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
 date: "`r Sys.Date()`"
+outputoff:
+  html_document
 output:
-  html_document
-outputoff:
   word_document:
@@
 
+## Titles
+
+???
+
 ^1^ Berlin Institute of Health at Charité – Universitätsmedizin Berlin, Core Unit Bioinformatics, Charitéplatz 1, 10117 Berlin, Germany
@@
 
-Venn diagrams are frequently used to illustrate the results of
-differential expression analysis and further
-downstream functional analysis. Here we show that both, the use
-of Venn diagrams to find genes which are thought to be specific for a
-certain comparison, as well as gene set enrichment analysis applied to such
-subsets is a fallacy.  A statistically correct approach involves testing for interaction.
-Moreover, since genes which show a significant change in one condition but
-not in another are likely to be false negatives in the latter case, these
-genes considered to be specific for one condition are likely characteristic
-for both of the compared groups, and enriched in functions related to the
-research hypothesis. Thus, the combination of incorrect statistical
-analysis and gene set enrichment analysis may result in particularly
-misleading artifacts.
+Version 1: In statistics, comparing the effect of a treatment in two different
+conditions requires the use of an appropriate statistical test to assert
+the significance of interaction between the treatment and the condition.
+However, a common procedure is to define genes specifically regulated in
+one, but not the other condition by selecting genes which show a
+statistically significant difference in one condition only. Here we show
+that not only is this procedure incorrect, but moreover, when combined with
+gene set enrichment analyses, it leads to statistically significant, but
+incorrect results. These gene set enrichments are the effect of false
+negatives and reflect only the differences in gene expression due to
+treatment and not due to the difference between conditions. Thus, the
+combination of incorrect statistical analysis and gene set enrichment
+analysis may result in particularly misleading artifacts.
+
+Version 2: A common application of differential expression analysis is to
+ask which genes are differentially expressed upon treatment in one
+condition only. One of the approaches is to define genes specific to a
+condition by testing for significant difference in expression separately
+in the two conditions, and then selecting genes which show statistical
+significance in one condition only. This approach may then be combined with
+a gene set enrichment analysis to find pathways and gene sets
+characteristic for this condition only.
+
+Here we show that this procedure is incorrect. Instead, the interaction
+between treatment and condition should be tested. Moreover, we show that
+gene set enrichment analysis applied to incorrectly defined genes specific
+to one condition only may result in misleading artifacts. Genes which show
+significant difference upon treatment in one, but not the other condition
+are enriched in gene sets related to the treatment, but not to the
+condition. Thus, while the results are apparently related to the treatment,
+they do not reflect differences in expression specific to a given
+condition.
+
+Version 3:
 
@@
 lfc_thr  <- 1
-n_stratum <- 20 # 2 strata per condition 
+n_stratum <- 10 # 2 strata (male and female) per condition/group
+                # total number of samples n_stratum * 2 * 2 * 2
+lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
 ```
 
-## Introduction: Venn diagrams are a showcase of erroneous interpretation of interactions
+## Introduction
 
-Venn diagrams (VDs) are commonly used to visualize high throughput data
-such as transcriptomic profiles. Frequently, a Venn diagram serves the
-purpose of comparing the results from two experiments or two comparisons.
-For example, a VD may illustrate up- or down-regulated genes in two strains
-of mice upon infection, showing which genes are regulated in both strains,
-and which are regulated only in one of them.  The VD serves as a basis for
-the statement that the regulation of certain genes is "specific" for one
-strain or another. A gene is, in this context, considered "specific" or
-"unique" for a given comparison when it shows a statistically significant
-difference in that comparison only.  We argue that not only is this
-inference incorrect, but it may also lead to misleading – although
-appealing – artifacts when combined with downstream analyses.
+Transcriptomic analyses commonly include more than one factor. Often, the
+question asked is whether there is a difference between conditions (first
+factor) with respect to reaction to a particular treatment (second
+condition). For example, we may ask whether there are differentially
+expressed genes (DEGs) which are specific to a particular group of
+patients, such as interferon response elicited by a virus in one group, but
+silenced in the other group of patients. In other words, we ask whether the
+difference between control (healthy subjects) and treatment (infected
+patients) is different between two groups of individuals. This "difference
+of differences" is known in statistics as an interaction
+[@blalock1965theory]. To find out whether it is statistically significant,
+an appriopriate statistical test for interaction should be employed.
 
-Consider a gene for which the expression has been analysed in four groups:
-two different mouse strains (wildtype, WT and knockout, KO) and two different experimental conditions
-(naive versus infected). We find that the gene expression significantly
-differs between infected and naive KO animals, but that there is no
-significant difference in the WT strain. Such a gene may be incorrectly
-considered as "specific" to KO, and will be accordingly entered in a Venn
-diagram.
+However, another approach is widely spread [@nieuwenhuis2011erroneous].
+Instead of testing the interaction, one tests the effect of the treatment
+separately in both groups. Next, a difference between conditions is inferred if the difference
+is significant in one comparison, but not significant in the other. This
+approach is not correct from statistical point of view, as "the difference
+between significant and not significant is not itself statistically
+significant" [@gelman2006difference]. For example, the p-value in the first
+comparison may be 0.009, and in the other comparison 0.011, thus at an
+alpha level of 0.01 the difference will be statistically significant in the
+first, but not significant in the other comparison.
 
-However, this is a common fallacy [@nieuwenhuis2011erroneous], since lack
-of statistical significance is not the same as testing for lack of
-difference. In other words, the fact that we failed to detect a significant
-difference in the WT does not mean that the difference is absent and
-itself significantly different from the difference in the KO. This "difference of
-differences" is known is statistics as interaction [@blalock1965theory].
-Here, it is the interaction between strain and treatment.  In fact, the
-obtained p-values might be just over the assumed p-value threshold in one, but just
-under it in the other strain (e.g. 0.009 vs 0.011), with the difference in
-both strains being essentially the same. The correct analysis is to test
-the significance of interaction using an ANOVA model. Notably, in 2011
-Nieuwenhuis et al. found that in over 150 papers in top neuroscience
-journals where the authors could make this particular statistical error,
-half of them did.
+In transcriptomics, the statistical tests are performed for thousands of
+genes. As in the general case, the correct inference of differences between
+the conditions is by testing the significance of interaction. However, in
+another approach the differences between treatment and control are tested
+in the two conditions separately, and the results are then compared. This
+can be visualized using a Venn diagram (VD, Fig. 1) showing the overall
+number of DEGs significant in both comparisons (the intersection in the VD)
+or significant in only one comparison (the remaining two fields on a VD).
+The genes which are significant in only one comparison may then be
+incorrectly defined as specific to the corresponding condition. Following
+this, gene set enrichment analysis may be applied to these genes in an
+attempt to test which pathways are specific to one, but not the other
+condition.
 
-Using VDs to show genes "specific" for a condition amounts to
-counting the times a comparison for a gene was statistically significant in
-one condition, but not significant in another.
-More disturbingly, VDs are a visual illustration of a procedure
-that exacerbates this problem by applying a downstream analysis to
-supposedly specific sets of genes. Several cells in a Venn diagram contain
-numbers corresponding to features (such as genes), for which there was a significant test result in
-one, but not in another comparison. The sets of genes in a Venn diagram cell may
-then be subsequently analysed to test whether they share a particular
-characteristics.  For example, gene set enrichment analysis may be used to
-interpret the biological function of genes that are considered "specific" to one
-condition. It turns out that due to the
-fallacious nature of this procedure, it is likely to produce results that
-appear to make sense in the given biological context.
-
-In this paper, we illustrate this rather simple statistical statement with
-a real world data set, demonstrating how choosing this sort of approach
-results in apparently sound gene set enrichment results which are, in fact,
-artifacts. Next, we dissect the underlying mechanism of how these artifacts
-are generated.  Finally, we discuss alternative approaches.
+In this paper, we show that under reasonable assumptions this approach may
+result in apparent enrichments even if there are no real differences
+between the groups. To this end, we have randomly split a cohort into two
+groups, compared the treatment (viral infection) with controls in each of
+the groups separately and then applied gene set enrichment to the sets of
+genes significantly different in one, but not the other group.
+Moreover, we show that the resulting gene set enrichments correspond to the
+differential expression between treatment and control. Thus, the enriched
+terms are relevant to the studied project, yet while they do reflect real
+processes linked to treatment, they are not demonstrative of differences
+between the study groups. Finally, we use literature search to show that
+this incorrect approach to study group-specific treatment effects is not
+uncommon.
 
@@
 the groups G1 and G2 separately using standard bioinformatic tools. For each comparison, we defined
-genes with significantly different gene expression 
-(differentially expressed genes, DEGs) 
+DEGs
 by using an false discovery rate (FDR) threshold of 
@@
 `r sum(!res.merged$DEG.g1 & res.merged$DEG.g2)` 
-were significant in G2 only ("specific" for G2; see Fig. 1, panel *A*).
+were significant in G2 only (see Fig. 1, panel *A*).
 A naive interpretation of these results implies that there is a substantial
@@
 selected only genes which are DEGs in that group, but not the other,
-mimicking a naive approach for finding "specifically" regulated pathways.
+mimicking a naive approach for finding pathways regulated in one patient
+group only.
 Here, a similar picture
@@
 
-### Influence of sample size and cut-off thresholds on number of artifacts
-
-In the example above, the groups have been randomly sampled from a larger
-data set only once. To investigate whether the sampling had an impact on
-the artifacts, we have repeated the above procedure for 100 replicates,
-each containing a different set of samples randomly assigned to the two
-groups.  Furthermore, we tested whether the selection of the log~2~ fold
-change threshold
-or different sample size might have an impact on the extent of the arising
-artifacts. To this end, we have tested `r length(lfc_thresholds)` different
-threshold values for the log~2~ fold changes and three sample sizes: 20, 40 or 80 samples
-per group (corresponding to 10, 20 or 40 samples per group/treatment
-combination).
-
-**Fig. 3.Influence of log~2~ fold change treshold (A, C) and the sample
-size (B, D) on the number of genes significant in one, but not in the other
-group (A, B) and the number of significant gene sets found in one group, but
-not the other (C, D).**  Top row shows the influence of log~2~ fold change
-threshold for sample size of 40 per group (20 per group/treatment
-combination), bottom row shows the influence of sample size for a log~2~
-fold change threshold set to `r lfc_thr`.
-
-```{r montecarlo,cache=TRUE,fig.width=5.5,fig.height=5.5}
-## note: given that the monte carlo simulation is quite time consuming, we
-## have included the code for it in another file, `montecarlo.R`. If trying
-## to replicate this manuscript, please execute that file first to generate
-## the necessary replicates.
-
-outdir <- "montecarlo"
-
-merged_res <- readRDS(file.path(outdir, "merged_res.rds"))
-
-lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
-
-by_lfc_thr <- map_dfc(lfc_thresholds, ~ 
-                      get_specific_num(merged_res[["10"]], lfc_thr=.x,
-                                       padj_thr=padj_thr))
-colnames(by_lfc_thr) <- lfc_thresholds
-by_lfc_thr <- by_lfc_thr %>% pivot_longer(everything(), names_to = "lfc")
-
-g1 <- ggplot(by_lfc_thr, aes(x=lfc, y=value)) + geom_boxplot() +
-  ylab("Number of DEGs") + xlab("log2 fold change threshold")
-
-gsea_files <- list.files(path=outdir, pattern="gsea_.*rds", full.names=TRUE)
-gsea_by_lfc_thr_files <- gsea_files[ grepl("gsea_res_10", gsea_files) ]
-names(gsea_by_lfc_thr_files) <- gsub("montecarlo/gsea_res_10_([0-9.]*)\\.rds", "\\1",
-                                     gsea_by_lfc_thr_files)
-gsea_by_lfc_thr_files <- gsea_by_lfc_thr_files[ order(as.numeric(names(gsea_by_lfc_thr_files))) ]
-gsea_by_lfc_thr <- map_dfc(gsea_by_lfc_thr_files, gsea_specific_num)
-gsea_by_lfc_thr <- gsea_by_lfc_thr %>% pivot_longer(everything(), names_to = "lfc")
-
-g2 <- ggplot(gsea_by_lfc_thr, aes(x=lfc, y=value)) + geom_boxplot() +
-  ylab("Number of enriched gene sets") + xlab("log2 fold change threshold")
-
-by_ssize <- map_dfc(c("5", "10", "20"), 
-                    ~ get_specific_num(merged_res[[.x]], lfc_thr=1,
-                                       padj_thr=padj_thr))
-colnames(by_ssize) <- c(10, 20, 40)
-by_ssize <- by_ssize %>% pivot_longer(everything(), names_to = "ss")
-
-g3 <- ggplot(by_ssize, aes(x=ss, y=value)) + geom_boxplot() +
-  ylab("Number of DEGs") + xlab("Number of samples per group/treatment")
-
-gsea_by_ss_files <- gsea_files[ grepl("gsea_res_[0-9]*_1\\.rds", gsea_files) ]
-names(gsea_by_ss_files) <- gsub("montecarlo/gsea_res_([0-9]*)_.*\\.rds", "\\1",
-                                     gsea_by_ss_files)
-gsea_by_ss_files <- gsea_by_ss_files[ order(as.numeric(names(gsea_by_ss_files))) ]
-gsea_by_ss <- imap(gsea_by_ss_files, ~ {
-                     data.frame(ss=.y, value= gsea_specific_num(.x))
-                                     })
-gsea_by_ss <- Reduce(rbind, gsea_by_ss)
-gsea_by_ss$ss <- factor(as.numeric(gsea_by_ss$ss) * 2, levels=c(10, 20, 40))
-
-g4 <- ggplot(gsea_by_ss, aes(x=ss, y=value)) + geom_boxplot() +
-  ylab("Number of enriched gene sets") + xlab("Number of samples per group/treatment")
-
-plot_grid(g1, g2, g3, g4, labels = LETTERS[1:4])
-```
-
 ### Artifacts arise because of false negatives
@@
 To illustrate this phenomenon, we have analysed the full data set from which G1 and G2
-were drawn (Fig. 4), comparing the 
+were drawn (Fig. 3), comparing the 
 `r sum(covar$disease == "no") `
@@
 
-**Fig. 4. Area-proportional Venn diagram showing overlaps in DEGs between
+**Fig. 3. Area-proportional Venn diagram showing overlaps in DEGs between
 G1, G2 and the full data set.** The majority of genes which have been labeled
@@
 
-## Discussion
+### Influence of sample size and cut-off thresholds on number of artifacts
+
+In the example above, the groups have been randomly sampled from a larger
+data set only once. Arguably, the observations might differ if the groups
+were to be resampled. Furthermore, we have chosen a group size of 
+`r n_stratum * 4` (`r n_stratum * 2` per group/treatment combination). 
+Larger sample sizes are known to increase robustness of gene set enrichment
+analysis, and group size of 20 has been shown to be relatively robust [@maleki2019size].
+However, a smaller or larger group size might change the proportion of FNs in the
+results and thus influence the results of gene set enrichment. Finally,
+we have used a log~2~ fold change threshold of `r lfc_thr`, because raising
+it would increase the fraction of FNs. However, selecting more biologically
+only relevant genes which a substantially higher effect size may influence
+the observed enrichments. On the other hand, raising the log~2~ fold change
+threshold or lowering the sample size may lead to a smaller number of
+defined DEGs, thus making the hypergeometric test less powerful and lead to
+fewer gene set enrichment.
+
+To investigate whether the sampling had an impact on
+the artifacts, we have repeated the above procedure for 100 replicates,
+each containing a different set of samples randomly assigned to the two
+groups.  Furthermore, we tested whether the selection of the log~2~ fold
+change threshold or different sample size might have an impact on the
+extent of the arising artifacts. To this end, we have tested 
+`r length(lfc_thresholds)` 
+different threshold values for the log~2~ fold
+changes and three sample sizes: 20, 40 or 80 samples per group
+(corresponding to 10, 20 or 40 samples per group/treatment combination).
+
+**Fig. 4. Influence of log~2~ fold change treshold (A, C) and the sample
+size (B, D) on the number of genes significant in one, but not in the other
+group (A, B) and the number of significant gene sets found in one group, but
+not the other (C, D).**  Top row shows the influence of log~2~ fold change
+threshold for sample size of 40 per group (20 per group/treatment
+combination), bottom row shows the influence of sample size for a log~2~
+fold change threshold set to `r lfc_thr`.
+
+```{r montecarlo,cache=TRUE,fig.width=7.5,fig.height=6.5}
+## note: given that the monte carlo simulation is quite time consuming, we
+## have included the code for it in another file, `montecarlo.R`. If trying
+## to replicate this manuscript, please execute that file first to generate
+## the necessary replicates.
+
+outdir <- "montecarlo"
+
+merged_res <- readRDS(file.path(outdir, "merged_res.rds"))
+
+lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
+
+fdr_thresholds <- c(.8, .7, .6, .5, .4, .3, .2, .1, .05, .01, .001, .0001, 0.00001, 1e-6)
+by_fdr <- map_dfc(fdr_thresholds, ~ {
+                      a <- get_specific_num(merged_res[["10"]], lfc_thr=0, padj_thr=.x)
+                      b <- get_total_num(merged_res[["10"]], lfc_thr=0, padj_thr=.x)
+                      (b - a)/b }
+)
+colnames(by_fdr) <- fdr_thresholds
+by_fdr <- by_fdr %>% pivot_longer(everything(), names_to = "lfc")
+by_fdr$lfc <- factor(by_fdr$lfc, levels=fdr_thresholds)
+g0 <- ggplot(by_fdr, aes(x=lfc, y=value)) + geom_violin() +
+  ylab("Number of DEGs") + xlab("FDR threshold")
+
+by_lfc_thr <- map_dfc(lfc_thresholds, ~ 
+                      get_specific_num(merged_res[["10"]], lfc_thr=.x,
+                                       padj_thr=padj_thr))
+colnames(by_lfc_thr) <- lfc_thresholds
+by_lfc_thr <- by_lfc_thr %>% pivot_longer(everything(), names_to = "lfc")
+
+g1 <- ggplot(by_lfc_thr, aes(x=lfc, y=value)) + geom_boxplot() +
+  ylab("Number of DEGs") + xlab("log2 fold change threshold")
+
+gsea_files <- list.files(path=outdir, pattern="gsea_.*rds", full.names=TRUE)
+gsea_by_lfc_thr_files <- gsea_files[ grepl("gsea_res_10", gsea_files) ]
+names(gsea_by_lfc_thr_files) <- gsub("montecarlo/gsea_res_10_([0-9.]*)\\.rds", "\\1",
+                                     gsea_by_lfc_thr_files)
+gsea_by_lfc_thr_files <- gsea_by_lfc_thr_files[ order(as.numeric(names(gsea_by_lfc_thr_files))) ]
+gsea_by_lfc_thr <- map_dfc(gsea_by_lfc_thr_files, gsea_specific_num)
+gsea_by_lfc_thr <- gsea_by_lfc_thr %>% pivot_longer(everything(), names_to = "lfc")
+
+g2 <- ggplot(gsea_by_lfc_thr, aes(x=lfc, y=value)) + geom_boxplot() +
+  ylab("Number of enriched gene sets") + xlab("log2 fold change threshold")
+
+by_ssize <- map_dfc(c("5", "10", "20"), 
+                    ~ get_specific_num(merged_res[[.x]], lfc_thr=1,
+                                       padj_thr=padj_thr))
+colnames(by_ssize) <- c(10, 20, 40)
+by_ssize <- by_ssize %>% pivot_longer(everything(), names_to = "ss")
+
+g3 <- ggplot(by_ssize, aes(x=ss, y=value)) + geom_boxplot() +
+  ylab("Number of DEGs") + xlab("Number of samples per group/treatment")
+
+gsea_by_ss_files <- gsea_files[ grepl("gsea_res_[0-9]*_1\\.rds", gsea_files) ]
+names(gsea_by_ss_files) <- gsub("montecarlo/gsea_res_([0-9]*)_.*\\.rds", "\\1",
+                                     gsea_by_ss_files)
+gsea_by_ss_files <- gsea_by_ss_files[ order(as.numeric(names(gsea_by_ss_files))) ]
+gsea_by_ss <- imap(gsea_by_ss_files, ~ {
+                     data.frame(ss=.y, value= gsea_specific_num(.x))
+                                     })
+gsea_by_ss <- Reduce(rbind, gsea_by_ss)
+gsea_by_ss$ss <- factor(as.numeric(gsea_by_ss$ss) * 2, levels=c(10, 20, 40))
+
+g4 <- ggplot(gsea_by_ss, aes(x=ss, y=value)) + geom_boxplot() +
+  ylab("Number of enriched gene sets") + xlab("Number of samples per group/treatment")
+
+plot_grid(g1, g2, g3, g4, labels = LETTERS[1:4])
+```
+
+Setting a higher log~2~ fold change threshold reduces the number of
+DEGs as well as of the observed artifactual gene set enrichments 
+(Fig. 4, A and B). However, even for log~2~
+threshold of 3 (DEGs defined by 8-fold change and FDR < `r padj_thr`) the
+number of replicates in which artifacts can be observed is 
+`r sum(gsea_by_lfc_thr %>% filter(lfc == 3) %>% pull(value) > 0)`
+(out of 100),
+and in at least 
+`r sum(gsea_by_lfc_thr %>% filter(lfc == 3) %>% pull(value) >= 5)`
+replicates, 5 or more gene sets were significantly
+enriched. Thus, setting a more conservative threshold while retaining the
+incorrect procedure cannot fully protect from the arising artifacts.
+The number of artifacts rose with sample size
+(Fig. 4, C and D). For total sample size of 160 (40 samples per
+group/treatment combination) the number of artifacts was almost 
+`r round(
+  mean(gsea_by_ss %>% filter(ss == 40) %>% pull(value)) /
+  mean(gsea_by_ss %>% filter(ss == 20) %>% pull(value)))
+`
+times higher than for total sample size of 80.
+
+### Incorrect analysis of interactions is common in transcriptomics
+
+@nieuwenhuis2011erroneous observed incorrect analyses of interactions in
+about half of the analysed papers from top neuroscience journals where the
+authors considered an interaction. We wanted to know if this problem is
+common in transcriptomics, too. To this end, we have searched three
+journals – from broad to specialized – for the occurence of the terms
+"differential expression" with "venn diagrams" (Tab. 2). Next, we analysed
+the selected papers from year 2020 (and years 2015-2020 in one case) to
+decide whether the VD was described or referred to as showing genes
+"specific" or "unique" to a particular condition or whether a test for
+interaction was performed. Finally, we checked whether gene set enrichment
+analysis was applied to genes significant in one, but not the other
+condition in order to find gene sets and pathways regulated by one, but not
+the other condition.
 
@@
 
-In an informal survey of articles from three journals (Tab. 2) we found
+We found 
 that of the 
 `r tab2$Analysed[nrow(tab2)]` analysed articles
- which used the term "venn diagram" and
+ which used the terms "venn diagram" and
 "differential expression", at least 
@@
 coupled the VDs with some form of gene set enrichment analysis on the
-set of supposedly "specific" DEGs.
+set of supposedly "specific" DEGs. In summary, in at least a quarter of the
+papers on differential expression in which a venn diagram was used, it was
+illustrating an incorrect statistical procedure which may result in
+artifactual gene set enrichments.
+
+## Discussion
 
@@
 same level of evidence as rejecting it. However, when such an incorrect
-analysis is combined with downstream functional analysis – i.e., gene set
-enrichments of genes "specific" to one or another comparison – the
-resulting pathways or gene ontologies are misleadingly relevant, yielding
-gene sets associated with immune answer for research hypotheses involving
-an infectious disease, or cancer pathways if the underlying research
-hypothesis involved cancer treatment. Such results may be hard to resist
-for an involved researcher, especially if the correct analysis of
-interactions does not show any significant differences.
+analysis is combined with downstream functional analysis, the
+resulting pathways or gene ontologies are misleadingly relevant. For
+example, the identified gene sets are associated with immune answer for
+research hypotheses involving an infectious disease, or cancer pathways if
+the underlying research hypothesis involved cancer treatment. Such results
+may appear reasonable in the given context, especially if the correct
+analysis of interactions does not show any significant differences. This
+effect is persistent or even exacerbated for larger sample sizes (Fig.
+4).
 
-That is not to say that VDs are not a useful tool, even in the context of
-transcriptomics and gene set enrichments, if used correctly. For example, analysis of an
+We found that this type of incorrect analysis occurs in more than a quarter
+of papers where the procedure was illustrated with a VD.  That is not to
+say that VDs are not a useful tool, even in the context of transcriptomics
+and gene set enrichments, if used correctly. For example, analysis of an
 intersection of DEGs (i.e., by considering genes from the overlap in a VD)
@@
 
+Our literature survey showed that even in cases where a VD is absent from a
+paper, at least two other ways are not uncommon. Firstly, the direct
+comparison of gene set enrichment results: that is, drawing conclusions
+from the fact that a gene set enrichment result was significant in one
+comparison only.  Second, while Venn diagrams are often used to illustrate
+the numbers of "specific" DEGs and so present a mean to find examples of
+this fallacy in scientific literature, researches often test for enrichment
+these "specific" genes without using the phrase "Venn diagram" or even
+clearly stating how the lists of "specific" genes were derived. In all
+these cases, the analysis boils down to comparing results significant in
+one, but not in another comparison.
+
 As an alternative to Venn diagrams and the downstream gene set enrichment
 analysis, two approaches can be considered.
-
-Statistical analysis, as shown above, involves a test for interaction which
+The correct statistical approach, as shown above, involves a test for interaction which
 can reveal genes for which the impact of treatment significantly differ
@@
 
-While the use of VDs in combination with the incorrect gene set
-enrichment analysis is common, it is not the only way of achieving
-artifactual and incorrect conclusions by means of comparing significant
-with non-significant results in transcriptomics and other high-throughput
-applications. VDs appear to be symptomatic of the assumption that if a gene
-is significantly regulated in one comparison, but does not show a
-significant difference in another one, then this gene is "specific" to this
-first comparison. However, VDs may also lead one to this reasoning; for
-example, the VD on Fig. 1A practically begs the question: what do these 
-`r sum(res.merged$DEG.g1 & !res.merged$DEG.g2)` 
-genes, which are significant in G1 only, have in common?
-
-Our literature survey showed that even in cases where a VD is absent from a
-paper, at least two other ways are not uncommon. Firstly, the direct
-comparison of gene set enrichment results: that is, drawing conclusions
-from the fact that a gene set enrichment result was significant in one
-comparison only.  Second, while Venn diagrams are often used to illustrate
-the numbers of "specific" DEGs and so present a mean to find examples of
-this fallacy in scientific literature, researches often test for enrichment
-these "specific" genes without using the phrase "Venn diagram" or even
-clearly stating how the lists of "specific" genes were derived. In all
-these cases, the analysis boils down to comparing results significant in
-one, but not in another comparison.
-
-The use of Venn diagrams to illustrate specific differences between
-comparisons should therefore be abandoned in favor of statistically correct
-approaches. Furthermore, gene set enrichment analysis must never be applied
+Defining condition-specific genes based on significant difference in one,
+but no significant difference in another comparison is thus not only a
+statistical fallacy, but when combined with gene set enrichment analysis
+also one which can lead to potentially sound-looking results. This method
+of obtaining specific differences between conditions should therefore be
+abandoned in favor of statistically correct approaches.
+Furthermore, gene set enrichment analysis must never be applied
 to sets of genes defined as significant in one comparison, but not the
@@
 
-An informal literature survey was performed using Google Scholar to
+A literature survey was performed using Google Scholar to
 estimate the frequency of the incorrect use of Venn diagrams. We searched
@@
 considered incorrect.
+The results of the literature survey (including links to papers classified
+as incorrectly analysing the interaction) are included in the manuscript
+sources.
 

### 2022-01-16 presentation/venn_diagrams_weiner_2022_01_06.rmd
major rewrite

@@
 
-```{r fig3,dpi=300,dev="png"}
+.pull-left[
+
+```{r fig3,width=5,height=5,dpi=300,dev="png"}
 tmp <- merge(res.merged, res$full, by="ENSEMBL")
@@
 
+]
+
+.pull-right[
+
+In a larger data set, majority of the "specific" genes are significant.
+
+]
+
 ---
@@
 
-```{r montecarlo,cache=TRUE,fig.width=7.5,fig.height=6.5}
+ * Increase the sample size to have a higher sensitivity (fewer FNs)
+
+ * Increase the LFC threshold to have fewer false positives
+
+---
+
+## Can we get rid of it in another way?
+
+```{r montecarlo,cache=TRUE,fig.width=7.5,fig.height=5.5}
 ## note: given that the monte carlo simulation is quite time consuming, we
@@
 
-plot_grid(g1, g2, g3, g4, labels = LETTERS[1:4])
+plot_grid(g3, g4, labels = LETTERS[1:2])
 ```
@@
 ---
+
+```{r montecarlo2,cache=TRUE,fig.width=7.5,fig.height=5}
+plot_grid(g1, g2, labels = LETTERS[1:2])
+```
+
+---
+
+## Thank you!
+
+.pull-left[
+
+Source code for this presentation:
+[https://github.com/bihealth/manuscript_venn_diagrams](https://github.com/bihealth/manuscript_venn_diagrams)
+
+]
+
+.pull-right[
+
+```{r}
+library(qrcode)
+plot(qr_code("github.com/bihealth/manuscript_venn_diagrams"))
+```
+
+]

### 2022-01-16 mk4/functions.R

@@ -106,4 +106,18 @@ get_specific_num <- function(merged_res, lfc_thr=0, padj_thr=0.05) {
   n_spec
 }
 
+## returns a numeric vector. Each number represents the number of
+## DEGs (in both groups) for one replicate.
+get_total_num <- function(merged_res, lfc_thr=0, padj_thr=0.05) {
+
+  n_spec <- map_int(merged_res, ~ {
+    res <- .x %>%
+      mutate(DEG.g1 = !is.na(padj.g1) & abs(log2FoldChange.g1) > lfc_thr & padj.g1 < padj_thr) %>%
+      mutate(DEG.g2 = !is.na(padj.g2) & abs(log2FoldChange.g2) > lfc_thr & padj.g2 < padj_thr) 
+    sum((res$DEG.g1 | res$DEG.g2))
+  })
+
+  n_spec
+}
+
 

### 2022-01-18 mk4/response_to_editor.md
forgotten update

--- 2022-01-05 mk4/response_to_editor.md
@@
 specific for a certain comparison". It is just a visualization tool. 
-
-**We do agree in principle. However, the term "venn diagram analysis" (and
-related formulations) is firmly (even though rather unfortunately)
-established in the field of differential expression and many authors are
-using it essentially as a shorthand for stating that the supposedly
-treatment specific genes are found by comparing which genes are significant
-in one, but not the other comparison. We have now explained this in more
-detail in the Introduction as follows:**
-
-XXX
-
 The flaw
@@
 
-**We fully agree; we have already addressed this issue in the
-"Discussion".**
-
 The title and abstract are misleading and more journalistic
@@
 would be to condemn the use of a knife for murders. 
-
-**We understand that point. However, we do observe a substantial fraction of
-papers using Venn diagrams misinterpreting the results. The given analogy
-would only work if every fourth use of a knife resulted in a homicide. This
-point has already been addressed in the Discussion. Nonetheless we have now
-removed references to Venn diagrams in the title and in the abstract.**
-
 This style is misleading
@@
 
-**We have completely rewritten the title and the abstract and hope that
-they are now acceptable.**
+**We agree with these observations and understand your concerns. We have
+reformulated abstract and discussion to state more clearly that Venn
+diagrams are not the cause of flawed statistics but are very easily
+misinterpreted and a readily identifiable symptom of erroneous statistical
+reasoning, since we do observe a substantial fraction of papers using Venn
+diagrams misinterpreting the results. Extending your metaphor, if knives
+were like VDs in this context, then every fourth use of a knife would
+result in a homicide.**
+
+**We regret if our manuscript is perceived as too journalistic, even though
+we think that communicating such a matter is to a large part an educational
+mission and requires somewhat less complex and more pointed language. We
+hope that our rewritten title, abstract and discussion strike a good
+balance between scientific restraint and accessibility for
+non-specialists.**
 
@@
 
-**We do not want to single out a paper or point fingers at specific
+**We do not want to single out specific papers or point fingers at specific
 authors.  However, given that the sources of all our calculations are
-available from github, we placed there the list of links to papers which –
+available from github, we placed there the lists of links to papers which –
 in our view – have incorrectly interpreted Venn diagrams. For example, the
@@
 
-**They have been all along. We have now included the following statement in the
+**We have now included the following statement in the
 Methods section:**
@@
 sources."*
-
-XXX
 
@@
 
-**This conclusion is based on the assumption that the fraction of FP among
+**Our conclusion is based on the assumption that the fraction of FP among
 significant genes is controlled by the Benjamini-Hochberg procedure, in
 other words, that the false discovery rate (FDR) is not more than 5% (as
-stated in the quoted paragraph).**
+stated in the quoted paragraph). This implies that in the whole data set
+there are at most 30 DEGs which are FP. Since 199+99 = 298
+supposedly specific genes are significant when analysing the whole data
+set, even if all 30 FPs are among them, there are still 268 genes which are
+not FP and which appear to be group specific. 100% * 268 / (431 + 278) =
+37%. We have now made these simple calculations more explicit in the
+text.**
 
@@
 smaller sample sizes for each group. 
-
-**We agree, but the total number of samples in the analysis is still 80,
-with 40 samples per group and 20 per group/treatment combination. This is
-not an unusually small sample size for this type of transcriptomic
-analysis (see below for more on the chosen sample size). Also, the design of our in
-silico experiment *requires* splitting the data set into two randomly selected groups,
-because that is the whole idea of the paper: take a situation in which 
-there should not be any significant enrichments (because there are no
-real differences between the groups) and show that with an incorrect (but
-widely spread) procedure you will get convincing evidence for differences
-between groups.**
-
 This comes with two major shortcomings:
@@
 and outliers, resulting in false positives
-
-**We find the claim that the number of false positives is higher for small
-sample sizes surprising. Statistical theory tells us that the FDR is
-controlled by the BH false discovery rate controlling procedure.  If BH
-(which is a widely accepted method for the analysis of DE data) is correct,
-we should not expect more than 5% FP among significant results
-independently of sample size.  See Benjamini, Yoav, and Yosef Hochberg.
-"Controlling the false discovery rate: a practical and powerful approach to
-multiple testing." Journal of the Royal statistical society: series B
-(Methodological) 57.1 (1995): 289-300. We are not aware of any sources that
-would claim otherwise.**
-
 , and (2) the test for differential
@@
 
-**We fully agree that FNs are the core of the problem! However, we know
-from other studies (and basic statistics) that even in most favorable
-circumstances the FNR will likely be above 20%; and that in case of transcriptomic
-studies, the FNR may be much larger (please see the relevant fragment already
-included in the paper). In other words, it does not matter whether one
-looks at 80 samples or 193 samples: FNs are bound to occur. That is the
-reason – we claim – for the fact that the gene set enrichments found in
-this manner are related to the studied problem. However, we have addressed
-this isssue by testing different sample sizes (see below).**
+**We agree that splitting the dataset results in decreased statistical
+power. However, we don't find it obvious that the false discovery rate will
+increase, since it's controlled by the standard BH false discovery rate
+procedure and the FDR cutoff of 0.05. Instead, we fully agree that FNs are
+the core of the problem and used the FDR cutoff of 0.05 to not unduely
+inflate the number of false negatives (a more conservative threshold would
+result in a higher false negative rate). Even in most favorable
+circumstances the FNR will likely be above 20%, especially for
+transcriptomic studies (see also
+https://academic.oup.com/bib/article/19/4/713/2920205).**
 
-Therefore,
-the design of the experiment is biased in favour of the "desired" conclusion.
+**In any case, there are 80 samples in our analysis (40 samples per
+group and 20 per group/treatment combination), which is not a small sample
+size for this type of transcriptomic analysis. In fact, a survey on RNA
+sample size in real world data (Baccarella et al.) finds that half of the
+studies in humans used sample sizes sizes of 6 or less per group. Nevertheless, we
+welcome the suggestion to analyze the sample size dependence: in fact, the
+larger the sample size, the more artefactual enrichments are observed,
+because the differential expression yields more results and the
+hypergeometric test gets more powerful with more DEGs. We have now included
+a section comparing the results for 10, 20 (our example) and 40 samples per
+group/treatment combination (thus, total sample sizes of 40, 80 and 160).
+For each sample size, we have run 100 replicates of the group sampling,
+replicated the whole procedure and gathered all test results. We hope that
+the new results shown ion the new Fig. 4, C and D provide a useful
+illustration of this reasoning.**
 
-**We have mimicked an RNA-Seq experiment using a sample size
-well above the median used by others precisely to avoid this allegation. We
-have now included a relevant reference (Baccarella et al.) which includes a
-literature survey on RNA sample size in real world data. In there you will
-see that half of the studies used sample sizes sizes of 6 or less per
-group.
+Therefore, the design of the experiment is biased in favour of the
+"desired" conclusion.
 
-Nonetheless, we think that showing how the artifacts depend on sample size
-is, in itself, interesting, not least because it shows that sample size has
-a direct effect on the number of observed artifacts: the larger the sample
-size, the more artifacts are observed. We have now included a section
-comparing the results for 10, 20 (our example) and 40 samples per
-group/treatment combination (thus, total sample sizes of 40, 80 and 160).
-The reason for this is, of course, that the more DEGs you observe, the more
-powerfull the hypergeometric test is.
+**We respectfully disagree. Our approach is quite similar to
+cross-validation and a straightforward way to construct two groups that
+should not have differences and then show how an incorrect (but widely
+used) statistical procedure will result in convincing evidence for
+differences between them.**
 
-For each sample size, we have run 100 replicates of the group sampling,
-replicated the whole procedure and gathered all test results. Results are
-shown on the new Fig. 3, C and D.
-
-However, we also think that the core of the problem is the incorrect
+**Essentially, we think that the core of the problem is the incorrect
 statistical procedure (selecting genes by comparing significance with lack
-thereof) which will not become more correct by increasing the sample size,
-even should we fail to observe any artifacts (which is not the case,
-artifacts appear even for abs(lfc) >= 3).**
-
-XXX
+thereof), which will not become more correct by increasing the sample size.**
 
@@
 
-**Statistical theory (e.g. the above reference to the original Benjamini &
-Hochberg paper) tells us that the false discovery rate in differential
-expression analysis is controlled by the BH procedure, not by the log fold
-change threshold, except incidentally (beause log fold change is correlated
-to the p-value, thus setting a more conservative log2 fold change results
-in eliminating most of the larger p-values). Genes with large fold changes
-may still not be significant and vice versa, true positives may have small
-log2 fold changes. Therefore you cannot predict what will be the precise
-effect of raising the LFC threshold on FDR except in very broad terms,
-whereas setting the FDR threshold gives you clear information about the
-expected proportion of false positives in your results.
-
-This is analogous to using an effect size filter (say, Cohen's d) on
-results of a t-test. The p-value from the t-test controls the type I error
-rate, but a statistically significant difference does not necessarily
-indicate a biologically meaningfull effect. For this, we turn to the effect
-size (see for example Sullivan, Gail M., and Richard Feinn. "Using effect
-size—or why the P value is not enough." Journal of graduate medical
-education 4.3 (2012): 279-282.).
-
-Thus, the threshold serves to eliminate genes that do show statistically
-significant change (even if they are true positives), but which are not
-biologically relevant (because the effect size, for which log2 fold change
-is a proxy, is too small). The proportion of false positives is 5% even with no
-log2 FC treshold.  Nonetheless, we have compared the effect of the log2
-fold change threshold on the results.**
-
-Higher values of log-fold change should also be used, and the results
-should be compared and discussed.
+**log2-fold change cutoffs (on top of FDR control) are widely used in the
+literature, and a cutoff of 1 is in our experience not small for human
+data. The threshold serves to eliminate genes that show statistically
+significant change (they are true positives), but are not biologically
+relevant (because the effect size, for which log2 fold change is a proxy,
+is too small). In any case, the false discovery rate in differential
+expression analysis is controlled via BH by a cutoff on the p-value, not
+the log fold change, except indirectly (beause log fold change is
+correlated to the p-value, setting a more conservative log2 fold change
+results in eliminating most of the larger p-values). The precise effect of
+raising the LFC threshold on FDR is therefore unclear except in very broad
+terms.**
 
 **We have now included a section in the manuscript which shows the
-dependence of the artifacts on the log2 fold change thresholds used.
-For this, we have replicated the whole procedure of group randomization 100
-times, and repeated the analysis for different threshold values. Summary
-overview can be found in the new Fig. 3.**
+dependence of the artifacts on the log2 fold change thresholds used. For
+this, we have replicated the whole procedure of group randomization 100
+times. Summary overview can be found in the new Fig. 4. We find that
+increasing the threshold can, of course, result in not observing any
+enrichments (see Fig. 4, A and B). Without a sufficient number of DEGs, no
+enrichments can be calculated using a hypergeometric test. However, even
+though increasing the log2 fold change cutoff may sometimes eliminate the
+symptoms, it does not remedy the core problem of erroneous analysis of
+interactions. Many replicates in the simulations have artifactual
+enrichments even at high log fold change thresholds.**
 
-The paper also needs to be edited to have a more scientific tone, improve long and complicated sentences, and correct the typos and grammatical errors.
+The paper also needs to be edited to have a more scientific tone, improve
+long and complicated sentences, and correct the typos and grammatical
+errors.
 

### 2022-01-18 mk4/manuscript.rmd
forgotten update

@@
 ---
-title: ""
+title: "Venn diagrams as indicators of erroneous statistical reasoning in transcriptomics."
 author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
@@
 
-Version 1: In statistics, comparing the effect of a treatment in two different
-conditions requires the use of an appropriate statistical test to assert
-the significance of interaction between the treatment and the condition.
-However, a common procedure is to define genes specifically regulated in
-one, but not the other condition by selecting genes which show a
-statistically significant difference in one condition only. Here we show
-that not only is this procedure incorrect, but moreover, when combined with
-gene set enrichment analyses, it leads to statistically significant, but
-incorrect results. These gene set enrichments are the effect of false
-negatives and reflect only the differences in gene expression due to
-treatment and not due to the difference between conditions. Thus, the
-combination of incorrect statistical analysis and gene set enrichment
-analysis may result in particularly misleading artifacts.
-
-Version 2: A common application of differential expression analysis is to
-ask which genes are differentially expressed upon treatment in one
-condition only. One of the approaches is to define genes specific to a
-condition by testing for significant difference in expression separately
-in the two conditions, and then selecting genes which show statistical
-significance in one condition only. This approach may then be combined with
+A common application of differential expression analysis is to
+find genes that are differentially expressed upon treatment in only one
+out of several groups of samples. One of the approaches is to test 
+for significant difference in expression between treatment and control separately
+in the two groups, and then select genes that show statistical
+significance in one group only. This approach is then often combined with
 a gene set enrichment analysis to find pathways and gene sets
-characteristic for this condition only.
-
-Here we show that this procedure is incorrect. Instead, the interaction
-between treatment and condition should be tested. Moreover, we show that
-gene set enrichment analysis applied to incorrectly defined genes specific
-to one condition only may result in misleading artifacts. Genes which show
-significant difference upon treatment in one, but not the other condition
-are enriched in gene sets related to the treatment, but not to the
-condition. Thus, while the results are apparently related to the treatment,
-they do not reflect differences in expression specific to a given
-condition.
-
-Version 3:
+regulated by treatment in only this group.
+
+Here we show that this procedure is statistically incorrect and that the interaction
+between treatment and group should be tested instead. Moreover, we show
+that gene set enrichment analysis applied to such incorrectly defined genes
+group-specific genes only may result in misleading artifacts. Genes which
+show significant difference upon treatment in one, but not another group
+are enriched in gene sets associated with the treatment in general.  Thus,
+the results appear related to the problem at hand, but do not reflect the
+group-specific effect of a treatment. A literature search revealed that
+more than a quarter of papers which used a Venn diagram to illustrate the
+results of separate differential analysis  have also applied this incorrect
+reasoning.
 
@@
 lfc_thr  <- 1
-n_stratum <- 10 # 2 strata (male and female) per condition/group
-                # total number of samples n_stratum * 2 * 2 * 2
+n_stratum <- 10 # 2 strata (male and female) per group/treatment
+                # total number of samples = n_stratum * 2 * 2 * 2
 lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
@@
 
-Transcriptomic analyses commonly include more than one factor. Often, the
-question asked is whether there is a difference between conditions (first
+Experimental designs for transcriptomic analyses commonly include more than one factor. Often, the
+question asked is whether there is a difference between groups (first
 factor) with respect to reaction to a particular treatment (second
-condition). For example, we may ask whether there are differentially
+factor). For example, we may ask whether there are differentially
 expressed genes (DEGs) which are specific to a particular group of
-patients, such as interferon response elicited by a virus in one group, but
-silenced in the other group of patients. In other words, we ask whether the
+patients, e.g. interferon response elicited by a virus in one group, but
+not regulated in another group of patients. In other words, we ask whether the
 difference between control (healthy subjects) and treatment (infected
@@
 However, another approach is widely spread [@nieuwenhuis2011erroneous].
-Instead of testing the interaction, one tests the effect of the treatment
-separately in both groups. Next, a difference between conditions is inferred if the difference
+Instead of testing the interaction, the effect of the treatment is tested
+separately in both groups. Next, a difference between groups is inferred if the difference
 is significant in one comparison, but not significant in the other. This
@@
 
-In transcriptomics, the statistical tests are performed for thousands of
-genes. As in the general case, the correct inference of differences between
-the conditions is by testing the significance of interaction. However, in
-another approach the differences between treatment and control are tested
-in the two conditions separately, and the results are then compared. This
-can be visualized using a Venn diagram (VD, Fig. 1) showing the overall
-number of DEGs significant in both comparisons (the intersection in the VD)
-or significant in only one comparison (the remaining two fields on a VD).
-The genes which are significant in only one comparison may then be
-incorrectly defined as specific to the corresponding condition. Following
-this, gene set enrichment analysis may be applied to these genes in an
-attempt to test which pathways are specific to one, but not the other
-condition.
+In transcriptomics, statistical tests are performed for thousands of
+genes. As in the general case, the inference of differences between the
+groups should correctly be done by testing the significance of interaction
+between the group and the treatment. However, often the differences between
+treatment and control are tested in the two groups separately, and the
+results are then compared. This can be visualized using a Venn diagram (VD,
+Fig. 1) showing the overall number of DEGs significant in both comparisons
+(the intersection in the VD) or significant in only one comparison (the
+remaining two fields on a VD).  The genes which are significant in only one
+comparison may then be incorrectly defined as specifically to the corresponding
+group. Following this, gene set enrichment analysis may be applied to
+these genes in an attempt to test which pathways are specific to one, but
+not the other group.
 
@@
 result in apparent enrichments even if there are no real differences
-between the groups. To this end, we have randomly split a cohort into two
+between the groups. To this end, we randomly split a cohort into two
 groups, compared the treatment (viral infection) with controls in each of
@@
 this incorrect approach to study group-specific treatment effects is not
-uncommon.
+uncommon. In fact, while VDs are a useful visualisation tool also in
+transcriptomics, in more than quarter of the papers where VDs were used,
+group-specific genes were defined as significant in one, but not other
+groups, and in 19% a gene set enrichment was performed.
 
@@
 `r sum(res$g1$DEG)` 
-DE genes in the G1 group, and
+DEGs in the G1 group, and
 `r sum(res$g2$DEG)`
@@
 **Fig. 1. Results of differential gene expression analysis
-and gene set enrichment analysis using an incorrect Venn diagram driven approach.**
+and gene set enrichment analysis using an incorrect approach.**
 **A**, Venn diagram showing numbers of differentially expressed genes (DEG)
@@
 from a Venn diagram-driven gene set enrichment analysis are artefactual. Closer
-inspection of genes which belong to DEG in one group, but not the other
+inspection of genes which are DEGs in one group, but not the other
 reveals the underlying statistical fallacy (Fig. 2, *A* – *D*), that is, that difference
@@
 genes for which the adjusted p~adj~ < 0.05 there are at most 5% false
-positives.
+positives irrespective of the sample size.
 
@@
 genes significant in G1, but not in G2, 
-`r .n <- res$full %>% filter(DEG & ENSEMBL %in% g1_spec) %>% nrow; .n`
-(`r sprintf("%.0f%%", .n/length(g1_spec) * 100)`) 
+`r .n1 <- res$full %>% filter(DEG & ENSEMBL %in% g1_spec) %>% nrow; .n1`
+(`r sprintf("%.0f%%", .n1/length(g1_spec) * 100)`) 
 are significant in the full data set; 
@@
 genes significant in G2, but not in G1, 
-`r .n <- res$full %>% filter(DEG & ENSEMBL %in% g2_spec) %>% nrow; .n`
-(`r sprintf("%.0f%%", .n/length(g2_spec) * 100)`) 
+`r .n2 <- res$full %>% filter(DEG & ENSEMBL %in% g2_spec) %>% nrow; .n2`
+(`r sprintf("%.0f%%", .n2/length(g2_spec) * 100)`) 
 are significant in the full data set.
 Given that G1 and G2 were sampled from the total population, and since the
-FDR was set to `r padj_thr`, we can assume that at least between a third and a half of
+FDR was set to `r padj_thr`, we do not expect more than 
+`r .nfp <- ceiling(padj_thr * sum(res$full$DEG)); .nfp`
+FPs in the full data set, 
+which implicates that at least 
+`r .n1 + .n2 - .nfp` out of the `r length(g1_spec) + length(g2_spec)` 
+"specific" DEGs are true positives in the full data set.
+Thus, we can assume that at least a third of
 the genes that appeared to be "specific" in the initial analysis were, in
@@
 about half of the analysed papers from top neuroscience journals where the
-authors considered an interaction. We wanted to know if this problem is
+authors considered an experimental design allowing for a test for
+interaction. We wanted to know if this problem is
 common in transcriptomics, too. To this end, we have searched three
@@
 decide whether the VD was described or referred to as showing genes
-"specific" or "unique" to a particular condition or whether a test for
+"specific" or "unique" to a particular group or whether a test for
 interaction was performed. Finally, we checked whether gene set enrichment
 analysis was applied to genes significant in one, but not the other
-condition in order to find gene sets and pathways regulated by one, but not
-the other condition.
+group in order to find gene sets and pathways regulated by one, but not
+the other group.
 
@@
 
-Our literature survey showed that even in cases where a VD is absent from a
-paper, at least two other ways are not uncommon. Firstly, the direct
+While VDs appear to be frequently associated with an incorrect statistical
+reasoning, the use of VDs is not the cause.  In the absence of a VD
+illustrating the DEGs common and unique to the different study groups, two
+incorrect approaches may still be found. Firstly, the direct
 comparison of gene set enrichment results: that is, drawing conclusions
 from the fact that a gene set enrichment result was significant in one
-comparison only.  Second, while Venn diagrams are often used to illustrate
+comparison only.  Second, while VDs are often used to illustrate
 the numbers of "specific" DEGs and so present a mean to find examples of
-this fallacy in scientific literature, researches often test for enrichment
+this fallacy in scientific literature, researches test for enrichment
 these "specific" genes without using the phrase "Venn diagram" or even
@@
 
-Defining condition-specific genes based on significant difference in one,
+Defining group-specific genes based on significant difference in one,
 but no significant difference in another comparison is thus not only a
@@
 also one which can lead to potentially sound-looking results. This method
-of obtaining specific differences between conditions should therefore be
+of obtaining specific differences between groups should therefore be
 abandoned in favor of statistically correct approaches.
@@
 
+### Simulation study
+
+We have generated replicates of the example study for different log~2~ fold
+change thresholds (`r paste(lfc_thresholds, sep=", ")`) and three different
+sample sizes (40, 80 and 160, corresponding to sample size per
+group/treatment combination of 10, 20 and 40). For each replicate, the full
+procedure as described above was repeated, and numbers of DEGs and
+significantly enriched terms were collected.
+
 ### Literature survey
@@
 for articles from 2020 mentioning the phrases "differential expression" and "venn
-diagram" in the Scientific Reports journal. The journal was
-selected because it represents a wide spectrum of research areas, the
-articles are available through an open access license and there is a large
-number of publications per year. For each of the papers identified, we
+diagram" in three journals: Scientific Reports (2020), Nature Communications (2020) and
+Science Immunology (2015-2020). For each of the papers identified, we
 checked whether (i) the authors used the VD to show differentially expressed
@@
 analysis is any form. Articles which (i) focused only on the intersections
-of the Venn diagrams (genes common to all conditions), or (ii) which used the
+of the Venn diagrams (genes common to all groups), or (ii) which used the
 Venn diagrams for a purpose other than to compare genes significant in one
-condition, but not significant in another condition or (iii) for which a
+groups, but not significant in other groups or (iii) for which a
 clear-cut error could not be indicated past any reasonable doubt were not

### 2022-03-08 mk5/manuscript.rmd
version 5

--- 2022-01-18 mk4/manuscript.rmd
@@
 ---
-title: "Venn diagrams as indicators of erroneous statistical reasoning in transcriptomics."
+title: "Venn diagrams may indicate erroneous statistical reasoning in transcriptomics."
 author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
@@
 
-## Titles
-
-???
-
 ^1^ Berlin Institute of Health at Charité – Universitätsmedizin Berlin, Core Unit Bioinformatics, Charitéplatz 1, 10117 Berlin, Germany
@@
 
-A common application of differential expression analysis is to
-find genes that are differentially expressed upon treatment in only one
+A common application of differential expression analysis is 
+finding genes that are differentially expressed upon treatment in only one
 out of several groups of samples. One of the approaches is to test 
@@
 
-Here we show that this procedure is statistically incorrect and that the interaction
-between treatment and group should be tested instead. Moreover, we show
-that gene set enrichment analysis applied to such incorrectly defined genes
-group-specific genes only may result in misleading artifacts. Genes which
-show significant difference upon treatment in one, but not another group
-are enriched in gene sets associated with the treatment in general.  Thus,
-the results appear related to the problem at hand, but do not reflect the
-group-specific effect of a treatment. A literature search revealed that
-more than a quarter of papers which used a Venn diagram to illustrate the
-results of separate differential analysis  have also applied this incorrect
-reasoning.
+Here we show that this procedure is statistically incorrect and that the
+interaction between treatment and group should be tested instead. Moreover,
+we show that gene set enrichment analysis applied to such incorrectly
+defined genes group-specific genes may result in misleading artifacts.
+Due to the presence of false negatives, genes significant in one, but not
+the other group are enriched in gene sets which which correspond to the
+overall effect of the treatment.  Thus, the results appear related to the
+problem at hand, but do not reflect the group-specific effect of a
+treatment. A literature search revealed that more than a quarter of papers
+which used a Venn diagram to illustrate the results of separate
+differential analysis  have also applied this incorrect reasoning.
 
@@
 
-Experimental designs for transcriptomic analyses commonly include more than one factor. Often, the
+Experimental designs for transcriptomic analyses frequently include more than one factor. Often, the
 question asked is whether there is a difference between groups (first
@@
 patients, e.g. interferon response elicited by a virus in one group, but
-not regulated in another group of patients. In other words, we ask whether the
-difference between control (healthy subjects) and treatment (infected
-patients) is different between two groups of individuals. This "difference
-of differences" is known in statistics as an interaction
+absent in another group of patients. In other words, we ask whether the
+difference between the control group (healthy subjects) and the treatment
+group (infected patients) is different between two groups of individuals.
+This "difference of differences" is known in statistics as an interaction
 [@blalock1965theory]. To find out whether it is statistically significant,
@@
 Instead of testing the interaction, the effect of the treatment is tested
-separately in both groups. Next, a difference between groups is inferred if the difference
+separately in both groups. Next, a difference between groups is inferred if the 
+effect of treatment
 is significant in one comparison, but not significant in the other. This
@@
 significant" [@gelman2006difference]. For example, the p-value in the first
-comparison may be 0.009, and in the other comparison 0.011, thus at an
+comparison may be 0.009, and in the other comparison 0.011. At an
 alpha level of 0.01 the difference will be statistically significant in the
@@
 groups should correctly be done by testing the significance of interaction
-between the group and the treatment. However, often the differences between
-treatment and control are tested in the two groups separately, and the
-results are then compared. This can be visualized using a Venn diagram (VD,
-Fig. 1) showing the overall number of DEGs significant in both comparisons
-(the intersection in the VD) or significant in only one comparison (the
-remaining two fields on a VD).  The genes which are significant in only one
-comparison may then be incorrectly defined as specifically to the corresponding
-group. Following this, gene set enrichment analysis may be applied to
-these genes in an attempt to test which pathways are specific to one, but
-not the other group.
+between the group and the treatment. In practice, the differences between
+treatment and control are frequently tested in the two groups separately.
+This can be visualized using a Venn diagram (VD, Fig. 1) showing the
+overall number of DEGs significant in both comparisons (the intersection in
+the VD) or significant in only one comparison (the remaining two fields on
+a VD).  The genes which are significant in only one comparison are
+sometimes incorrectly considered as specific for the corresponding group.
+Following this, gene set enrichment analysis may be used
+in an attempt to test which pathways are specific to one, but not the other
+group.
 
 In this paper, we show that under reasonable assumptions this approach may
-result in apparent enrichments even if there are no real differences
+result in apparent enrichments even if there are no real statistically
+significant differences
 between the groups. To this end, we randomly split a cohort into two
@@
 differential expression between treatment and control. Thus, the enriched
-terms are relevant to the studied project, yet while they do reflect real
-processes linked to treatment, they are not demonstrative of differences
-between the study groups. Finally, we use literature search to show that
-this incorrect approach to study group-specific treatment effects is not
-uncommon. In fact, while VDs are a useful visualisation tool also in
+terms are relevant to the biological question at hand, yet while they do
+reflect real processes linked to viral infection, they do not correspond to
+the differences between the study groups. Finally, we use literature search to
+show that this incorrect approach to study group-specific treatment effects
+is not uncommon. In fact, while VDs are a useful visualisation tool also in
 transcriptomics, in more than quarter of the papers where VDs were used,
 group-specific genes were defined as significant in one, but not other
-groups, and in 19% a gene set enrichment was performed.
+groups, and in 19% of the papers a gene set enrichment was performed.
 
@@
 used the data set GSE156063 [@mick2020upper] in two approaches (an
-incorrect and the correct one) which arrive at opposite conclusions.
+incorrect and the correct one) to arrive at opposite conclusions.
 
@@
 the groups G1 and G2 separately using standard bioinformatic tools. For each comparison, we defined
-DEGs
-by using an false discovery rate (FDR) threshold of 
+DEGs as genes for which the 
+false discovery rate (FDR) was lower than 
 `r padj_thr`
-and absolute log~2~ fold change (LFC) threshold of
+and absolute log~2~ fold change (LFC) was higher than
 `r lfc_thr`. There were
@@
 difference between these two groups of individuals, as evidenced by a small
-overlap in commonly regulated genes, while the majority of DEGs is
+overlap in commonly regulated genes. The majority of DEGs is
 significant in one comparison only.
@@
 pathways are upregulated in each of the two groups, we used a standard generation I
-gene set enrichment analysis – a hypergeometric test – on the DEGs in each
+gene set enrichment analysis — a hypergeometric test — on the DEGs in each
 group. Gene sets for the gene set enrichment analysis were taken from the Gene Ontology (GO) database.
@@
 Here, a similar picture
-emerged. Just
+emerged. Overall,
 `r nrow(gsea_res$g1)` gene sets were significantly enriched in G1, and
-`r nrow(gsea_res$g2)` gene sets were significantly enriched in G2. Again,
-both the Venn diagram (Fig. 1, B) and the results of
+`r nrow(gsea_res$g2)` gene sets were significantly enriched in G2. 
+Both the Venn diagram (Fig. 1, B) and the results of
 enrichments (Fig. 1, C and D) suggest that there is a
@@
 Importantly, the different GO terms enriched in the two groups were related
-to infectious diseases, and may tempt to speculate about the underlying
-biological differences between these two groups (e.g.  significance of
+to infection, and may tempt to speculate about the underlying
+biological differences between these two groups. For example, the significance of
 Toll like receptor 4 pathway in G1, but not G2; and, vice versa, significance of
-response to interleukin 7 in G2, but not in G1).
+response to interleukin 7 in G2, but not in G1 may be considered as
+evidence of altered immune response to the virus in G2 as compared to G1.
 
@@
 
-The groups G1 and G2 were randomly sampled from the same data set.
+However, the groups G1 and G2 were randomly sampled from the same data set.
 In fact, repeated re-sampling always results in some genes being found to
@@
 fact that one does not expect any major differences between sets of
-individuals randomly drawn from one population. Thus, the conclusions drawn
+individuals randomly drawn from a single population. Thus, the conclusions drawn
 from a Venn diagram-driven gene set enrichment analysis are artefactual. Closer
@@
 respectively. Thus, genes which are significant in one, but not in the
-other comparison often have have similar log~2~ fold changes (e.g. Fig.
+other comparison tend to have have similar log~2~ fold changes in both
+groups (e.g. Fig.
 2, *A*, *C* and *D*).
@@
 relevant to the research hypothesis being tested. Below, we will show why
-these terms (rather than random terms which have no obvious relevance for
-an infectious disease) appear in the results.
-
-To understand how gene set enrichment actually gives significant
-results in such randomly generated gene sets, despite absence of genes with
+these terms (rather than random terms which have no obvious relevance to an
+infectious disease) appear in the results.
+
+To understand how significant results appear in a gene set enrichment
+analysis in randomly generated groups despite absence of genes with
 significant interaction, it is first necessary to consider the definition
@@
 Benjamini-Hochberg procedure [@benjamini1995controlling] ensures that among
-genes for which the adjusted p~adj~ < 0.05 there are at most 5% false
+genes for which FDR < 0.05 there are at most 5% false
 positives irrespective of the sample size.
@@
 Even assuming a test power of 80%, the FNR is 20%. Clearly, false negatives (FNs) occur
-at much higher rates than false positives (FPs). In case of high throughput
+at much higher rates than false positives (FPs). In the case of high throughput
 data sets, where the FPR is controlled by Bejnamini-Hochberg procedure or a
@@
 
-These FNs will occur
+These FNs occur
 at a much higher rate within the sets of DEGs defined by the
@@
 we have used a log~2~ fold change threshold of `r lfc_thr`, because raising
-it would increase the fraction of FNs. However, selecting more biologically
-only relevant genes which a substantially higher effect size may influence
+it would increase the fraction of FNs. However, filtering for biologically
+relevant genes with a substantially higher effect size may influence
 the observed enrichments. On the other hand, raising the log~2~ fold change
@@
 outdir <- "montecarlo"
+merged_res_fn <- file.path(outdir, "merged_res.rds")
+
+if(!file.exists(merged_res_fn)) {
+  warning("Running simulations. This will take a long time.")
+  source("montecarlo.R")
+}
 
@@
 analysis was applied to genes significant in one, but not the other
-group in order to find gene sets and pathways regulated by one, but not
-the other group.
+group in order to find group-specific differences.
 
@@
 papers in which the Venn diagram was combined with comparing significance
-to non-significance; *Enrichment*, total numebr (percentage) of papers
+to non-significance; *Enrichment*, total number (percentage) of papers
 which combined the Venn diagram with a gene set enrichment analysis.
@@
 set of supposedly "specific" DEGs. In summary, in at least a quarter of the
-papers on differential expression in which a venn diagram was used, it was
+papers on differential expression in which a Venn diagram was used, it was
 illustrating an incorrect statistical procedure which may result in
@@
 resulting pathways or gene ontologies are misleadingly relevant. For
-example, the identified gene sets are associated with immune answer for
-research hypotheses involving an infectious disease, or cancer pathways if
+example, the identified gene sets are associated with immune response for
+a research hypotheses involving an infectious disease, or cancer pathways if
 the underlying research hypothesis involved cancer treatment. Such results
@@
 say that VDs are not a useful tool, even in the context of transcriptomics
-and gene set enrichments, if used correctly. For example, analysis of an
-intersection of DEGs (i.e., by considering genes from the overlap in a VD)
-is not an incorrect procedure.  Genes in the overlapping part of a VD are
-significant in both (or all) comparisons, hence no comparison between
-significance and non-significance is made.
+and gene set enrichments, if used correctly. For example, gene set
+enrichment analysis of an intersection of DEGs (i.e., by considering genes
+from the overlap in a VD) is not an incorrect procedure.  Genes in the
+overlapping part of a VD are significant in both (or all) comparisons,
+hence no comparison between significance and non-significance is made.
 
@@
 second comparison. This allows an intuitive assessment of the differences
-between the two comparisons, in especially with combination of color coding
-the genes either are significant in the interaction or using the disco
-score (see Fig. 2, *E* for an example).
+between the two comparisons, in especially in combination with color coding
+the genes which either are significant in the interaction or by coloring
+using the disco score (see Fig. 2, *E* for an example).
 
 Defining group-specific genes based on significant difference in one,
-but no significant difference in another comparison is thus not only a
-statistical fallacy, but when combined with gene set enrichment analysis
-also one which can lead to potentially sound-looking results. This method
+but no significant difference in another comparison is thus more than only a
+statistical fallacy leading to erroneous results. When combined with gene
+set enrichment analysis it can lead to potentially
+sound-looking, and therefore particularly misleading results. This method
 of obtaining specific differences between groups should therefore be
-abandoned in favor of statistically correct approaches.
-Furthermore, gene set enrichment analysis must never be applied
-to sets of genes defined as significant in one comparison, but not the
-other.
+abandoned in favor of statistically correct approaches.  Furthermore, gene
+set enrichment analysis must never be applied to sets of genes defined as
+significant in one comparison, but not the other.
 
@@
 
-This document has been written as an Rmarkdown file, containing all
-statistical calculations required to replicate the findings and figures.
-The markdown file, along with additional files required to recreate this
-manuscript have been uploaded to [https://github.com/bihealth/manuscript_venn_diagrams](https://github.com/bihealth/manuscript_venn_diagrams).
+This manuscript has been written in Rmarkdown [@xie2018r].  
+All statistical calculations required to replicate the findings and figures
+are contained in the source Rmarkdown file.  The Rmarkdown file, along with
+additional files required to recreate this manuscript as well as the
+results
+of literature survey have been uploaded to
+[https://github.com/bihealth/manuscript_venn_diagrams](https://github.com/bihealth/manuscript_venn_diagrams).
 
@@
 estimate the frequency of the incorrect use of Venn diagrams. We searched
-for articles from 2020 mentioning the phrases "differential expression" and "venn
+for articles including the phrases "differential expression" and "venn
 diagram" in three journals: Scientific Reports (2020), Nature Communications (2020) and
@@
 groups, but not significant in other groups or (iii) for which a
-clear-cut error could not be indicated past any reasonable doubt were not
+clear-cut error could not be identified past any reasonable doubt were not
 considered incorrect.
@@
 
+```{r}
+sessioninfo::session_info(to_file=TRUE)
+```
+
 ## Bibliography

### 2022-03-08 mk5/montecarlo.R

@@ -1,0 +1,181 @@
+## monte carlo simulation: 100 replicates of the procedure shown in the
+## paper
+
+## output directory
+outdir <- "montecarlo"
+dir.create(path=outdir, showWarnings = FALSE)
+
+## number of replicates
+n_sim <- 100
+
+## number of replicates per stratum (2 x n_stratum for one treatment/group)
+## total sample size is 8 x n_stratum
+n_stratum <- c(5, 10, 20)
+
+## thresholds for the analysis
+padj_thr <- 0.05
+lfc_thr  <- 1.5
+
+## lfc thresholds to test
+lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
+
+geo <- readRDS("GSE156063.geo")
+covar <- as(phenoData(geo), "data.frame")
+
+## remove columns with only one value
+boring <- map_lgl(covar, ~ length(unique(.x)) == 1)
+covar <- covar[ , !boring ] 
+
+## clean up
+covar <- covar %>% 
+  dplyr::rename(gender = "gender:ch1") %>%
+  mutate(disease = gsub(" .*", "", .data[["disease state:ch1"]])) %>%
+  mutate(label = description) %>%
+  mutate(group = disease) %>%
+  arrange(description) %>%
+  dplyr::select(all_of(c("title", "label", "gender", "disease", "group")))
+
+counts <- read_csv("GSE156063_swab_gene_counts.csv.gz")
+.tmp <- counts[[1]]
+counts <- as.matrix(counts[,-1])
+rownames(counts) <- .tmp
+#counts <- counts[ , covar$description ]
+counts <- counts[ , covar$label ]
+lcpm   <- edgeR::cpm(counts, log=TRUE)
+#stopifnot(all(colnames(counts) == covar$description))
+stopifnot(all(colnames(counts) == covar$label))
+
+annot <- data.frame(ENSEMBL = rownames(counts))
+
+.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("ENTREZID"), keytype="ENSEMBL")
+annot$ENTREZID <- .tmp[ match(annot$ENSEMBL, names(.tmp)) ]
+
+.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("SYMBOL"), keytype="ENSEMBL")
[…]

### 2022-03-08 mk5/manuscript.rmd
bug in figure 2: incorrect samples chosen for visualization

@@
 [@blalock1965theory]. To find out whether it is statistically significant,
-an appriopriate statistical test for interaction should be employed.
+an appropriate statistical test for interaction should be employed.
 
@@
 show that this incorrect approach to study group-specific treatment effects
-is not uncommon. In fact, while VDs are a useful visualisation tool also in
+is not uncommon. In fact, while VDs are a useful visualization tool also in
 transcriptomics, in more than quarter of the papers where VDs were used,
@@
 the differences between G1 and G2 in the response to infection. For
-example, we ask which genes or pathways are specifically upregulated by SC2
+example, we ask which genes or pathways are specifically up-regulated by SC2
 infection in G1 as compared to G2, and vice versa. In the following, we
@@
 covar  <- covar[ sel, ]
+lcpm   <- lcpm[ , sel ]
 covar$group <- as.character(covar$group)
@@
 To understand which
-pathways are upregulated in each of the two groups, we used a standard generation I
+pathways are up-regulated in each of the two groups, we used a standard generation I
 gene set enrichment analysis — a hypergeometric test — on the DEGs in each
@@
 individuals randomly drawn from a single population. Thus, the conclusions drawn
-from a Venn diagram-driven gene set enrichment analysis are artefactual. Closer
+from a Venn diagram-driven gene set enrichment analysis are based on
+artifacts. Closer
 inspection of genes which are DEGs in one group, but not the other
@@
 at a much higher rate within the sets of DEGs defined by the
-non-overlaping areas of the VDs, that is DEGs considered to be "specific"
+non-overlapping areas of the VDs, that is DEGs considered to be "specific"
 for one group or other in a naive approach.
-To illustrate this phenomenon, we have analysed the full data set from which G1 and G2
+To illustrate this phenomenon, we have analyzed the full data set from which G1 and G2
 were drawn (Fig. 3), comparing the 
@@
 as DEGs in only one of the groups G1, G2 are DEGs when all data were
-analysed.
+analyzed.
 
@@
 @nieuwenhuis2011erroneous observed incorrect analyses of interactions in
-about half of the analysed papers from top neuroscience journals where the
+about half of the analyzed papers from top neuroscience journals where the
 authors considered an experimental design allowing for a test for
@@
 common in transcriptomics, too. To this end, we have searched three
-journals – from broad to specialized – for the occurence of the terms
-"differential expression" with "venn diagrams" (Tab. 2). Next, we analysed
+journals – from broad to specialized – for the occurrence of the terms
+"differential expression" with "venn diagrams" (Tab. 2). Next, we analyzed
 the selected papers from year 2020 (and years 2015-2020 in one case) to
@@
 *Journal*, journal title; *Years*, publication dates; *Total*, total number
-of papers found using the search phrase; *Analysed*, total number of papers
-which have been analysed for correctness; *Incorrect*, total number
+of papers found using the search phrase; *Analyzed*, total number of papers
+which have been analyzed for correctness; *Incorrect*, total number
 (percentage) of
@@
                           Total=.x$total,
-                          Analysed=.x$analysed,
+                          Analyzed=.x$analysed,
                           Incorrect=.x$incorrect_count,
                           Enrichment=.x$incorrect_count - .x$incorrect_count_vdonly))  %>%
-  mutate(Analysed=as.numeric(Analysed)) %>%
+  mutate(Analyzed=as.numeric(Analyzed)) %>%
   mutate(Total=as.numeric(Total))
@@
              data.frame(Journal="**Total:**", Years="", Total=sum(tab$Total),
-                        Analysed=sum(tab$Analysed),
+                        Analyzed=sum(tab$Analyzed),
                         Incorrect=sum(tab$Incorrect),
@@
 tab2 <- tab %>%
-  mutate(Incorrect=sprintf("%d (%.0f%%)", Incorrect, 100 * Incorrect / Analysed)) %>%
-  mutate(Enrichment=sprintf("%d (%.0f%%)", Enrichment, 100 * Enrichment / Analysed))
+  mutate(Incorrect=sprintf("%d (%.0f%%)", Incorrect, 100 * Incorrect / Analyzed)) %>%
+  mutate(Enrichment=sprintf("%d (%.0f%%)", Enrichment, 100 * Enrichment / Analyzed))
 
@@
 that of the 
-`r tab2$Analysed[nrow(tab2)]` analysed articles
+`r tab2$Analyzed[nrow(tab2)]` analyzed articles
  which used the terms "venn diagram" and
@@
 common statistical fallacy [@gelman2006difference]. Just as absence of evidence is not evidence of
-absence, the failure to reject the null hypothesis does not consitute the
+absence, the failure to reject the null hypothesis does not constitute the
 same level of evidence as rejecting it. However, when such an incorrect
@@
 for example Fig. 2, *A-D*). However, the point of VDs is to show a grand
-overview of the whole analysis summarising thousands of results for the
-analysed genes. As an alternative of such an overview, we suggest plotting
+overview of the whole analysis summarizing thousands of results for the
+analyzed genes. As an alternative of such an overview, we suggest plotting
 the log~2 fold changes in one comparison against log~2~ fold changes in the

### 2022-03-09 mk5/manuscript.rmd
fixed small formatting issues

@@
 
-```{r montecarlo,cache=TRUE,fig.width=7.5,fig.height=6.5}
+```{r montecarlo,cache=TRUE,fig.width=8,fig.height=6}
 ## note: given that the monte carlo simulation is quite time consuming, we
@@
 g0 <- ggplot(by_fdr, aes(x=lfc, y=value)) + geom_violin() +
-  ylab("Number of DEGs") + xlab("FDR threshold")
+  ylab("DEGs") + xlab("FDR threshold")
 
@@
 g1 <- ggplot(by_lfc_thr, aes(x=lfc, y=value)) + geom_boxplot() +
-  ylab("Number of DEGs") + xlab("log2 fold change threshold")
+  ylab("DEGs") + 
+  xlab(expression(log[2]~fold~change~threshold)) +
+  theme(text=element_text(size=16))
 
@@
 g2 <- ggplot(gsea_by_lfc_thr, aes(x=lfc, y=value)) + geom_boxplot() +
-  ylab("Number of enriched gene sets") + xlab("log2 fold change threshold")
+  ylab("Enriched gene sets") + 
+  xlab(expression(log[2]~fold~change~threshold)) +
+  theme(text=element_text(size=16))
 
@@
 g3 <- ggplot(by_ssize, aes(x=ss, y=value)) + geom_boxplot() +
-  ylab("Number of DEGs") + xlab("Number of samples per group/treatment")
+  ylab("DEGs") + xlab("Samples per group/treatment") +
+  theme(text=element_text(size=16))
 
@@
 g4 <- ggplot(gsea_by_ss, aes(x=ss, y=value)) + geom_boxplot() +
-  ylab("Number of enriched gene sets") + xlab("Number of samples per group/treatment")
+  ylab("Enriched gene sets") + xlab("Samples per group/treatment") +
+  theme(text=element_text(size=16))
 

### 2022-04-05 mk5/manuscript.rmd
final update

@@
 Due to the presence of false negatives, genes significant in one, but not
-the other group are enriched in gene sets which which correspond to the
+the other group are enriched in gene sets which correspond to the
 overall effect of the treatment.  Thus, the results appear related to the
@@
 ## Set default options for the knitr RMD processing
-knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=FALSE,autodep=TRUE, results="hide")
+knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=FALSE,autodep=TRUE, results="hide",dev="jpeg")
 ```
@@
 
-```{r fig1, fig.width=15,fig.height=5.5,dpi=300,dev="png"}
+```{r fig1, fig.width=15,fig.height=5.5,dpi=300,dev="jpeg"}
 p1 <- ggvenn(list(G1=res$g1 %>% filter(DEG) %>% pull(ENSEMBL),
@@
              show_percentage=FALSE) +
-      ggtitle("Differentially expressed genes")
+      labs(title="Differentially expressed genes") +
+      ylim(-1, 1.4)
 p2 <- ggvenn(list(G1=gsea_res$g1 %>% pull(ID),
@@
               show_percentage=FALSE) +
-      ggtitle("Enriched GO terms")
+      labs(title="Enriched GO terms") +
+      ylim(-1, 1.4)
 
@@
 
-```{r fig2, fig.width=10, fig.height=5.5, dpi=300,dev="png"}
+```{r fig2, fig.width=10, fig.height=5.5, dpi=300,dev="jpeg"}
 #id    <- gsub(" ", "_", toupper("response to interferon beta"))
@@
 
-```{r fig3,dpi=300,dev="png"}
+```{r fig3,dpi=300,dev="jpeg"}
 tmp <- merge(res.merged, res$full, by="ENSEMBL")
@@
 
-This manuscript has been written in Rmarkdown [@xie2018r].  
+This manuscript has been written in R markdown [@xie2018r].  
 All statistical calculations required to replicate the findings and figures
-are contained in the source Rmarkdown file.  The Rmarkdown file, along with
+are contained in the source R markdown file.  The R markdown file, along with
 additional files required to recreate this manuscript as well as the
